### 2019-09-09 assignment_one/article/main.tex
Working on data generation.

\documentclass[article, a4paper, oneside, 12pt]{memoir}
\linespread{1.25}
\setlength{\parskip}{1em}
\setlength{\parindent}{0em}

\usepackage[]{mathtools}
\usepackage[]{amssymb}
\usepackage[]{libertine}
\usepackage[]{libertinust1math}
\usepackage[T1]{fontenc}
\usepackage[]{microtype}
\usepackage[]{geometry}
\usepackage[]{amsthm}
\usepackage[]{thmtools}
\usepackage[usedvipsnames]{xcolor}
\usepackage[]{bm}
\usepackage[noabbrev, capitalize]{cleveref}
\usepackage[]{amsmath}
\usepackage[]{commath}

\declaretheorem[style=remark, name=Remark,
mdframed={
}]{remark}

\title{\textsc{Regression Techniques \\
  FYS-STK4155 \\
Assignment One}}
\author{Ivar Haugal{\o}kken Stangeby}

\newcommand{\x}{\bm{x}}
\newcommand{\C}{\mathbb{C}}
\newcommand{\X}{\bm{X}}

\newcommand{\mat}[1]{\bm{#1}}
\newcommand{\y}{\bm{y}}
\newcommand{\data}{\bm{\Omega}}
\newcommand{\cost}{\mathcal{C}}
\newcommand{\N}{\mathcal{N}}
\newcommand{\basis}{\mathcal{B}}
\newcommand{\set}[1]{\left\{\, #1 \right\}}

\begin{document}
  \maketitle

  \chapter{Introduction}

  Regression analysis is the act of estimating relationships among variables.
  In this project, we study various regression methods in more detail. In
  particular, we compare the \emph{ordinary least squares} (OLS) method with
  the \emph{Ridge regression} and \emph{Lasso regression} techniques. As we
  shall see, these three methods are all variations over the same theme. We
  start by testing the methods on noisy data sampled from a function known as
  \emph{Franke's} bivariate test function, which has been in widespread use in
  the evaluation of interpolation and data fitting techniques. Finally, we run
  regression on real terrain data, comparing the aforementioned methods.

  \chapter{Least Squares}

  We start by giving an introduction to ordinary least square regression. This
  method is perhaps the most widely used regression technique due to its
  simplicity, both in its derivation, and its implementation. The goal of
  regression analysis is to fit a \emph{model function} \( f(\x, \beta) \) to a
  set of \( n \) data points \( \data = (\x_i, y_i)_{i=1}^n \). A simple example
  is that of a linear polynomial with two parameters:
  \begin{equation}
    f(x, \beta) = \beta_0 + \beta_1 x.
  \end{equation}
  The \emph{model parameters} \( \beta \) are determined in order to minimize a
  suitable \emph{cost function} \(\cost(\data, \beta)\) which measures to which
  extent the model function manages to capture trends in the data \( \data \).

  One common cost function is the one which involves the sum of squared
  residuals (or squared errors):
  \begin{equation}
    \label{eq:mse}
    \cost(\data, \beta) = \frac{1}{n}\sum_{i = 1}^n \varepsilon_i^2 \coloneqq \frac{1}{n}\sum_{i=1}^n (y_i - \hat{y}_i)^2
  \end{equation}
  where \( \hat{y}_i \coloneqq f(\x_i, \beta) \). The method involving the
  minimization of this specific cost function is known as \emph{least squares}.

  \begin{remark}
    It is often assumed a priori that the data is infact generated from a noisy model, such that each \( y_i \) can be described 
    as
    \begin{equation}
      y_i = f(\x_i) + \varepsilon_i
    \end{equation}
    where each \( \varepsilon_i \sim \N(0, \sigma^2) \) is normally distributed
    with zero mean and variance \( \sigma^2 \). 
  \end{remark}

  \section{General linear models.} We are often interested in finding the
  best model function in a specific function space.  Assuming this space has a
  basis \( \basis = \set{\varphi_i}_{i=1}^M\), we may write our model function
  in terms of the basis functions and the model parameters as
  \begin{equation}
    \label{eq:gen_lin_mod}
    f(\x, \beta) \coloneqq \sum_{i = 1}^M \beta_i\varphi_i(\x).
  \end{equation}
  With this formulation, we can represent the approximation \( \hat{\y} \) to
  \( \y \) as a matrix product 
  \begin{equation}
    \hat{\y} = \X\beta
  \end{equation}
  where \( \X \) is the \emph{design matrix} defined by components \(\X_{ij} =
  \varphi_j(\x_i)\), and \( \beta = [\beta_0, \ldots, \beta_M] \) is the model
  parameters. With this notation in mind, we can also write the cost function
  as
  \begin{equation}
    \cost(\data, \beta) = \frac{1}{n} (\y - \hat{\y})^T (\y - \hat{\y}).
  \end{equation}

  \paragraph{Optimizing the parameters of a general linear model.}

  Assume now that our model is of the form given in \cref{eq:gen_lin_mod} and
  that our cost function is the mean squared error defined in \cref{eq:mse}. We
  are interested in finding the parameters \( \beta \) that minimize the cost
  function \( \cost(\data, \beta) \). Since \( \cost(\data, \beta) \) is
  convex, it suffices to differentiate with respect to \( \beta \) and equating
  to zero.

  We have that
  \begin{align}
    \dpd{\cost(\data, \beta)}{\beta} = \X^T(\y - \X\beta), 
  \end{align}
  and equating this to zero yields the familiar \emph{normal equations}
  \begin{equation}
    \X^T \y = \X^T\X\beta.
  \end{equation}
  If the matrix \( \X^T\X \) is invertible, we may obtain the solution by
  direct numerical matrix inversion. In this case, the optimal model parameters
  are found directly by
  \begin{equation}
    \beta = (\X^T\X)^{-1}\X^T\y.
  \end{equation}
  However, these matrices may be ill-conditioned when the number of equations
  are very large, and it is therefore common to apply approximate solvers for
  the inverse, for instance using low-rank approximation based on
  \textsc{SVD}-decomposition, which we will briefly turn to in the following.

  \paragraph{Singular value decomposition.}

  Recall that any matrix \( \mat{A} \in \C^{n, m} \) can be decomposed as
  \begin{equation}
    \mat{A} = \mat{U} \mat{\Sigma}\mat{V}^T,
  \end{equation}
  where \( \mat{U} \) and \( \mat{V} \) comprised of the eigenvectors of \(
  \mat{A}\mat{A}^T \) and \( \mat{A}^T\mat{A} \) respectively. As these
  eigenvectors are orthonormal, it follows that both \( U \) and \( V \) are
  unitary matrices. Furthermore, \( \mat{\Sigma} \) is a matrix 
  \begin{equation}
    \mat{\Sigma} = \begin{bmatrix}
      \Sigma_1 & 0 \\
       0 & 0
    \end{bmatrix},
  \end{equation}
  where \( \Sigma_1 \) is a square diagonal matrix of size \( r \times r \)
  with the non-zero singular values of \( \mat{A} \). The integer \( r \) is
  the \emph{rank} of \( \mat{A} \). As the matrix \( \Sigma \) is mostly
  containing zeros, the information stored in \( \mat{A} \) is attributed to
  only some parts of \( \mat{U} \) and \( \mat{V} \). We can remove the
  redundant parts, and more compactly express \( \mat{A} \) as \( \mat{A} =
  \mat{U}_1 \mat{\Sigma}_1 \mat{V}_1^T \). Here, \( \mat{U}_1 \) is \( m \times
  r \) and \( \mat{V}_1 \) is \( n \times r \).
\end{document}

### 2019-09-10 assignment_one/article/main.tex
Working on article.

@@
 \linespread{1.25}
-\setlength{\parskip}{1em}
-\setlength{\parindent}{0em}
+%\setlength{\parskip}{1em}
+%\setlength{\parindent}{0em}
 
@@
 \usepackage[]{bm}
-\usepackage[noabbrev, capitalize]{cleveref}
 \usepackage[]{amsmath}
 \usepackage[]{commath}
+\usepackage[noabbrev, capitalize]{cleveref}
 
@@
 
-\title{\textsc{Regression Techniques \\
+\title{\textsc{Regression And Resampling Techniques \\
   FYS-STK4155 \\
@@
 \newcommand{\X}{\bm{X}}
+\newcommand{\diag}[1]{\mathrm{diag}(#1)}
+\newcommand{\MSE}[1]{\mathrm{MSE}(#1)}
+\newcommand{\R}[1]{\mathrm{R}^2(#1)}
 
@@
 
-  \chapter{Least Squares}
+  \chapter{Regression Techniques}
 
-  We start by giving an introduction to ordinary least square regression. This
-  method is perhaps the most widely used regression technique due to its
-  simplicity, both in its derivation, and its implementation. The goal of
-  regression analysis is to fit a \emph{model function} \( f(\x, \beta) \) to a
-  set of \( n \) data points \( \data = (\x_i, y_i)_{i=1}^n \). A simple example
-  is that of a linear polynomial with two parameters:
+  In general, the goal of regression analysis is to fit a \emph{model function}
+  \( f(\x, \beta) \) to a set of \( n \) data points \( \data = (\x_i,
+  y_i)_{i=1}^n \). A simple example is that of a linear polynomial with two
+  parameters:
   \begin{equation}
@@
   extent the model function manages to capture trends in the data \( \data \).
+  It is the choice of cost function \( \matcal{C}(\data, \beta)\) which
+  distinguishes  the three regression techniques, OLS, Lasso regression, and
+  Ridge regression.
+
+  \begin{remark}
+    It is often assumed a priori that the data is infact generated from a noisy
+    model, such that each \( y_i \) can be described as
+    \begin{equation}
+      y_i = f(\x_i) + \varepsilon_i
+    \end{equation}
+    where each \( \varepsilon_i \sim \N(0, \sigma^2) \) is normally distributed
+    with zero mean and variance \( \sigma^2 \). This assumption on the error
+    gives rise to what is known as \emph{general linear models}.
+  \end{remark}
+
+  \paragraph{The design matrix.} We are often interested in finding the best
+  model function in a specific function space.  Assuming this space has a basis
+  \( \basis = \set{\varphi_i}_{i=1}^M\), we may write our model function in
+  terms of the basis functions and the model parameters as
+  \begin{equation}
+    \label{eq:gen_lin_mod}
+    f(\x, \beta) \coloneqq \sum_{i = 1}^M \beta_i\varphi_i(\x).
+  \end{equation}
+  As an example, if we were to use the space \( \Pi_2^2 \) of bi-quadratic
+  polynomials, our basis functions would be
+  \begin{equation}
+    \basis = \set{1, x, x^2, y, y^2, xy}.
+  \end{equation}
+  With this formulation, we can represent the approximation \( \hat{\y} \) to
+  \( \y \) as a matrix product 
+  \begin{equation}
+    \label{eq:matrix_prod}
+    \hat{\y} = \X\beta
+  \end{equation}
+  where \( \X \) is the \emph{design matrix} defined by components \(\X_{ij} =
+  \varphi_j(\x_i)\), and \( \beta = [\beta_1, \ldots, \beta_M] \) is the model
+  parameters.
+
+  \paragraph{Performance metrics.}
+  In order to evaluate the how accurately the model function \( f(\data, \beta)
+  \) captures trends in the target data \( \y \), a few standard performance
+  metrics are  used.  Firstly, the \emph{mean squared error}:
+  \begin{equation}
+    \MSE{\y, \hat{\y}} \coloneqq \frac{1}{n} \sum_{i=1}^n (y_i - \hat{y}_i)^2
+  \end{equation}
+  which simply averages the squared error over all samples and estimates.
+  Secondly, we have the \emph{coefficient of determination} or
+  \emph{\(R^2\)-score}:
+  \begin{equation}
+    \R{\y, \hat{y}} \coloneqq 1 - \frac{\sum_{i=1}^n (y_i -
+    \hat{y}_i)^2}{\sum_{i=1}^n (y_i - \overline{y})^2}.
+  \end{equation}
+  The \(R^2\)-score measures much of the variation in \( \y \) which can be
+  attributed to a simple linear relation between \( \x \) and \( \y \). It is a
+  ratio, thus a value of one tells us that all the variation in the data can be
+  attributed to an approximate linear relationship between the data \(\x\) and
+  the response variable \( \y \). A value of zero indicates that a non-linear
+  model may be preferrable.
+
+  \section{Ordinary Least Squares}
 
@@
   minimization of this specific cost function is known as \emph{least squares}.
-
-  \begin{remark}
-    It is often assumed a priori that the data is infact generated from a noisy model, such that each \( y_i \) can be described 
-    as
-    \begin{equation}
-      y_i = f(\x_i) + \varepsilon_i
-    \end{equation}
-    where each \( \varepsilon_i \sim \N(0, \sigma^2) \) is normally distributed
-    with zero mean and variance \( \sigma^2 \). 
-  \end{remark}
-
-  \section{General linear models.} We are often interested in finding the
-  best model function in a specific function space.  Assuming this space has a
-  basis \( \basis = \set{\varphi_i}_{i=1}^M\), we may write our model function
-  in terms of the basis functions and the model parameters as
-  \begin{equation}
-    \label{eq:gen_lin_mod}
-    f(\x, \beta) \coloneqq \sum_{i = 1}^M \beta_i\varphi_i(\x).
-  \end{equation}
-  With this formulation, we can represent the approximation \( \hat{\y} \) to
-  \( \y \) as a matrix product 
-  \begin{equation}
-    \hat{\y} = \X\beta
-  \end{equation}
-  where \( \X \) is the \emph{design matrix} defined by components \(\X_{ij} =
-  \varphi_j(\x_i)\), and \( \beta = [\beta_0, \ldots, \beta_M] \) is the model
-  parameters. With this notation in mind, we can also write the cost function
-  as
+  With the matrix notation from above in mind, we can also write the cost
+  function as
   \begin{equation}
@@
 
-  \paragraph{Optimizing the parameters of a general linear model.}
+  \paragraph{Optimizing the parameters in OLS.}
 
@@
   \begin{equation}
+    \label{eq:optimal}
     \beta = (\X^T\X)^{-1}\X^T\y.
@@
   \end{equation}
-  where \( \mat{U} \) and \( \mat{V} \) comprised of the eigenvectors of \(
+  where \( \mat{U} \) and \( \mat{V} \) are comprised of the eigenvectors of \(
   \mat{A}\mat{A}^T \) and \( \mat{A}^T\mat{A} \) respectively. As these
-  eigenvectors are orthonormal, it follows that both \( U \) and \( V \) are
-  unitary matrices. Furthermore, \( \mat{\Sigma} \) is a matrix 
+  eigenvectors are orthonormal, it follows that both \( \mat{U} \) and \(
+  \mat{V} \) are unitary matrices. Furthermore, \( \mat{\Sigma} \) is a matrix 
   \begin{equation}
@@
   redundant parts, and more compactly express \( \mat{A} \) as \( \mat{A} =
-  \mat{U}_1 \mat{\Sigma}_1 \mat{V}_1^T \). Here, \( \mat{U}_1 \) is \( m \times
-  r \) and \( \mat{V}_1 \) is \( n \times r \).
+  \mat{U}_1 \mat{\Sigma}_1 \mat{V}_1^T \) without loss of accuracy in the
+  decomposition. Here, \( \mat{U}_1 \) is \( m \times r \) and \( \mat{V}_1 \)
+  is \( n \times r \).
+
+  Applying the singular value decomposition to the matrix \(\X = \mat{U}
+  \mat{\Sigma} \mat{V}^T\) we can analyze the expression for the prediction \( \hat{\y} \)
+  in terms of the matrix \( \X^T \X \).  First of all, we have that
+  \begin{align}
+    \begin{split}
+    \X^T \X &= \mat{V}\mat{\Sigma}^T\mat{U}^T\mat{U}\mat{\Sigma} \mat{V}^T\\
+            &= \mat{V}\mat{D}\mat{V}^T, 
+    \end{split}
+  \end{align}
+  where we have used that \( \mat{U}\mat{U}^T = \mat{I} \) and defined \(
+  \mat{D} = \diag{\sigma_1^2, \ldots, \sigma_n^2} \).  Thus, plugging this into
+  \cref{eq:optimal,eq:matrix_prod}, we obtain the expression
+  \begin{align}
+    \begin{split}
+      \hat{\y} = \X \beta = \X (\mat{V}\mat{D}\mat{V}^T)^{-1} \mat{X}^T \y.
+    \end{split}
+  \end{align}
+  Finally, substituting \( \X = \mat{U}\mat{\Sigma}\mat{V}^T \) and noting that
+  diagonal matrices always commute, we end up with
+  \begin{align}
+    \begin{split}
+      \hat{\y} = \mat{U}\mat{U}^T \mat{y}.
+    \end{split}
+  \end{align}
+
 \end{document}

### 2019-09-10 assignment_one/article/main.tex
Working on article.

@@
-\documentclass[article, a4paper, oneside, 12pt]{memoir}
+\documentclass[dvipsnames, article, a4paper, oneside, 12pt]{memoir}
 \linespread{1.25}
@@
 \usepackage[]{commath}
+\usepackage[]{todonotes}
 \usepackage[noabbrev, capitalize]{cleveref}
+\usepackage[]{graphicx}
+\usepackage{titling}
 
@@
 }]{remark}
+\declaretheorem[style=remark, name=Assumption,
+mdframed={
+}]{assumption}
 
@@
 \newcommand{\R}[1]{\mathrm{R}^2(#1)}
+\newcommand{\expect}{\mathbb{E}}
+\newcommand{\OLS}{\mathrm{OLS}}
+\newcommand{\ridge}{\mathrm{ridge}}
 
@@
 \newcommand{\basis}{\mathcal{B}}
-\newcommand{\set}[1]{\left\{\, #1 \right\}}
+%\newcommand{\set}[1]{\left\{\, #1 \right\}}
+
+\pretitle{%
+  \begin{center}
+    \includegraphics[scale=0.7]{images/franke_shadow.pdf}\\[\bigskipamount]
+}
+\posttitle{\end{center}}
 
@@
   \maketitle
+
+  \tableofcontents
 
@@
   start by testing the methods on noisy data sampled from a function known as
-  \emph{Franke's} bivariate test function, which has been in widespread use in
-  the evaluation of interpolation and data fitting techniques. Finally, we run
-  regression on real terrain data, comparing the aforementioned methods.
+  \emph{Franke's} bivariate test function (visualized on the title page), which
+  has been in widespread use in the evaluation of interpolation and data
+  fitting techniques. Finally, we run regression on real terrain data,
+  comparing the aforementioned methods.
 
@@
   extent the model function manages to capture trends in the data \( \data \).
-  It is the choice of cost function \( \matcal{C}(\data, \beta)\) which
+  It is the choice of cost function \( \mathcal{C}(\data, \beta)\) which
   distinguishes  the three regression techniques, OLS, Lasso regression, and
@@
 
-  \begin{remark}
+  \begin{assumption}
+    \label{as:general_linear}
     It is often assumed a priori that the data is infact generated from a noisy
@@
     gives rise to what is known as \emph{general linear models}.
-  \end{remark}
+  \end{assumption}
 
@@
 
-  \paragraph{Optimizing the parameters in OLS.}
+  \subsection{Optimizing the parameters in OLS.}
 
@@
     \label{eq:optimal}
-    \beta = (\X^T\X)^{-1}\X^T\y.
+    \beta_{\OLS} = (\X^T\X)^{-1}\X^T\y.
   \end{equation}
@@
 
-  \paragraph{Singular value decomposition.}
+  \subsection{Singular value decomposition.}
 
@@
 
+  \subsection{The bias-variance tradeoff}
+
+  Since we are attempting to estimate a true distribution based on finite
+  samples, there are certain statistical properties it is worthwile discussing.
+  We follow the derivation in.  Recall \cref{as:general_linear} which states
+  that our true data is generated from a noisy model
+  \begin{equation}
+    y = f(\x) + \varepsilon
+  \end{equation}
+  with \( \varepsilon \sim \N(0, \sigma^2) \).  We select a finite sample from
+  the true data, which constitutes our finite data-set \( \data \).
+  Furthermore, we have a model function \( f(\data, \beta) \) which depends on
+  the selected data-set and the model parameters \( \beta \). Thus, our
+  cost-function \( \cost \) will depend on the specific sample taken from the
+  true distribution. It is of interest to examine the expected out-of-sample
+  error \( \expect_{\data, \varepsilon}[\cost(\data, \beta)] \) of the model
+  function over the sampled data set \( \data \) and the noise \( \varepsilon
+  \). This can be decomposed into three distinct parts,
+  \emph{\textcolor{Maroon}{variance}}, \emph{\textcolor{Blue}{bias}}, and
+  \emph{\textcolor{Green}{noise}}:
+  \begin{equation}
+    \expect_{\data, \varepsilon}[\cost(\data, \beta)] = \frac{1}{n} \textcolor{Blue}{\sum_{i=1}^n \left(f(\x_i) - \expect_{\data}\left[\hat{y}_i\right]\right)^2}  + \frac{1}{n}\textcolor{Maroon}{\sum_{i=1}^n \expect_{\data}\big[\left(\hat{y}_i - \expect_{\data} [\hat{y}_i]\right)^2\big]} + \textcolor{Green} \sigma^2.
+  \end{equation}
+  See \cref{ap:derivation} for the full derivation. The bias-term measures the
+  deviation of the expectation of the model function \(f(\data, \beta)\) from
+  the true value \( f \), while the variance-term measures the fluctuations
+  arising due to using finite samples.
+
+  \section{Ridge Regression}
+
+  When the dimensionality of the matrix \( \X \) is high, the problem of
+  near-collinear columns often arise, as mentioned earlier. Thus, the model
+  parameters \( \beta_{\OLS} \) of the ordinary least squares regression cannot
+  be determined. The propsed fix to this problem is to slightly perturb the
+  matrix \( \X^T \X \) by adding small positive values along the diagonal to
+  ensure its invertability. This is known as \emph{ridge regression} and yields
+  a model parameter estimator
+  \begin{equation}
+    \beta_{\ridge}(\lambda) = (\X^T\X + \lambda\mat{I})^{-1}\X^T\y
+  \end{equation}
+  where \( \lambda \geq 0 \) is a hyper parameter. To see why the resulting
+  modified matrix is invertible, one has to look at the singular values.  It
+  turns out that the model parameters \( \beta_{\ridge}(\lambda) \) directly
+  minimize the \emph{ridge loss function} defined as
+  \begin{equation}
+    \cost_{\ridge}(\data, \beta;  \lambda) \coloneqq \frac{1}{n} \sum_{i=1}^n (y_i - \hat{y}_i)^2 + \lambda \sum_{i=1}^p \beta_i^2 = \cost(\data, \beta) + \lambda \beta^T \beta,
+  \end{equation}
+  which we recognize as standard penalized least squares.
+
+  Ridge regregression carries several benefits. One of the more important is
+  its resilience to overfitting. In a high-dimensional setting, if many of the
+  data-features can be attributed to a few covariates, then the remaining
+  covariates do nothing more than fit to the noise. Large values of \( \beta_i
+  \) then a good indication that your model might be overfitting as some
+  covariates are given more weight than others.  The Ridge regression solves
+  this by directly penalizing large model parameters in the cost function.
+
+  \appendix
+  \chapter{Derivation of the bias-variance-noise relationship}
+  \label{ap:derivation}
+
+  \if 0
+  To this end --- by adding and subtracting \( f(\x_i) \) --- we have that 
+  \begin{align}
+    \begin{split}
+      \expect_{\data, \varepsilon} \big[\cost(\data, \beta)\big] &= \frac{1}{n}\expect_{\data, \varepsilon} \left[\sum_{i = 1}^n (y_i - \hat{y}_i)^2\right] \\
+                                                                 &= \frac{1}{n}\expect_{\data, \varepsilon} \left[\sum_{i=1}^n (y_i - f(\x_i) + f(\x_i) - \hat{y}_i)^2\right] \\
+                                                                 &= \frac{1}{n}\sum_{i=1}^n\left\{\overbrace{\expect_{\varepsilon} \left[(y_i - f(\x_i))^2\right]}^{= \sigma^2} + \expect_{\data}\left[(f(\x_i) - \hat{y}_i)^2\right] + 2\overbrace{\expect_{\varepsilon}\left[y_i - f(\x_i)\right]}^{= 0}\expect_{\data}\left[f(\x_i) - \hat{y}_i\right] \right\} \\
+                                                                 &= \frac{1}{n}\left\{n\sigma^2 + \expect_{\data} \left[(f(\x_i) - \hat{y}_i)^2\right]\right\} \\
+                                                                 &= \sigma^2 + \frac{1}{n}\sum_{i=1}^n\expect_{\data} \left[(f(\x_i) - \hat{y}_i)^2\right]\right\}.
+  \end{split}
+  \end{align}
+
+  Using the same trick of adding and subtracting, we may further decompose the second term as
+  \begin{align}
+    \begin{split}
+      \expect_{\data} \big[ (f(\x_i) - \hat{y}_i)^2\big] = \left(f(\x_i) - \expect_{\data}\big[\hat{y_i}\big]\right)^2 + \expect_{\data} \left[\left(\hat{y}_i - \expect_{\data}\left[\hat{y}_i\right]\right)^2\right]
+    \end{split}
+  \end{align}
+  \fi
+  Combining the two previous equations, we obtain
+
 \end{document}

### 2019-09-10 assignment_one/src/aux.py

@@ -48,24 +48,7 @@ def r2_score(y, y_hat):
     return 1 - num / den
 
 
-if __name__ == '__main__':
-    import matplotlib.pyplot as plt
-    from mpl_toolkits.mplot3d import Axes3D
-
-    x = np.linspace(0, 1, 50)
-    y = np.linspace(0, 1, 50)
-
-    X, Y = np.meshgrid(x, y)
-    Z = franke_function(X, Y)
-
-    fig = plt.figure()
-    axs = Axes3D(fig)
-
-    axs.plot_surface(X, Y, Z)
-    plt.show()
-
-
-def generate_data(N, noise=True, seed=None):
+def generate_data(N, noise=True, seed=None, return_mesh=False):
     """
     Samples (optionally noisy) data from the franke function with n parameter values in
     each parameter direction.
@@ -88,4 +71,24 @@ def generate_data(N, noise=True, seed=None):
     if noise:
         Z += np.random.normal(0, 1 / N, size=Z.shape)
 
-    return params, Z
+    if return_mesh:
+        return params, Z, X, Y
+    else:
+        return params, Z
+
+
+if __name__ == '__main__':
+    import matplotlib.pyplot as plt
+    from mpl_toolkits.mplot3d import Axes3D
+
+    x = np.linspace(0, 1, 50)
+    y = np.linspace(0, 1, 50)
+
+    X, Y = np.meshgrid(x, y)
+    Z = franke_function(X, Y)
+
+    fig = plt.figure()
+    axs = Axes3D(fig)
+
+    axs.plot_surface(X, Y, Z)
+    plt.show()

### 2019-09-10 assignment_one/src/franke_sampling.py

@@ -1,4 +1,4 @@
-import numpy as np
+from matplotlib import cm
 
 from assignment_one.src import ordinary_least_squares as ols
 from assignment_one.src.aux import generate_data, mean_squared_error, r2_score
@@ -30,15 +30,31 @@ def compute_beta_variance(beta):
 if __name__ == '__main__':
 
     import tqdm
+    import numpy as np
+    import matplotlib.pyplot as plt
+    from mpl_toolkits.mplot3d import Axes3D
 
     poly_degs = range(6)
-    N = 2000
+    N = 50
     seed = 42
     noise = True
-    params, z = generate_data(N, noise, seed)
+    params, z, X, Y = generate_data(N, noise, seed, return_mesh=True)
 
     for d in tqdm.tqdm(poly_degs, disable=True):
         beta, z_hat = perform_franke_ols(params, z, polynomial_degree=d)
+
+        z_hat = z_hat.reshape(N, N)
+        z = z.reshape(N, N)
+
+        fig = plt.figure()
+        ax1 = fig.add_subplot(111, projection='3d')
+        ax1.set_zlim3d(-0.2, 1.2)
+
+        ax1.plot_surface(X, Y, z_hat, alpha=0.5, cmap=cm.coolwarm)
+        ax1.scatter(X, Y, z, alpha=1, s=1, color='black')
+
+        plt.title(f'Polynomial degree $p = {d}$')
+        plt.show()
 
         z_hat = z_hat.ravel()
         z = z.ravel()
@@ -46,6 +62,5 @@ if __name__ == '__main__':
         print('Polynomial degree = ', d)
         print(f'\tMSE = {mean_squared_error(z, z_hat):.3f}')
         print(f'\tR2  = {r2_score(z, z_hat):.3f}')
-        np.var
         print(f'Predicted beta variance = {beta.var():.3f}')
         print('===========================================')

### 2019-09-10 assignment_one/src/ordinary_least_squares.py
Added ridge regression.

@@ -40,6 +40,32 @@ def ordinary_least_squares(X, y, pseudo_inv=True):
     return b
 
 
+def ridge_regression(X, y, l=0.1, pseudo_inv=True):
+    """
+    Solves the ridge regression problem by solving the equations X^T y = (X^T X + LI)b for b.
+
+    :param l: ridge hyper parameter
+    :param X: design matrix
+    :param y: ground truth / response variable
+    :param pseudo_inv: whether to use the svd-based inverse
+    :return: the parameters b minimizing the penalized mean squared error.
+    """
+
+    n = X.shape[0]
+    dot = X.T.dot(X) + l * np.eye(n)
+
+    if y.ndim != 1:
+        y = y.ravel()
+
+    if pseudo_inv:
+        dot_inv = svd_inv(dot)
+    else:
+        dot_inv = np.linalg.inv(dot)
+
+    b = dot_inv.dot(X.T).dot(y)
+    return b
+
+
 def svd_inv(A):
     """
     Returns the inverse of the matrix A based on a singular value decomposition of A.

### 2019-09-10 assignment_one/src/franke_sampling.py
Added ridge regression.

@@ -4,7 +4,7 @@ from assignment_one.src import ordinary_least_squares as ols
 from assignment_one.src.aux import generate_data, mean_squared_error, r2_score
 
 
-def perform_franke_ols(params, z, polynomial_degree=5):
+def perform_regression(params, z, polynomial_degree=5, ridge=False, l=0.1):
     """
     Performs a polynomial fit to sampled data from the Franke function.
     :return: parameters beta, and sampled points from predicted surface.
@@ -12,7 +12,11 @@ def perform_franke_ols(params, z, polynomial_degree=5):
     """
     n = int(np.sqrt(z.shape[0]))
     v = ols.design_matrix_from_parameters(params, polynomial_degree=polynomial_degree)
-    beta = ols.ordinary_least_squares(v, z)
+
+    if ridge:
+        beta = ols.ridge_regression(v, z, l=l)
+    else:
+        beta = ols.ordinary_least_squares(v, z)
     z_hat = (v @ beta).reshape(n, n)
 
     return beta, z_hat
@@ -25,6 +29,8 @@ def compute_beta_variance(beta):
     :param beta:
     :return:
     """
+
+    pass
 
 
 if __name__ == '__main__':
@@ -41,7 +47,7 @@ if __name__ == '__main__':
     params, z, X, Y = generate_data(N, noise, seed, return_mesh=True)
 
     for d in tqdm.tqdm(poly_degs, disable=True):
-        beta, z_hat = perform_franke_ols(params, z, polynomial_degree=d)
+        beta, z_hat = perform_regression(params, z, polynomial_degree=d, ridge=True, l=0.1)
 
         z_hat = z_hat.reshape(N, N)
         z = z.reshape(N, N)

### 2019-09-10 assignment_one/src/ordinary_least_squares.py

@@ -36,6 +36,7 @@ def ordinary_least_squares(X, y, pseudo_inv=True):
     else:
         dot_inv = np.linalg.inv(dot)
 
+    print(dot_inv.shape, X.T.shape, y.shape)
     b = dot_inv.dot(X.T).dot(y)
     return b
 
@@ -51,9 +52,8 @@ def ridge_regression(X, y, l=0.1, pseudo_inv=True):
     :return: the parameters b minimizing the penalized mean squared error.
     """
 
-    n = X.shape[0]
+    n = X.shape[1]
     dot = X.T.dot(X) + l * np.eye(n)
-
     if y.ndim != 1:
         y = y.ravel()
 
@@ -62,6 +62,7 @@ def ridge_regression(X, y, l=0.1, pseudo_inv=True):
     else:
         dot_inv = np.linalg.inv(dot)
 
+    print(dot_inv.shape, X.T.shape, y.shape)
     b = dot_inv.dot(X.T).dot(y)
     return b
 

### 2019-09-12 assignment_one/article/main.tex
Working on article

@@
 \linespread{1.25}
-%\setlength{\parskip}{1em}
-%\setlength{\parindent}{0em}
+\setlength{\parskip}{1em}
+\setlength{\parindent}{0em}
 
@@
 \usepackage[]{todonotes}
-\usepackage[noabbrev, capitalize]{cleveref}
 \usepackage[]{graphicx}
 \usepackage{titling}
+\usepackage[]{FiraMono}
+\usepackage[]{FiraSans}
+
+\usepackage[backend=biber, style=authoryear]{biblatex}
+\addbibresource{../../FYS-STK4155.bib}
+\usepackage[noabbrev, capitalize]{cleveref}
 
@@
 \newcommand{\ridge}{\mathrm{ridge}}
+\newcommand{\lasso}{\mathrm{lasso}}
+\newcommand{\python}{\textsf{Python}{}}
 
@@
   samples, there are certain statistical properties it is worthwile discussing.
-  We follow the derivation in.  Recall \cref{as:general_linear} which states
-  that our true data is generated from a noisy model
+  We follow the derivation from \autocite{mehtaHighbiasLowvarianceIntroduction2019}.  Recall
+  \cref{as:general_linear} which states that our true data is generated from a
+  noisy model
   \begin{equation}
@@
   covariates do nothing more than fit to the noise. Large values of \( \beta_i
-  \) then a good indication that your model might be overfitting as some
+  \) is then a good indication that your model might be overfitting as some
   covariates are given more weight than others.  The Ridge regression solves
   this by directly penalizing large model parameters in the cost function.
+
+  \section{Lasso Regression}
+
+  One drawback of the OLS and Ridge regression approaches is that irrelevant
+  features may still be present in the model. Lasso regression fixes this by
+  --- in addition to penalizing large model parameters --- also setting
+  sufficiently small parameters to zero. The associated cost function reads
+  \begin{equation}
+    \cost_{\lasso}(\data, \beta; \lambda) \coloneqq \frac{1}{n} \sum_{i=1}^n (y_i - \hat{y}_i)^2 + \lambda \sum_{i=1}^n |\beta_i|.
+  \end{equation}
+  Again, \( \lambda \leq 0 \) is a model hyper-parameter. Note that the
+  penalization-term now involves absolute values of the model parameters. As
+  the Lasso estimator may set model parameters to zero, it is effectively
+  producing a sparser model.
+
+  For further information on Ridge and Lasso regression, the excellent article
+  \citetitle{vanwieringenLectureNotesRidge2015} by
+  \textcite{vanwieringenLectureNotesRidge2015} is highly recommended.
+
+  \chapter{Implementation}
+
+  The implementations are all written in the programming language \python{} with
+  extensive use of the \python-packages \textsf{numpy} and \textsf{sklearn}.
+
+  \chapter{Numerical Results}
+
+  \section{Toy Data}
+
+  \section{Terrain Data}
 
@@
 
+\printbibliography
 \end{document}

### 2019-09-12 assignment_one/article/main.tex
Work work.

@@
 
-\usepackage[backend=biber, style=authoryear]{biblatex}
+\usepackage[backend=biber, style=alphabetic, url=false, doi=false]{biblatex}
 \addbibresource{../../FYS-STK4155.bib}
@@
 \newcommand{\python}{\textsf{Python}{}}
+\newcommand{\sgn}{\,\mathrm{sign}{}}
 
@@
   \chapter{Regression Techniques}
+  \label{chp:regression}
 
@@
     \label{eq:mse}
-    \cost(\data, \beta) = \frac{1}{n}\sum_{i = 1}^n \varepsilon_i^2 \coloneqq \frac{1}{n}\sum_{i=1}^n (y_i - \hat{y}_i)^2
+    \cost_\OLS(\data, \beta) = \frac{1}{n}\sum_{i = 1}^n \varepsilon_i^2 \coloneqq \frac{1}{n}\sum_{i=1}^n (y_i - \hat{y}_i)^2
   \end{equation}
@@
   \begin{equation}
-    \cost(\data, \beta) = \frac{1}{n} (\y - \hat{\y})^T (\y - \hat{\y}).
+    \cost_\OLS(\data, \beta) = \frac{1}{n} (\y - \hat{\y})^T (\y - \hat{\y}).
   \end{equation}
@@
   \begin{align}
-    \dpd{\cost(\data, \beta)}{\beta} = \X^T(\y - \X\beta), 
+    \dpd{\cost_\OLS(\data, \beta)}{\beta} = \X^T(\y - \X\beta), 
   \end{align}
@@
   the true value \( f \), while the variance-term measures the fluctuations
-  arising due to using finite samples.
+  arising due to using different samples of the true data.
+
+  Now, note that the equation above suggests that we should minimize both the
+  variance and the bias of the model, in order to obtain the smallest
+  generalization error of the model function. The variance term will be larger
+  for complex models, as performing a fit on one sampled data-set will not
+  neccessarily translate well to another sampled data-set. The bias term
+  measures errors introduced by making certain assumptions about the relations
+  between the covariates \( \X \) and the response \( \y \). In our case, we
+  are assuming a linear relationship (recall \emph{linear regression}), and
+  thus if the true underlying relation is non-linear, our bias will be high.
+  More complex models will in general have lower bias.
+
+  This is known as the \emph{bias-variance tradeoff} as having a highly complex
+  model, we reduce the bias, yet increase the variance.  Having a simple model
+  (for instance a linear model), we increase the bias, and reduce the variance.
 
@@
   \begin{equation}
-    \cost_{\ridge}(\data, \beta;  \lambda) \coloneqq \frac{1}{n} \sum_{i=1}^n (y_i - \hat{y}_i)^2 + \lambda \sum_{i=1}^p \beta_i^2 = \cost(\data, \beta) + \lambda \beta^T \beta,
+    \cost_{\ridge}(\data, \beta;  \lambda) \coloneqq \frac{1}{n} \sum_{i=1}^n
+    (y_i - \hat{y}_i)^2 + \lambda \sum_{i=1}^p \beta_i^2 = \cost(\data, \beta)
+    + \lambda \beta^T \beta,
   \end{equation}
@@
 
+  Attempting to differentiate the Lasso cost function with respect to the model parameters, we see that
+  \begin{equation}
+    \dpd{\cost_\lasso(\data, \beta; \lambda)}{\beta_i} = \dpd{\cost_\OLS(\data, \beta)}{\beta_i} + \lambda \sgn(\beta_i).
+  \end{equation}
+  The analysis of the optimal Lasso parameters is made difficult due to the
+  discontinuity of the sign function, and no closed form solution as in OLS and
+  Ridge-regression can be found in the general case. Thus, numerical
+  optimzation techniques are often used, for instance gradient ascent/descent.
+
   For further information on Ridge and Lasso regression, the excellent article
@@
 
-  \chapter{Implementation}
+  \chapter{Implementation and methodology}
 
-  The implementations are all written in the programming language \python{} with
-  extensive use of the \python-packages \textsf{numpy} and \textsf{sklearn}.
+  In order to verify the implementation, we test on sampled data from the
+  Franke function \cite{frankeCriticalComparisonMethods1979}. This bivariate
+  function \(F(x, y) \) is defined for \( x, y \in [0, 1] \) as follows:
+  \begin{align}
+    \begin{split}
+      F(x,y) \coloneqq &\frac{3}{4} \exp\left( -\frac{(9x - 2)^2}{4} - \frac{(9y - 2)^2}{4}\right) \\
+      + &\frac{3}{4}\exp\left( -\frac{(9x + 1)^2}{49} - \frac{(9 y + 1)^2}{10}\right) \\
+      + &\frac{1}{2}\exp\left(-\frac{(9x-7)^2}{4} - \frac{(9y-3)^2}{4}\right) \\
+      - &\frac{1}{5}\exp\left(-(9x - 4)^2 - (9y - 7)^2\right).
+    \end{split} 
+  \end{align}
+  We are interested in fitting a bivariate polynomial surface to this data, by
+  means of the regression techniques discussed in \cref{chp:regression}. That
+  is, our model function \( f(\data, \beta) \) lies in the polynomial space \(
+  \Pi_{2}^d \) of bi-degree \( (d, d) \) polynomials. We are mainly interested
+  in bi-quintic polynomial surfaces, and will therefore restrict our attention
+  to \( d = 5 \) in the numerical experiments.
+
+  We sample the function \( F \) both with and without noise at \( N = 5000 \) points in each
+  parameter direction, yielding a matrix of function values
+  \begin{align}
+    \begin{split}
+      \mat{Z} &= \left[F(x_i, y_j) \right]_{ij}, \\
+      \mat{Z}^\star &= \left[F(x_i, y_j) + \varepsilon_{ij}\right]_{ij},
+    \end{split}
+  \end{align}
+  where each \( \varepsilon_{ij} \sim \N(0, \sigma^2) \). The design matrix \(
+  \mat{X} \) is computed using the \textsc{PolynomialFeatures}-method from the
+  \textsc{sklearn} package. The dimension of the bi-quintic polynomial space is
+  21, yielding a design matrix of size \( N^2 \times 21 \).
+
+  For the real data set, we use the supplied \textsc{Tiff} data of an area
+  close to Stavanger in Norway, as displayed in \cref{fig:terrain}. The
+  data-set consists of approximately \( 1800 \times 3600 \) data-points. 
+  \begin{figure}
+    \centering
+    \includegraphics[width=1\linewidth]{images/terrain.pdf}
+    \caption{Real terrain data from a region close to Stavanger, Norway. For
+    the regression tasks, we select data from small rectangular regions to keep
+  the data-sizes manageable. The data is supplied in the file
+\texttt{SRTM\_data\_1\_Norway.tif}.}
+  \label{fig:terrain}
+  \end{figure}
+
+  The ordinary least squares and the Ridge regression is performed by solving
+  the corresponding normal equations. We use the Lasso regression available in
+  \textsc{sklearn} package.
 

### 2019-09-12 assignment_one/src/franke_sampling.py

@@ -1,28 +1,11 @@
+import numpy as np
 from matplotlib import cm
 
-from assignment_one.src import ordinary_least_squares as ols
 from assignment_one.src.aux import generate_data, mean_squared_error, r2_score
+from assignment_one.src.ordinary_least_squares import perform_regression, svd_inv
 
 
-def perform_regression(params, z, polynomial_degree=5, ridge=False, l=0.1):
-    """
-    Performs a polynomial fit to sampled data from the Franke function.
-    :return: parameters beta, and sampled points from predicted surface.
-
-    """
-    n = int(np.sqrt(z.shape[0]))
-    v = ols.design_matrix_from_parameters(params, polynomial_degree=polynomial_degree)
-
-    if ridge:
-        beta = ols.ridge_regression(v, z, l=l)
-    else:
-        beta = ols.ordinary_least_squares(v, z)
-    z_hat = (v @ beta).reshape(n, n)
-
-    return beta, z_hat
-
-
-def compute_beta_variance(beta):
+def compute_beta_variance(noise_variance, design_matrix):
     """
     Computes the variance of the model parameters beta.
 
@@ -30,43 +13,59 @@ def compute_beta_variance(beta):
     :return:
     """
 
-    pass
+    return noise_variance * svd_inv(design_matrix.T.dot(design_matrix))
 
 
 if __name__ == '__main__':
 
     import tqdm
-    import numpy as np
     import matplotlib.pyplot as plt
     from mpl_toolkits.mplot3d import Axes3D
+    from sklearn.model_selection import KFold
 
     poly_degs = range(6)
-    N = 50
+    N = 1500
     seed = 42
-    noise = True
+    noise = False
     params, z, X, Y = generate_data(N, noise, seed, return_mesh=True)
 
-    for d in tqdm.tqdm(poly_degs, disable=True):
-        beta, z_hat = perform_regression(params, z, polynomial_degree=d, ridge=True, l=0.1)
+    kf = KFold(n_splits=10)
+    for train_idx, test_idx in kf.split(X):
+        for d in tqdm.tqdm(poly_degs, disable=True):
+            X_train, X_test = X[train_idx], X[test_idx]
+            Y_train, Y_test = Y[train_idx], Y[test_idx]
+            z_train, z_test = z[train_idx], z[test_idx]
 
-        z_hat = z_hat.reshape(N, N)
-        z = z.reshape(N, N)
[…]

### 2019-09-12 assignment_one/src/ordinary_least_squares.py

@@ -94,3 +94,24 @@ if __name__ == '__main__':
     X = design_matrix_from_parameters(x, y, polynomial_degree=1)
     X_ = svd_inv(X)
     b = ordinary_least_squares(X, y_, pseudo_inv=False)
+
+
+def perform_regression(x, y, z, polynomial_degree=5, ridge=False, l=0.1):
+    """
+    Performs a polynomial fit to sampled data.
+    :return: parameters beta, and sampled points from predicted surface.
+
+    """
+    N, M = x.shape[0], y.shape[1]
+    params = np.dstack((x, y)).reshape(-1, 2)
+    n = int(np.sqrt(z.shape[0]))
+    v = design_matrix_from_parameters(params, polynomial_degree=polynomial_degree)
+
+    if ridge:
+        beta = ridge_regression(v, z, l=l)
+    else:
+        beta = ordinary_least_squares(v, z)
+    print(v.shape, beta.shape)
+    z_hat = (v @ beta).reshape(N, M)
+
+    return beta, z_hat

### 2019-09-12 assignment_one/src/regression.py

@@ -1,0 +1,12 @@
+class Regression(object):
+
+    def __init__(self, X, y, lmbd=0):
+        self.X = X
+        self.y = y
+
+    def fit(self, X, y):
+        pass
+
+
+    def predict(self, X):
+        pass

### 2019-09-12 assignment_one/src/terrain_regression.py

@@ -1,0 +1,35 @@
+import numpy as np
+import matplotlib.pyplot as plt
+from mpl_toolkits.mplot3d import Axes3D
+from PIL import Image
+from sklearn.model_selection import KFold
+
+from assignment_one.src.ordinary_least_squares import perform_regression
+
+terrain = np.array(Image.open('../data/SRTM_data_Norway_2.tif'))
+aspect_ratio = terrain.shape[0] / terrain.shape[1]
+print(f'Aspect ratio of terrain data = {aspect_ratio : 2.3f}')
+# terrain = terrain / np.linalg.norm(terrain)
+terrain = (terrain - np.min(terrain)) / np.max(terrain)
+
+plt.imshow(terrain)
+plt.show()
+
+N = terrain.shape[0]
+M = terrain.shape[1]
+
+x = np.linspace(0, 1, N)
+y = np.linspace(0, aspect_ratio, M)
+
+X, Y = np.meshgrid(x, y)
+
+params = np.dstack((X, Y)).reshape(-1, 2)
+beta, z_hat = perform_regression(X, Y, terrain.ravel(), polynomial_degree=2, ridge=False, l=0.1)
+
+fig = plt.figure()
+ax1 = fig.add_subplot(211, projection='3d')
+ax2 = fig.add_subplot(212, projection='3d')
+
+ax1.plot_surface(X, Y, terrain.T)
+ax2.plot_surface(X, Y, z_hat)
+plt.show()

### 2019-09-16 assignment_one/article/main.tex
Working on confidence intervals

@@
 \newcommand{\expect}{\mathbb{E}}
+\newcommand{\var}{\mathbb{V}}
+\newcommand{\cov}{\mathrm{Cov}}
 \newcommand{\OLS}{\mathrm{OLS}}
+\newcommand{\SD}{\mathrm{SD}}
 \newcommand{\ridge}{\mathrm{ridge}}
@@
     \centering
-    \includegraphics[width=1\linewidth]{images/terrain.pdf}
+    \includegraphics[width=1\linewidth]{images/terrain.png}
     \caption{Real terrain data from a region close to Stavanger, Norway. For
@@
   \textsc{sklearn} package.
+
+  \section{Model statistics and resampling methods}
+
+  We are interested in computing certain statistical quantities pertaining to
+  our data and corresponding model function. Firstly, the \emph{variance} of
+  the model parameters \( \beta \) over all possible data sets tells us how
+  sensitive \( \beta \) is to changes in the sampled data set. Using
+  \cref{eq:optimal} for the optimal OLS-parameters and setting \( \mat{A} =
+  (\X^T\X)^{-1}\X^T\), we compute:
+  \begin{align}
+	  \label{eq:ols_variance}
+	  \cov\left[\beta_{\OLS}\right] &= \cov\left[\mat{A}\y\right] = \mat{A} \cov\left[\y\right] \mat{A}^T = \mat{A}\mat{A}^T\sigma^2 = (\X^T{\X})^{-1} \sigma^2.
+  \end{align}
+  In this calculation, we have used \cref{as:general_linear} to deduce that \(
+  \cov[ \y ] = \diag{\sigma}\), a diagonal matrix, which commutes with other
+  matrices. It then follows that the estimated variance of \( \beta_{\OLS} \) is given by
+  \begin{equation}
+	  \var[\beta_{\OLS}] = [\sigma^2 (\X^T\X)^{-1}_{ii} ]_{i=1}^n.
+  \end{equation}
+  Furthermore, by direct computation, we see that \( \beta_{\OLS} \) is an \emph{unbiased} estimator,
+  as 
+  \begin{equation}
+	  \expect[\bias_{\OLS}] = \beta.
+  \end{equation}
+  With both the expected value and the variance of \( \beta \), we may
+  formulate a confidence interval for \( \beta \). Thus for an estimated
+  parameter \( \beta_i \) we construct the
+  \(95\%\)-confidence interval as
+  \begin{align}
+	  \begin{split}
+		  I_i &= (\expect[\beta_i] - 1.96 \times \SD[\beta_i], \expect[\beta_i] + 1.96 \times \SD[\beta_i]) \\
+		      &= \left(\beta_i - 1.96 \times \sigma \sqrt{(\X^T\X)^{-1}_{ii}}, \beta_i + 1.96 \times \sigma \sqrt{(\X^T\X)^{-1}_{ii}}\right).
+	  \end{split}
+  \end{align}
+  Here \( \SD[\beta_i] \) denotes the standard deviation of \( \beta_i \).  The
+  interpretation of this interval is  ``under repeated experiments, the number
+  of computed confidence intervals that actually include the true parameter,
+  will tend to \( 95\% \)''.
 

### 2019-09-17 assignment_one/article/main.tex
Working on article

@@
 \usepackage{titling}
+\usepackage{mdframed}
 \usepackage[]{FiraMono}
@@
 
-\title{\textsc{Regression And Resampling Techniques \\
+\title{\Large\textsc{Regression And Resampling Techniques \\
   FYS-STK4155 \\
@@
   \maketitle 
+  \clearpage
+
+  \begin{abstract}
+	  In this project, we implement and compare three different variants of
+	  linear regression. Namely, ordinary least squares (OLS), Ridge
+	  regression, and Lasso regression. We test the methods on terrain data
+	  taken from a region near Stavanger, Norway. Initial implementation
+	  tests are done on an artificial data-set based on noisy samples of
+	  the Franke function. Statistical quantities pertaining to the various
+	  model functions are computed and compared. We also give an
+	  introduction to the theory and motivation behind the three methods.
+  \end{abstract}
 
@@
 
+  \chapter{Conclusion}
+
+  \section{Further work}
+
+  In this project we have discussed three different methods of linear
+  regression.  In all three cases, we have fitted polynomial surfaces to the
+  data sets. One draw-back with using polynomial surfaces is that they are
+  inherently global methods. Thus, in order to capture local trends in the
+  data, a very complex model is required, which (in the polynomial case) means
+  surfaces of higher polynomial degree. In recent years, several attempts at
+  constructing efficient local methods have been proposed. Spline-based
+  (piecewise polynomial) methods in particular have been important, and several
+  approaches to locally refined spline spaces have started to gain momentum.
+  We note in particular that of \emph{hierarchical B-splines} (HB),
+  \emph{truncated hierarchical B-splines} (THB) and \emph{locally refined
+  B-splines} (LRB). These methods allow us to place more degrees of freedom in regions of high
+  data-variance, thus locally increasing the model complexity.
+
+  It would be interesting to make a comparison of the three methods in such a
+  simple setting as linear regression, as done in this project, taking into
+  account adaptive refinement of the spline spaces with respect to the sampled
+  data set.
+
   \appendix

### 2019-09-18 assignment_one/src/regression_class.py
Added regression classes

@@ -1,0 +1,81 @@
+import numpy as np
+import sklearn.linear_model
+
+
+class Regression(object):
+
+    def __init__(self, X, y, inversion_method='direct'):
+        self.X = X
+        self.y = y
+        self.xtx = np.dot(X.T, X)
+        self.inv = np.linalg.inv if inversion_method == 'direct' else None
+
+    def fit(self):
+        raise NotImplementedError()
+
+    def predict(self):
+        return self.X.dot(self.beta)
+
+    @property
+    def beta(self):
+        raise NotImplementedError()
+
+
+class OLS(Regression):
+
+    @property
+    def beta(self):
+        if hasattr(self, '_beta'):
+            return self._beta
+        else:
+            self._beta = self.inv(self.xtx).dot(self.X.T).dot(self.y)
+            return self._beta
+
+    def fit(self):
+        xtx_inv = self.inv(self.xtx)
+        self._beta = xtx_inv.dot(self.X.T).dot(self.y)
+
+
+class Ridge(Regression):
+
+    def fit(self):
+        pass
+
+    def __init__(self, X, y, inversion_method='direct', lmbd=0.1):
+        super().__init__(X, y, inversion_method)
+        self.lmbd = lmbd
+
+    @property
+    def beta(self):
+        if hasattr(self, '_beta'):
+            return self._beta
+        else:
+            self._beta = self.inv(self.xtx + self.lmbd * np.eye(self.xtx.shape[0])).dot(self.X.T).dot(self.y)
+            return self._beta
+
+
+class Lasso(Regression):
+
+    def __init__(self, X, y, inversion_method='direct', lmbd=0.1):
+        super().__init__(X, y, inversion_method)
+        self.lmbd = lmbd
+        self._regressor = sklearn.linear_model.Lasso(alpha=self.lmbd, fit_intercept=False)
+        self._regressor.fit(self.X, self.y)
+
+    @property
+    def beta(self):
+        return self._regressor.coef_
+
+
+if __name__ == '__main__':
+    x = np.random.randint(0, 10, (10, 6))
+    y = np.random.randint(0, 10, (10,))
+
+    O = OLS(x, y)
+    R = Ridge(x, y, lmbd=0)
+    L = Lasso(x, y, lmbd=0)
+    print(O.beta)
+    print(R.beta)
+    print(L.beta)
+
+    print(O.predict())

### 2019-09-18 assignment_one/src/regression_class.py
Added variance estimates for parameters.

@@ -14,15 +14,31 @@ class Regression(object):
         raise NotImplementedError()
 
     def predict(self):
-        return self.X.dot(self.beta)
+        self.y_hat = self.X.dot(self.beta)
+        return self.y_hat
+
+    @property
+    def y_variance_estimate(self):
+        if not hasattr(self, 'y_hat'):
+            self.predict()
+        n, p = self.X.shape
+
+        return 1 / (n - p - 1) * np.sum((self.y - self.y_hat) ** 2)
 
     @property
     def beta(self):
         raise NotImplementedError()
 
+    @property
+    def beta_covariance_estimate(self):
+        return self.inv(self.xtx) * self.y_variance_estimate
+
+    @property
+    def beta_variance_estimate(self):
+        return np.diag(self.beta_covariance_estimate)
+
 
 class OLS(Regression):
-
     @property
     def beta(self):
         if hasattr(self, '_beta'):
@@ -72,10 +88,13 @@ if __name__ == '__main__':
     y = np.random.randint(0, 10, (10,))
 
     O = OLS(x, y)
-    R = Ridge(x, y, lmbd=0)
-    L = Lasso(x, y, lmbd=0)
+    R = Ridge(x, y, lmbd=0.1)
+    L = Lasso(x, y, lmbd=0.1)
     print(O.beta)
     print(R.beta)
     print(L.beta)
 
     print(O.predict())
+    for e in O, R, L:
+        print(e.y_variance_estimate)
+        print(e.beta_variance_estimate)

### 2019-09-18 assignment_one/src/regression_class.py
Added some docstrings.

@@ -19,6 +19,13 @@ class Regression(object):
 
     @property
     def y_variance_estimate(self):
+        """
+        Return an unbiased estimate of the response variance (assuming all y_i are uncorrelated
+        and constant variance sigma^2).
+
+        :return: estimate of sigma^2
+        """
+
         if not hasattr(self, 'y_hat'):
             self.predict()
         n, p = self.X.shape
@@ -31,10 +38,21 @@ class Regression(object):
 
     @property
     def beta_covariance_estimate(self):
+        """
+        Returns the estimated covariance matrix for beta, which is a matrix of shape (n, n)
+
+        :return: covariance matrix estimate
+        """
         return self.inv(self.xtx) * self.y_variance_estimate
 
     @property
     def beta_variance_estimate(self):
+        """
+        Returns the estimated variances for beta, simply by extracting the diagonal from the covariance matrix.
+        This is an array of shape (n,)
+
+        :return: variance of beta
+        """
         return np.diag(self.beta_covariance_estimate)
 
 

### 2019-09-18 assignment_one/src/regression_class.py
Added performance metrics - r2 score and mse score.

@@ -55,6 +55,32 @@ class Regression(object):
         """
         return np.diag(self.beta_covariance_estimate)
 
+    def mse(self):
+        """
+        Computes the mean squared error of the prediction y_hat to the ground truth y.
+
+        :return: mean squared error
+        """
+
+        n = self.y.shape[0]
+        return ((self.y - self.y_hat) ** 2).sum() / n
+
+    def r2(self):
+        """
+        Computes the r2-score (or coefficient of determination) of the prediction y_hat to the ground truth y.
+
+        :param y: ground truth
+        :param y_hat: prediction
+        :return: r2-score
+        """
+
+        y_mean = self.y.mean()
+
+        num = ((self.y - self.y_hat) ** 2).sum()
+        den = ((self.y - y_mean) ** 2).sum()
+
+        return 1 - num / den
+
 
 class OLS(Regression):
     @property
@@ -102,12 +128,14 @@ class Lasso(Regression):
 
 
 if __name__ == '__main__':
-    x = np.random.randint(0, 10, (10, 6))
-    y = np.random.randint(0, 10, (10,))
+    x = np.linspace(0, 1, 10000)
+    X = np.ones((10000, 2))
+    X[:, 1] = x
+    y = 4*x**4 + np.random.normal(0, np.sqrt(9), 10000)
 
-    O = OLS(x, y)
-    R = Ridge(x, y, lmbd=0.1)
-    L = Lasso(x, y, lmbd=0.1)
+    O = OLS(X, y)
+    R = Ridge(X, y, lmbd=0.1)
+    L = Lasso(X, y, lmbd=0.1)
     print(O.beta)
     print(R.beta)
     print(L.beta)
@@ -116,3 +144,4 @@ if __name__ == '__main__':
     for e in O, R, L:
         print(e.y_variance_estimate)
         print(e.beta_variance_estimate)
+        print(e.r2(), e.mse())

### 2019-09-18 assignment_one/src/regression_class.py
Added performance metrics - r2 score and mse score.

@@ -131,7 +131,7 @@ if __name__ == '__main__':
     x = np.linspace(0, 1, 10000)
     X = np.ones((10000, 2))
     X[:, 1] = x
-    y = 4*x**4 + np.random.normal(0, np.sqrt(9), 10000)
+    y = 4 * x ** 4 + np.random.normal(0, np.sqrt(9), 10000)
 
     O = OLS(X, y)
     R = Ridge(X, y, lmbd=0.1)

### 2019-09-18 assignment_one/src/regression.py
Deleted old regression class. Added SVD-inv as a static-method to Regression base class.

@@ -1,12 +1,0 @@
-class Regression(object):
-
-    def __init__(self, X, y, lmbd=0):
-        self.X = X
-        self.y = y
-
-    def fit(self, X, y):
-        pass
-
-
-    def predict(self, X):
-        pass

### 2019-09-18 assignment_one/src/regression_class.py

@@ -81,6 +81,25 @@ class Regression(object):
 
         return 1 - num / den
 
+    @staticmethod
+    def svd_inv(A):
+        """
+        Returns the inverse of the matrix A based on a singular value decomposition of A.
+
+        :param A: matrix to invert
+        :return: inv(A)
+        """
+
+        u, sigma, v_transposed = np.linalg.svd(A)
+        m = u.shape[0]
+        n = v_transposed.shape[0]
+
+        D = np.zeros((m, n))
+        for i in range(n):
+            D[i, i] = sigma[i]
+
+        return v_transposed.T @ (np.linalg.pinv(D) @ u.T)
+
 
 class OLS(Regression):
     @property

### 2019-09-18 assignment_one/src/regression_class.py
Added standard deviation property.

@@ -8,7 +8,7 @@ class Regression(object):
         self.X = X
         self.y = y
         self.xtx = np.dot(X.T, X)
-        self.inv = np.linalg.inv if inversion_method == 'direct' else None
+        self.inv = np.linalg.inv if inversion_method == 'direct' else Regression.svd_inv
 
     def fit(self):
         raise NotImplementedError()
@@ -54,6 +54,15 @@ class Regression(object):
         :return: variance of beta
         """
         return np.diag(self.beta_covariance_estimate)
+
+    @property
+    def beta_std_dev_estimate(self):
+        """
+        Returns the estimated standard deviations for beta by taking the square root of the variance estimate.
+
+        :return: standard deviation of beta
+        """
+        return np.sqrt(self.beta_variance_estimate)
 
     def mse(self):
         """
@@ -152,7 +161,7 @@ if __name__ == '__main__':
     X[:, 1] = x
     y = 4 * x ** 4 + np.random.normal(0, np.sqrt(9), 10000)
 
-    O = OLS(X, y)
+    O = OLS(X, y, inversion_method='svd')
     R = Ridge(X, y, lmbd=0.1)
     L = Lasso(X, y, lmbd=0.1)
     print(O.beta)

### 2019-09-18 assignment_one/src/regression_class.py
Added optional X-input for validation prediction!

@@ -13,9 +13,18 @@ class Regression(object):
     def fit(self):
         raise NotImplementedError()
 
-    def predict(self):
-        self.y_hat = self.X.dot(self.beta)
-        return self.y_hat
+    def predict(self, X=None):
+        if X is None:
+            X = self.X
+        self._y_hat = X.dot(self.beta)
+        return self._y_hat
+
+    @property
+    def y_hat(self):
+        if not hasattr(self, '_y_hat'):
+            return self.predict()
+        else:
+            return self._y_hat
 
     @property
     def y_variance_estimate(self):

### 2019-09-18 assignment_one/article/main.tex
Added terrain.pdf image
Added optional X-input for validation prediction!

@@
 \documentclass[dvipsnames, article, a4paper, oneside, 12pt]{memoir}
+\newsubfloat{figure}
 \linespread{1.25}
@@
 \newcommand{\expect}{\mathbb{E}}
-\newcommand{\var}{\mathbb{V}}
-\newcommand{\cov}{\mathrm{Cov}}
+\DeclareMathOperator{\var}{Var}
+\DeclareMathOperator{\cov}{Cov}
 \newcommand{\OLS}{\mathrm{OLS}}
-\newcommand{\SD}{\mathrm{SD}}
+\DeclareMathOperator{\SD}{SD}
 \newcommand{\ridge}{\mathrm{ridge}}
@@
       \mat{Z} &= \left[F(x_i, y_j) \right]_{ij}, \\
-      \mat{Z}^\star &= \left[F(x_i, y_j) + \varepsilon_{ij}\right]_{ij},
+      \mat{Z}^\star &= \left[F(x_i, y_j) + \alpha\varepsilon_{ij}\right]_{ij},
     \end{split}
   \end{align}
-  where each \( \varepsilon_{ij} \sim \N(0, \sigma^2) \). The design matrix \(
-  \mat{X} \) is computed using the \textsc{PolynomialFeatures}-method from the
-  \textsc{sklearn} package. The dimension of the bi-quintic polynomial space is
-  21, yielding a design matrix of size \( N^2 \times 21 \).
+  where each \( \varepsilon_{ij} \sim \N(0, \sigma^2) \) and \( \alpha \) is
+  the \emph{signal-to-noise ratio}. The design matrix \( \mat{X} \) is computed
+  using the \textsc{PolynomialFeatures}-method from the \textsc{sklearn}
+  package. The dimension of the bi-quintic polynomial space is 21, yielding a
+  design matrix of size \( N^2 \times 21 \).
 
@@
     \centering
-    \includegraphics[width=1\linewidth]{images/terrain.png}
+    \includegraphics[width=1\linewidth]{images/terrain.pdf}
     \caption{Real terrain data from a region close to Stavanger, Norway. For
@@
   \begin{equation}
-	  \expect[\bias_{\OLS}] = \beta.
+	  \expect[\beta_{\OLS}] = \beta.
   \end{equation}
@@
 
+  In this section we present some of the results of the numerical simulations.
+  We start simple by first considering the effects of model complexity --- in
+  terms of polynomial degree --- using OLS regression on the Franke dataset. We
+  assess the model accuracy by means of the mean squared error and the \(
+  R^2\)-score discussed previously. In order to increase the reliability of the
+  statistical estimates, we employ \( k\)-fold cross validation.
+
+  We then take a closer look at Ridge and Lasso regression techniques, and find
+  the optimal hyper-parameter \( \lambda \). Again, we use the Franke dataset.
+  In this context, we also see the effects of the bias-variance tradeoff.
+
+  Finally, we compare our methods on real world data, namely the terrain data
+  in \cref{fig:terrain}.
+
   \section{Toy Data}
 
+  \begin{figure}
+    \centering
+    \subbottom{%
+      \includegraphics[width=0.45\linewidth]{images/OLS_MSE_score.pdf}
+    }
+    \subbottom{%
+      \includegraphics[width=0.45\linewidth]{images/OLS_R2_score.pdf}
+    }
+    \caption{Ordinary least squares regression on samples from the Franke
+      function. The mean squared error (left) and the \(R^2\)-score (right) as
+      a function of the polynomial degree \( d \). The values are displayed
+      both for the true sample data (solid) and the noisy samples (dashed). The
+      noise is distributed as \( \N(0, \tfrac{5}{100})\).}
+    \label{fig:ols_scores}
+  \end{figure}
+
+  \subsubsection{Ordinary Least Squares}
+
+  We sample \( N = 1000 \) points in each parameter direction and compute the
+  corresponding \( N^2 \) funciton values of the Franke function both with and
+  without added noise. The noise was selected to be distributed as \( \N(0,
+  1) \) with a signal-to-noise ratio \( \alpha = 0.05\).
+
+  We then perform ordinary least squares, and compute the mean squared error as
+  well as the \( R^2 \)-score for both the true and the noisy data. These are
+  presented in \cref{fig:ols_scores}. With the Franke function being a weighted
+  sum of exponentials, the MSE-error tends to zero as the polynomial degree
+  increases. The added noise reduces the convergence rate, as is to be
+  expected.
+
+  In addition to the evaluation metrics, we also compute confidence intervals
+  for each model parameter. These are displayed in \cref{tab:ols_confidence}
+  for each polynomial degree up to and including \( d = 5 \).
+
+  \begin{table}
+    \caption{}
+    \label{tab:ols_confidence}
+  \end{table}
   \section{Terrain Data}

### 2019-09-18 assignment_one/src/aux.py

@@ -18,7 +18,7 @@ def franke_function(x, y):
     return a + b + c - d
 
 
-def mean_squared_error(y, y_hat):
+def mse(y, y_hat):
     """
     Computes the mean squared error of the prediction y_hat to the ground truth y.
 
@@ -31,7 +31,7 @@ def mean_squared_error(y, y_hat):
     return ((y - y_hat) ** 2).sum() / n
 
 
-def r2_score(y, y_hat):
+def r2(y, y_hat):
     """
     Computes the r2-score (or coefficient of determination) of the prediction y_hat to the ground truth y.
 

### 2019-09-18 assignment_one/src/franke_sampling.py

@@ -1,71 +1,0 @@
-import numpy as np
-from matplotlib import cm
-
-from assignment_one.src.aux import generate_data, mean_squared_error, r2_score
-from assignment_one.src.ordinary_least_squares import perform_regression, svd_inv
-
-
-def compute_beta_variance(noise_variance, design_matrix):
-    """
-    Computes the variance of the model parameters beta.
-
-    :param beta:
-    :return:
-    """
-
-    return noise_variance * svd_inv(design_matrix.T.dot(design_matrix))
-
-
-if __name__ == '__main__':
-
-    import tqdm
-    import matplotlib.pyplot as plt
-    from mpl_toolkits.mplot3d import Axes3D
-    from sklearn.model_selection import KFold
-
-    poly_degs = range(6)
-    N = 1500
-    seed = 42
-    noise = False
-    params, z, X, Y = generate_data(N, noise, seed, return_mesh=True)
-
-    kf = KFold(n_splits=10)
-    for train_idx, test_idx in kf.split(X):
-        for d in tqdm.tqdm(poly_degs, disable=True):
-            X_train, X_test = X[train_idx], X[test_idx]
-            Y_train, Y_test = Y[train_idx], Y[test_idx]
-            z_train, z_test = z[train_idx], z[test_idx]
-
-            print(X_train)
-            beta, z_hat = perform_regression(X_train, Y_train, z_train, polynomial_degree=d, ridge=True, l=0.1)
-
-            M = X_train.shape[0]
-            z_hat = z_hat.reshape(M, M)
-            z_train = z_train.reshape(M, M)
-
-            fig = plt.figure()
-            ax1 = fig.add_subplot(211, projection='3d')
-            ax2 = fig.add_subplot(212, projection='3d')
-
-            ax1.view_init(azim=30)
-            ax2.view_init(azim=30)
-            ax1.set_zlim3d(-0.2, 1.2)
-
-            ax1.plot_surface(X, Y, z_hat, alpha=0.5, cmap=cm.coolwarm)
-            # ax1.scatter(X, Y, z, alpha=1, s=1, color='black')
-
-            ax2.plot_surface(X, Y, np.abs(z_hat - z), cmap=cm.coolwarm)
-            ax2.set_zlim3d(0, 1)
-
-            plt.title(f'Polynomial degree $p = {d}$')
-            plt.show()
-
-            z_hat = z_hat.ravel()
-            z = z.ravel()
-
-            print('Polynomial degree = ', d)
[…]

### 2019-09-18 assignment_one/src/part_a.py

@@ -1,0 +1,55 @@
+import numpy as np
+import sklearn.preprocessing
+import matplotlib.pyplot as plt
+import tqdm
+from matplotlib import cm
+from mpl_toolkits.mplot3d import Axes3D
+from assignment_one.src.aux import franke_function
+from assignment_one.src.plotting import latexify, format_axes
+from assignment_one.src.regression_class import OLS
+
+if __name__ == '__main__':
+    N = 1000
+    polynomial_degrees = range(0, 11)
+
+    x = np.linspace(0, 1, N)
+    y = np.linspace(0, 1, N)
+
+    X, Y = np.meshgrid(x, y)
+    cartesian_product = np.dstack((X, Y)).reshape(-1, 2)
+    z_grid = franke_function(X, Y)
+    z_grid_noise = z_grid + np.random.normal(0, 0.05, size=(z_grid.shape))
+    z = z_grid.ravel()
+    z_noise = z_grid_noise.ravel()
+    mse_scores = []
+    r2_scores = []
+    mse_scores_noise = []
+    r2_scores_noise = []
+    for d in tqdm.tqdm(polynomial_degrees):
+        design_matrix = sklearn.preprocessing.PolynomialFeatures(degree=d).fit_transform(cartesian_product)
+        ols = OLS(design_matrix, z)
+        ols_noise = OLS(design_matrix, z_noise)
+
+        mse_scores.append(ols.mse())
+        r2_scores.append(ols.r2())
+        mse_scores_noise.append(ols_noise.mse())
+        r2_scores_noise.append(ols_noise.r2())
+
+
+    latexify(3)
+    plt.semilogy(polynomial_degrees, mse_scores, ls='-', marker='o')
+    plt.semilogy(polynomial_degrees, mse_scores_noise, ls='--', marker='^')
+    plt.xlabel('$d$')
+    plt.ylabel('$\\mathrm{MSE}(\\mathbf{y}, \\hat{\\mathbf{y}})$')
+    plt.tight_layout()
+    format_axes(ax=plt.gca())
+    plt.savefig('../article/images/OLS_MSE_score.pdf')
+    plt.clf()
+
+    plt.plot(polynomial_degrees, r2_scores, ls='-', marker='o')
+    plt.plot(polynomial_degrees, r2_scores_noise, ls='--', marker='^')
+    plt.xlabel('$d$')
[…]

### 2019-09-18 assignment_one/src/part_b.py

@@ -1,0 +1,73 @@
+import numpy as np
+import sklearn.preprocessing
+import sklearn.model_selection
+import tqdm
+import matplotlib.pyplot as plt
+
+from assignment_one.src.aux import franke_function, mse, r2
+from assignment_one.src.plotting import latexify, format_axes
+from assignment_one.src.regression_class import OLS
+
+if __name__ == '__main__':
+    N = 1000
+
+    polynomial_degrees = range(0, 10)
+
+    x = np.linspace(0, 1, N)
+    y = np.linspace(0, 1, N)
+
+    X, Y = np.meshgrid(x, y)
+    cartesian_product = np.dstack((X, Y)).reshape(-1, 2)
+    z_grid = franke_function(X, Y)
+    z_grid_noise = z_grid + np.random.normal(0, 0.05, size=(z_grid.shape))
+    z = z_grid.ravel()
+    z_noise = z_grid_noise.ravel()
+
+    number_of_folds = 5
+    kf = sklearn.model_selection.KFold(n_splits=number_of_folds, shuffle=True)
+
+    mse_s = []
+    r2_s = []
+    beta_variance = []
+    beta_std = []
+    for d in tqdm.tqdm(polynomial_degrees):
+        mse_scores = []
+        r2_scores = []
+        bvar = []
+        for train_idx, test_idx in kf.split(cartesian_product):
+            x_train, x_test = cartesian_product[train_idx], cartesian_product[test_idx]
+            z_train, z_test = z[train_idx], z[test_idx]
+
+            X_train = sklearn.preprocessing.PolynomialFeatures(degree=d).fit_transform(x_train)
+            X_test = sklearn.preprocessing.PolynomialFeatures(degree=d).fit_transform(x_test)
+
+            ols = OLS(X_train, z_train)
+
+            z_hat_train = ols.predict(X_train)
+            z_hat_test = ols.predict(X_test)
+
+            mse_scores.append((mse(z_train, z_hat_train), mse(z_test, z_hat_test)))
+            r2_scores.append((r2(z_train, z_hat_train), r2(z_test, z_hat_test)))
+            bvar.append(ols.beta_variance_estimate)
+        mse_s.append(np.array(mse_scores).mean(axis=0))
+        r2_s.append(np.array(r2_scores).mean(axis=0))
[…]

### 2019-09-18 assignment_one/src/plotting.py

@@ -1,0 +1,87 @@
+import matplotlib
+
+SPINE_COLOR = 'gray'
+
+
+def latexify(fig_width=None, fig_height=None, columns=1):
+    """Set up matplotlib's RC params for LaTeX plotting.
+    Call this before plotting a figure.
+
+    Parameters
+    ----------
+    fig_width : float, optional, inches
+    fig_height : float,  optional, inches
+    columns : {1, 2}
+    """
+
+    # code adapted from http://www.scipy.org/Cookbook/Matplotlib/LaTeX_Examples
+
+    # Width and max height in inches for IEEE journals taken from
+    # computer.org/cms/Computer.org/Journal%20templates/transactions_art_guide.pdf
+
+    from math import sqrt
+    import matplotlib
+
+    assert columns in {1, 2}
+    if fig_width is None:
+        fig_width = 3.39 if columns==1 else 6.9 # width in inches
+
+    if fig_height is None:
+        golden_mean = (sqrt(5)-1.0)/2.0    # Aesthetic ratio
+        fig_height = fig_width*golden_mean # height in inches
+
+    MAX_HEIGHT_INCHES = 8.0
+    if fig_height > MAX_HEIGHT_INCHES:
+        print("WARNING: fig_height too large:" + fig_height +
+              "so will reduce to" + MAX_HEIGHT_INCHES + "inches.")
+        fig_height = MAX_HEIGHT_INCHES
+
+    params = {
+              'text.latex.preamble': [r'\usepackage{microtype}', r'\usepackage{libertine}', r'\usepackage{libertinust1math}'],
+              'axes.labelsize': 8, # fontsize for x and y labels (was 10)
+              'axes.titlesize': 8,
+              'axes.titlepad' : 20,
+              'font.size': 12, # was 10
+              'legend.fontsize': 8, # was 10
+              'xtick.labelsize': 8,
+              'ytick.labelsize': 8,
+              'text.usetex': True,
+              'figure.figsize': [fig_width,fig_height],
+              'font.family' : 'serif',
+              'lines.markersize' : 2,
+              'lines.linewidth' : 0.7,
+              'axes.prop_cycle' : "cycler('linestyle', ['solid', 'dashed', 'dashdot', 'dotted'])"
+    }
+
+    matplotlib.rcParams.update(params)
+
+
[…]

### 2019-09-18 assignment_one/src/regression_class.py

@@ -16,8 +16,10 @@ class Regression(object):
     def predict(self, X=None):
         if X is None:
             X = self.X
-        self._y_hat = X.dot(self.beta)
-        return self._y_hat
+        y_hat = X.dot(self.beta)
+        if X is None:
+            self._y_hat = y_hat
+        return y_hat
 
     @property
     def y_hat(self):

### 2019-09-19 assignment_one/src/aux.py
dsd

@@ -1,94 +1,0 @@
-import numpy as np
-
-
-def franke_function(x, y):
-    """
-    Evaluates the franke function over the mesh-grid X, Y.
-
-    :param x: x values
-    :param y: y values
-    :return: F(x, y)
-    """
-
-    a = 3 / 4 * np.exp(-(9 * x - 2) ** 2 / 4 - (9 * y - 2) ** 2 / 4)
-    b = 3 / 4 * np.exp(-(9 * x + 1) ** 2 / 49 - (9 * y + 1) / 10)
-    c = 1 / 2 * np.exp(-(9 * x - 7) ** 2 / 4 - (9 * y - 3) ** 2 / 4)
-    d = 1 / 5 * np.exp(-(9 * x - 4) ** 2 - (9 * y - 7) ** 2)
-
-    return a + b + c - d
-
-
-def mean_squared_error(y, y_hat):
-    """
-    Computes the mean squared error of the prediction y_hat to the ground truth y.
-
-    :param y: ground truth
-    :param y_hat: prediction
-    :return: mean squared error
-    """
-
-    n = y.shape[0]
-    return ((y - y_hat) ** 2).sum() / n
-
-
-def r2_score(y, y_hat):
-    """
-    Computes the r2-score (or coefficient of determination) of the prediction y_hat to the ground truth y.
-
-    :param y: ground truth
-    :param y_hat: prediction
-    :return: r2-score
-    """
-
-    y_mean = y.mean()
-
-    num = ((y - y_hat) ** 2).sum()
-    den = ((y - y_mean) ** 2).sum()
-
-    return 1 - num / den
-
-
-def generate_data(N, noise=True, seed=None, return_mesh=False):
-    """
-    Samples (optionally noisy) data from the franke function with n parameter values in
-    each parameter direction.
-
-    :param N: number of values to sample in each parameter direction
-    :param noise: whether to sample with added noise or not
-    :param seed: optional seed for the RNG
-    :return: x-values, y-values and F(x,y)-values in raveled arrays.
-    """
-    if noise and seed is not None:
-        np.random.seed(seed)
-
-    x = np.linspace(0, 1, N)
-    y = np.linspace(0, 1, N)
-
-    X, Y = np.meshgrid(x, y)
-    Z = franke_function(X, Y).ravel()
-    params = np.dstack((X, Y)).reshape(-1, 2)
-
-    if noise:
[…]

### 2019-09-19 assignment_one/src/regression_class.py

@@ -1,0 +1,89 @@
+import numpy as np
+
+
+class Regression():
+
+    def __init__(self, X, y, inversion_method='direct'):
+        self.X = X
+        self.y = y
+        self.XTX = np.dot(X.T, X)
+        self.inv_method = np.linalg.inv if inversion_method == 'direct' else None
+
+    def fit(self, X, y):
+        raise NotImplementedError()
+
+    def predict(self):
+        """
+        Returns the prediction corresponding to optimal model parameters
+        :return: y = self.X.dot(self.beta)
+        """
+
+        return self.X.dot(self.beta)
+
+
+class OLS(Regression):
+
+    def fit(self):
+        """
+        Returns the analytically optimal model parameters for ordinary least squares.
+        :return:
+        """
+
+        if hasattr(self, 'beta'):
+            return self.beta
+        else:
+            xtx_inv = self.inv_method(self.XTX)
+            self.beta = xtx_inv.dot(self.X.T).dot(self.y)
+            return self.beta
+
+
+class Ridge(Regression):
+
+    def __init__(self, X, y, inversion_method='direct', lambd=1.0):
+        super().__init__(X, y, inversion_method)
+        self.lambd = lambd
+
+    def fit(self):
+        """
+        Returns the analytically optimal model parameters for ordinary least squares.
+        :return:
+        """
+
+        if hasattr(self, 'beta'):
+            return self.beta
+        else:
+            xtx_inv = self.inv_method(self.XTX + self.lambd * np.eye(self.XTX.shape[0]))
+            self.beta = xtx_inv.dot(self.X.T).dot(self.y)
+            return self.beta
+
+
+if __name__ == '__main__':
+    x = np.linspace(-1, 5, 20)
+    X = np.ones((20, 3))
+    X[:, 1] = x
+    X[:, 2] = x ** 2
+    # X = sklearn.preprocessing.PolynomialFeatures(degree=2).fit_transform(x)
+    print(X.shape)
+    y = 3 * x ** 3 + x
+
+    import matplotlib.pyplot as plt
+
+    O = OLS(X, y)
+    b = O.fit()
+    y_hat = O.predict()
+    plt.plot(x, y_hat, label='OLS')
+    for l in [10 ** (- i - 1) for i in range(4)]:
+        R = Ridge(X, y, lambd=l)
+        b = R.fit()
+        y_hat = R.predict()
[…]

### 2019-09-19 assignment_one/src/aux.py
Move stuff over to auxiliary as aux.py was not functioning properly for some reason.

@@ -1,94 +1,0 @@
-import numpy as np
-
-
-def franke_function(x, y):
-    """
-    Evaluates the franke function over the mesh-grid X, Y.
-
-    :param x: x values
-    :param y: y values
-    :return: F(x, y)
-    """
-
-    a = 3 / 4 * np.exp(-(9 * x - 2) ** 2 / 4 - (9 * y - 2) ** 2 / 4)
-    b = 3 / 4 * np.exp(-(9 * x + 1) ** 2 / 49 - (9 * y + 1) / 10)
-    c = 1 / 2 * np.exp(-(9 * x - 7) ** 2 / 4 - (9 * y - 3) ** 2 / 4)
-    d = 1 / 5 * np.exp(-(9 * x - 4) ** 2 - (9 * y - 7) ** 2)
-
-    return a + b + c - d
-
-
-def mse(y, y_hat):
-    """
-    Computes the mean squared error of the prediction y_hat to the ground truth y.
-
-    :param y: ground truth
-    :param y_hat: prediction
-    :return: mean squared error
-    """
-
-    n = y.shape[0]
-    return ((y - y_hat) ** 2).sum() / n
-
-
-def r2(y, y_hat):
-    """
-    Computes the r2-score (or coefficient of determination) of the prediction y_hat to the ground truth y.
-
-    :param y: ground truth
-    :param y_hat: prediction
-    :return: r2-score
-    """
-
-    y_mean = y.mean()
-
-    num = ((y - y_hat) ** 2).sum()
-    den = ((y - y_mean) ** 2).sum()
-
-    return 1 - num / den
-
-
-def generate_data(N, noise=True, seed=None, return_mesh=False):
-    """
-    Samples (optionally noisy) data from the franke function with n parameter values in
-    each parameter direction.
-
-    :param N: number of values to sample in each parameter direction
-    :param noise: whether to sample with added noise or not
-    :param seed: optional seed for the RNG
-    :return: x-values, y-values and F(x,y)-values in raveled arrays.
-    """
-    if noise and seed is not None:
-        np.random.seed(seed)
-
-    x = np.linspace(0, 1, N)
-    y = np.linspace(0, 1, N)
-
-    X, Y = np.meshgrid(x, y)
-    Z = franke_function(X, Y).ravel()
-    params = np.dstack((X, Y)).reshape(-1, 2)
-
-    if noise:
[…]

### 2019-09-19 assignment_one/src/auxiliary.py

@@ -1,0 +1,48 @@
+import numpy as np
+
+
+def franke_function(x, y):
+    """
+    Evaluates the franke function over the mesh-grid X, Y.
+
+    :param x: x values
+    :param y: y values
+    :return: F(x, y)
+    """
+
+    a = 3 / 4 * np.exp(-(9 * x - 2) ** 2 / 4 - (9 * y - 2) ** 2 / 4)
+    b = 3 / 4 * np.exp(-(9 * x + 1) ** 2 / 49 - (9 * y + 1) / 10)
+    c = 1 / 2 * np.exp(-(9 * x - 7) ** 2 / 4 - (9 * y - 3) ** 2 / 4)
+    d = 1 / 5 * np.exp(-(9 * x - 4) ** 2 - (9 * y - 7) ** 2)
+
+    return a + b + c - d
+
+
+def mse(y, y_hat):
+    """
+    Computes the mean squared error of the prediction y_hat to the ground truth y.
+
+    :param y: ground truth
+    :param y_hat: prediction
+    :return: mean squared error
+    """
+
+    n = y.shape[0]
+    return ((y - y_hat) ** 2).sum() / n
+
+
+def r2(y, y_hat):
+    """
+    Computes the r2-score (or coefficient of determination) of the prediction y_hat to the ground truth y.
+
+    :param y: ground truth
+    :param y_hat: prediction
+    :return: r2-score
+    """
+
+    y_mean = y.mean()
+
+    num = ((y - y_hat) ** 2).sum()
+    den = ((y - y_mean) ** 2).sum()
+
+    return 1 - num / den

### 2019-09-19 assignment_one/src/bias_variance_tradeoff.py

@@ -1,0 +1,72 @@
+import numpy as np
+import sklearn.model_selection
+import tqdm as tqdm
+import matplotlib.pyplot as plt
+
+from assignment_one.src.auxiliary import franke_function, mse, r2
+from assignment_one.src.plotting import latexify, format_axes
+from assignment_one.src.regression_class import OLS
+
+if __name__ == '__main__':
+    N = 1000
+
+    polynomial_degrees = range(0, 10)
+
+    x = np.linspace(0, 1, N)
+    y = np.linspace(0, 1, N)
+
+    X, Y = np.meshgrid(x, y)
+    cartesian_product = np.dstack((X, Y)).reshape(-1, 2)
+    z_grid = franke_function(X, Y)
+    z_grid_noise = z_grid + np.random.normal(0, 0.05, size=(z_grid.shape))
+    z = z_grid.ravel()
+    z_noise = z_grid_noise.ravel()
+
+    number_of_folds = 5
+    kf = sklearn.model_selection.KFold(n_splits=number_of_folds, shuffle=True)
+
+    mse_s = []
+    r2_s = []
+    beta_variance = []
+    beta_std = []
+    for d in tqdm.tqdm(polynomial_degrees):
+        mse_scores = []
+        r2_scores = []
+        bvar = []
+        for train_idx, test_idx in kf.split(cartesian_product):
+            x_train, x_test = cartesian_product[train_idx], cartesian_product[test_idx]
+            z_train, z_test = z[train_idx], z[test_idx]
+
+            X_train = sklearn.preprocessing.PolynomialFeatures(degree=d).fit_transform(x_train)
+            X_test = sklearn.preprocessing.PolynomialFeatures(degree=d).fit_transform(x_test)
+
+            ols = OLS(X_train, z_train)
+
+            z_hat_train = ols.predict(X_train)
+            z_hat_test = ols.predict(X_test)
+
+            mse_scores.append((mse(z_train, z_hat_train), mse(z_test, z_hat_test)))
+            r2_scores.append((r2(z_train, z_hat_train), r2(z_test, z_hat_test)))
+            bvar.append(ols.beta_variance_estimate)
+        mse_s.append(np.array(mse_scores).mean(axis=0))
+        r2_s.append(np.array(r2_scores).mean(axis=0))
[…]

### 2019-09-19 assignment_one/article/main.tex
Fixed bias_variance_tradeoff.

@@
 
+\begin{figure}
+    \centering
+    \subbottom{%
+      \includegraphics[width=0.45\linewidth]{images/OLS_MSE_score_crossval.pdf}
+    }
+    \subbottom{%
+      \includegraphics[width=0.45\linewidth]{images/OLS_R2_score_crossval.pdf}
+    }
+    \caption{Ordinary least squares regression on samples from the Franke
+      function with \( N = 100 \) points in each parameter direction, yielding a total data-size of \( N^2 = 10000 \) points.
+     We have used \( K \)-fold cross validation with \( K = 10 \).
+     The mean squared error (left) and the \(R^2\)-score (right) as
+      a function of the polynomial degree \( d \). The values are displayed
+      both for the training data (solid) and the test data (dashed). The
+      noise is distributed as \( \N(0, \tfrac{5}{100})\).
+Here we clearly see the effect of the bias-variance tradeoff.
+Note as the model complexity increases (higher polynomial degree), the model at some points overfits on the training data, yielding bad generalization scores, as signified by the sharp increase in \( \mathrm{MSE}_{\mathrm{test}} \).}
+    \label{fig:ols_scores}
+  \end{figure}
+
   \subsubsection{Ordinary Least Squares}

### 2019-09-19 assignment_one/src/bias_variance_tradeoff.py

@@ -8,65 +8,72 @@ from assignment_one.src.plotting import latexify, format_axes
 from assignment_one.src.regression_class import OLS
 
 if __name__ == '__main__':
-    N = 1000
-
-    polynomial_degrees = range(0, 10)
+    N = 100
+    NUMBER_OF_FOLDS = 10
+    SIGNAL_TO_NOISE = 0.05
 
     x = np.linspace(0, 1, N)
     y = np.linspace(0, 1, N)
-
     X, Y = np.meshgrid(x, y)
+
     cartesian_product = np.dstack((X, Y)).reshape(-1, 2)
-    z_grid = franke_function(X, Y)
-    z_grid_noise = z_grid + np.random.normal(0, 0.05, size=(z_grid.shape))
-    z = z_grid.ravel()
-    z_noise = z_grid_noise.ravel()
 
-    number_of_folds = 5
-    kf = sklearn.model_selection.KFold(n_splits=number_of_folds, shuffle=True)
+    z_values = franke_function(X, Y).ravel() + SIGNAL_TO_NOISE * np.random.normal(0, 1, size=(N ** 2))
 
-    mse_s = []
-    r2_s = []
-    beta_variance = []
-    beta_std = []
-    for d in tqdm.tqdm(polynomial_degrees):
-        mse_scores = []
-        r2_scores = []
-        bvar = []
-        for train_idx, test_idx in kf.split(cartesian_product):
-            x_train, x_test = cartesian_product[train_idx], cartesian_product[test_idx]
-            z_train, z_test = z[train_idx], z[test_idx]
+    polynomial_degrees = range(0, 16)
 
-            X_train = sklearn.preprocessing.PolynomialFeatures(degree=d).fit_transform(x_train)
-            X_test = sklearn.preprocessing.PolynomialFeatures(degree=d).fit_transform(x_test)
+    kf = sklearn.model_selection.KFold(NUMBER_OF_FOLDS)
 
-            ols = OLS(X_train, z_train)
+    mse_scores_polynomial = np.zeros((2, len(polynomial_degrees)))
+    r2_scores_polynomial = np.zeros((2, len(polynomial_degrees)))
+
+    for d in polynomial_degrees:
+
+        print(f'Performing {NUMBER_OF_FOLDS}-fold cross-validation for degree {d} polynomial')
+        # create design matrix for the given degree
[…]

### 2019-09-19 assignment_one/src/bias_variance_tradeoff.py
Fixed bias_variance_tradeoff.

@@ -60,6 +60,7 @@ if __name__ == '__main__':
     latexify(fig_width=4)
     plt.plot(polynomial_degrees, mse_scores_polynomial[0], ls='-', marker='o', label=r'$\mathrm{MSE}_{\mathrm{train}}(\mathbf{y}, \hat{\mathbf{y}})$')
     plt.plot(polynomial_degrees, mse_scores_polynomial[1], ls='--', marker='*', label=r'$\mathrm{MSE}_{\mathrm{test}}(\mathbf{y}, \hat{\mathbf{y}})$')
+    plt.xlabel(r'$d$')
     plt.legend()
     plt.tight_layout()
 
@@ -71,6 +72,7 @@ if __name__ == '__main__':
     latexify(fig_width=4)
     plt.plot(polynomial_degrees, r2_scores_polynomial[0], ls='-', marker='o', label=r'$\mathrm{R}^2_{\mathrm{train}}(\mathbf{y}, \hat{\mathbf{y}})$')
     plt.plot(polynomial_degrees, r2_scores_polynomial[1], ls='--', marker='*', label=r'$\mathrm{R}^2_{\mathrm{test}}(\mathbf{y}, \hat{\mathbf{y}})$')
+    plt.xlabel(r'$d$')
     plt.legend()
     plt.tight_layout()
 

### 2019-09-24 assignment_one/article/main.tex
Working on table

@@
 \usepackage[]{amssymb}
-\usepackage[]{libertine}
-\usepackage[]{libertinust1math}
+%\usepackage[]{libertine}
+%\usepackage[]{libertinust1math}
+\usepackage[]{mathpazo}
+\usepackage[]{eulervm}
 \usepackage[T1]{fontenc}
@@
 \usepackage[]{FiraSans}
+\usepackage[]{siunitx}
 
@@
     \caption{Ordinary least squares regression on samples from the Franke
-      function with \( N = 100 \) points in each parameter direction, yielding a total data-size of \( N^2 = 10000 \) points.
-     We have used \( K \)-fold cross validation with \( K = 10 \).
-     The mean squared error (left) and the \(R^2\)-score (right) as
-      a function of the polynomial degree \( d \). The values are displayed
-      both for the training data (solid) and the test data (dashed). The
-      noise is distributed as \( \N(0, \tfrac{5}{100})\).
-Here we clearly see the effect of the bias-variance tradeoff.
-Note as the model complexity increases (higher polynomial degree), the model at some points overfits on the training data, yielding bad generalization scores, as signified by the sharp increase in \( \mathrm{MSE}_{\mathrm{test}} \).}
+	    function with \( N = 100 \) points in each parameter direction,
+	    yielding a total data-size of \( N^2 = 10000 \) points.  We have
+	    used \( K \)-fold cross validation with \( K = 10 \).  The mean
+	    squared error (left) and the \(R^2\)-score (right) as a function of
+	    the polynomial degree \( d \). The values are displayed both for
+	    the training data (solid) and the test data (dashed). The noise is
+    distributed as \( \N(0, \tfrac{5}{100})\). Here we clearly see the effect
+    of the bias-variance tradeoff. Note as the model complexity increases
+    (higher polynomial degree), the model at some points overfits on the
+    training data, yielding bad generalization scores, as signified by the
+    sharp increase in \( \mathrm{MSE}_{\mathrm{test}} \).}
     \label{fig:ols_scores}
@@
   We sample \( N = 1000 \) points in each parameter direction and compute the
-  corresponding \( N^2 \) funciton values of the Franke function both with and
+  corresponding \( N^2 \) function values of the Franke function both with and
   without added noise. The noise was selected to be distributed as \( \N(0,
@@
   \begin{table}
-    \caption{}
+		\begin{tabular}{llllllr}
+	\toprule
+	{\(d\)} &         0 &         1 &         2 &         3 &         4 &         5 \\
+	\midrule
+	$1$       & \num{1.030e-05} & \num{2.010e-05} & \num{5.315e-05} & \num{6.349e-05} & \num{7.072e-05} & \num{6.696e-05} \\[1em]
+	$x$       &       {} & \num{3.435e-05} & \num{4.673e-04} & \num{1.682e-03} & \num{4.398e-03} & \num{8.402e-03} \\
+	$y$       &       {} & \num{3.435e-05} & \num{4.673e-04} & \num{1.683e-03} & \num{4.396e-03} & \num{8.400e-03} \\[1em]
+	$x^2$     &       {} &       {} & \num{3.686e-04} & \num{6.458e-03} & \num{4.694e-02} & \num{1.959e-01} \\
+	$x y$     &       {} &       {} & \num{2.948e-04} & \num{4.106e-03} & \num{2.873e-02} & \num{1.195e-01} \\
+	$y^2$     &       {} &       {} & \num{3.686e-04} & \num{6.460e-03} & \num{4.692e-02} & \num{1.959e-01} \\[1em]
+	$x^3$     &       {} &       {} &       {} & \num{2.575e-03} & \num{8.868e-02} & \num{9.884e-01} \\
+	$x^2 y$   &       {} &       {} &       {} & \num{1.985e-03} & \num{5.199e-02} & \num{5.432e-01} \\
+	$x y^2$   &       {} &       {} &       {} & \num{1.985e-03} & \num{5.199e-02} & \num{5.429e-01} \\
+	$y^3$     &       {} &       {} &       {} & \num{2.575e-03} & \num{8.864e-02} & \num{9.884e-01} \\[1em]
+	$x^4$     &       {} &       {} &       {} &       {} & \num{2.084e-02} & \num{1.078e+00} \\
+	$x^3 y$   &       {} &       {} &       {} &       {} & \num{1.585e-02} & \num{6.287e-01} \\
+	$x^2 y^2$ &       {} &       {} &       {} &       {} & \num{1.528e-02} & \num{5.443e-01} \\
+	$x y^3$   &       {} &       {} &       {} &       {} & \num{1.585e-02} & \num{6.285e-01} \\
+	$y^4$     &       {} &       {} &       {} &       {} & \num{2.083e-02} & \num{1.078e+00} \\[1em]
+	$x^5$     &       {} &       {} &       {} &       {} &       {} & \num{1.658e-01} \\
+	$x^4 y$   &       {} &       {} &       {} &       {} &       {} & \num{1.253e-01} \\
+	$x^3 y^2$ &       {} &       {} &       {} &       {} &       {} & \num{1.193e-01} \\
+	$x^2 y^3$ &       {} &       {} &       {} &       {} &       {} & \num{1.192e-01} \\
+	$x y^4$   &       {} &       {} &       {} &       {} &       {} & \num{1.253e-01} \\
+	$y^5$     &       {} &       {} &       {} &       {} &       {} & \num{1.658e-01} \\
+	\bottomrule
+	\end{tabular}
+	\caption{The variance of the model parameters \( beta \) with respect
+		to the polynomial degree \( d \). The corresponding basis
+		function is displayed to the left. Note in particular how the
+		variance of the model parameters seems to increase with the
+		polynomial degree, which is what is to be expected according to
+	the \emph{bias-variance tradeoff}.}
     \label{tab:ols_confidence}
   \end{table}
+
+  \subsubsection{Ridge and Lasso regression}
+
+  We now incorporate Ridge and Lasso regression in our numerical experiments.
+  We are still working with the noisy sampled data.
+
   \section{Terrain Data}

### 2019-09-24 assignment_one/src/part_b.py

@@ -1,45 +1,85 @@
 import numpy as np
+import pandas as pd
 import sklearn.preprocessing
 import sklearn.model_selection
 import tqdm
+import sympy
 import matplotlib.pyplot as plt
 
-from assignment_one.src.aux import franke_function, mse, r2
+from assignment_one.src.auxiliary import franke_function, mse, r2
 from assignment_one.src.plotting import latexify, format_axes
 from assignment_one.src.regression_class import OLS
 
+
+def plot_errors():
+    latexify(3)
+    plt.semilogy(polynomial_degrees, mse_s[:, 1], ls='-', marker='o')
+    plt.xlabel('$d$')
+    plt.ylabel('$\\mathrm{MSE}(\\mathbf{y}, \\hat{\\mathbf{y}})$')
+    plt.tight_layout()
+    format_axes(ax=plt.gca())
+    plt.savefig('../article/images/OLS_MSE_score_crossval.pdf')
+    plt.clf()
+    plt.plot(polynomial_degrees, r2_s[:, 1], ls='-', marker='o')
+    plt.xlabel('$d$')
+    plt.ylabel('$\\mathrm{R}^2(\\mathbf{y}, \\hat{\\mathbf{y}})$')
+    plt.tight_layout()
+    format_axes(ax=plt.gca())
+    plt.savefig('../article/images/OLS_R2_score_crossval.pdf')
+
+
+def format_feature_names(feature_names):
+    feature_names = list(map(lambda x: x.replace('x0', 'x'), feature_names))
+    feature_names = list(map(lambda x: x.replace('x1', 'y'), feature_names))
+    feature_names = list(map(lambda x: x.replace('x^0y^0', '1'), feature_names))
+    feature_names = list(map(lambda x: r'$' + x + r'$', feature_names))
+
+    return feature_names
+
+
+def create_variance_table(beta_variance, feature_names, polynomial_degrees):
+    formatted_features = format_feature_names(feature_names)
+
+    df = pd.DataFrame(columns=polynomial_degrees, index=formatted_features)
+    for i, d in enumerate(polynomial_degrees):
+        df[d][:len(beta_variance[i])] = beta_variance[i]
+
+    return df
+
+
 if __name__ == '__main__':
-    N = 1000
+    N = 100
 
-    polynomial_degrees = range(0, 10)
+    polynomial_degrees = range(0, 5)
 
-    x = np.linspace(0, 1, N)
-    y = np.linspace(0, 1, N)
-
[…]

### 2019-09-24 assignment_one/article/main.tex
Working on lasso and ridge

--- 2019-09-19 assignment_one/article/main.tex
@@
   \begin{table}
-    \caption{}
+		\begin{tabular}{llllllr}
+	\toprule
+	{\(d\)} &    2 &         3 &         4 &         5 \\
+	\midrule
+	$1$       & \num{5.315e-05} & \num{6.349e-05} & \num{7.072e-05} & \num{6.696e-05} \\[1em]
+	$x$       & \num{4.673e-04} & \num{1.682e-03} & \num{4.398e-03} & \num{8.402e-03} \\
+	$y$       & \num{4.673e-04} & \num{1.683e-03} & \num{4.396e-03} & \num{8.400e-03} \\[1em]
+	$x^2$     & \num{3.686e-04} & \num{6.458e-03} & \num{4.694e-02} & \num{1.959e-01} \\
+	$x y$     & \num{2.948e-04} & \num{4.106e-03} & \num{2.873e-02} & \num{1.195e-01} \\
+	$y^2$     & \num{3.686e-04} & \num{6.460e-03} & \num{4.692e-02} & \num{1.959e-01} \\[1em]
+	$x^3$     &       {} & \num{2.575e-03} & \num{8.868e-02} & \num{9.884e-01} \\
+	$x^2 y$   &       {} & \num{1.985e-03} & \num{5.199e-02} & \num{5.432e-01} \\
+	$x y^2$   &       {} & \num{1.985e-03} & \num{5.199e-02} & \num{5.429e-01} \\
+	$y^3$     &       {} & \num{2.575e-03} & \num{8.864e-02} & \num{9.884e-01} \\[1em]
+	$x^4$     &       {} &       {} & \num{2.084e-02} & \num{1.078e+00} \\
+	$x^3 y$   &       {} &       {} & \num{1.585e-02} & \num{6.287e-01} \\
+	$x^2 y^2$ &       {} &       {} & \num{1.528e-02} & \num{5.443e-01} \\
+	$x y^3$   &       {} &       {} & \num{1.585e-02} & \num{6.285e-01} \\
+	$y^4$     &       {} &       {} & \num{2.083e-02} & \num{1.078e+00} \\[1em]
+	$x^5$     &       {} &       {} &       {} & \num{1.658e-01} \\
+	$x^4 y$   &       {} &       {} &       {} & \num{1.253e-01} \\
+	$x^3 y^2$ &       {} &       {} &       {} & \num{1.193e-01} \\
+	$x^2 y^3$ &       {} &       {} &       {} & \num{1.192e-01} \\
+	$x y^4$   &       {} &       {} &       {} & \num{1.253e-01} \\
+	$y^5$     &       {} &       {} &       {} & \num{1.658e-01} \\
+	\bottomrule
+	\end{tabular}
+	\caption{The variance of the model parameters \( beta \) with respect
+		to the polynomial degree \( d \). The corresponding basis
+		function is displayed to the left. Note in particular how the
+		variance of the model parameters seems to increase with the
+		polynomial degree, which is what is to be expected according to
+	the \emph{bias-variance tradeoff}.}
     \label{tab:ols_confidence}
   \end{table}
+
+  \subsubsection{Ridge and Lasso regression}
+
+  We now incorporate Ridge and Lasso regression in our numerical experiments.
+  We are still working with the noisy sampled data.
+
   \section{Terrain Data}
@@
   \emph{truncated hierarchical B-splines} (THB) and \emph{locally refined
-  B-splines} (LRB). These methods allow us to place more degrees of freedom in regions of high
-  data-variance, thus locally increasing the model complexity.
+  B-splines} (LRB). These methods allow us to place more degrees of freedom in
+  regions of high data-variance, thus locally increasing the model complexity.
+
+  A study on least-squares surfaces using LR B-splines for representing terrain
+  data was presented in \cite{skyttLocallyRefinedSpline2015}. Here, regions of
+  high discrepancy between the true surface and the approximating surface was
+  locally refined until an acceptable tolerance was reached.
 

### 2019-09-24 assignment_one/src/auxiliary.py

@@ -1,4 +1,5 @@
 import numpy as np
+import pandas as pd
 
 
 def franke_function(x, y):
@@ -46,3 +47,22 @@ def r2(y, y_hat):
     den = ((y - y_mean) ** 2).sum()
 
     return 1 - num / den
+
+
+def format_feature_names(feature_names):
+    feature_names = list(map(lambda x: x.replace('x0', 'x'), feature_names))
+    feature_names = list(map(lambda x: x.replace('x1', 'y'), feature_names))
+    feature_names = list(map(lambda x: x.replace('x^0y^0', '1'), feature_names))
+    feature_names = list(map(lambda x: r'$' + x + r'$', feature_names))
+
+    return feature_names
+
+
+def create_variance_table(beta_variance, feature_names, polynomial_degrees):
+    formatted_features = format_feature_names(feature_names)
+
+    df = pd.DataFrame(columns=polynomial_degrees, index=formatted_features)
+    for i, d in enumerate(polynomial_degrees):
+        df[d][:len(beta_variance[i])] = beta_variance[i]
+
+    return df

### 2019-09-24 assignment_one/src/part_b.py

@@ -1,12 +1,9 @@
+import matplotlib.pyplot as plt
 import numpy as np
-import pandas as pd
-import sklearn.preprocessing
 import sklearn.model_selection
 import tqdm
-import sympy
-import matplotlib.pyplot as plt
 
-from assignment_one.src.auxiliary import franke_function, mse, r2
+from assignment_one.src.auxiliary import franke_function, mse, r2, create_variance_table
 from assignment_one.src.plotting import latexify, format_axes
 from assignment_one.src.regression_class import OLS
 
@@ -28,82 +25,76 @@ def plot_errors():
     plt.savefig('../article/images/OLS_R2_score_crossval.pdf')
 
 
-def format_feature_names(feature_names):
-    feature_names = list(map(lambda x: x.replace('x0', 'x'), feature_names))
-    feature_names = list(map(lambda x: x.replace('x1', 'y'), feature_names))
-    feature_names = list(map(lambda x: x.replace('x^0y^0', '1'), feature_names))
-    feature_names = list(map(lambda x: r'$' + x + r'$', feature_names))
-
-    return feature_names
-
-
-def create_variance_table(beta_variance, feature_names, polynomial_degrees):
-    formatted_features = format_feature_names(feature_names)
-
-    df = pd.DataFrame(columns=polynomial_degrees, index=formatted_features)
-    for i, d in enumerate(polynomial_degrees):
-        df[d][:len(beta_variance[i])] = beta_variance[i]
-
-    return df
-
-
-if __name__ == '__main__':
-    N = 100
-
-    polynomial_degrees = range(0, 5)
-
-    cartesian_product = np.random.random(size=(N, 2))
-    X, Y = np.meshgrid(cartesian_product[:, 0], cartesian_product[:, 1])
-    # cartesian_product = np.dstack((X, Y)).reshape(-1, 2)
-    z_grid = franke_function(X, Y)
-    z_grid_noise = z_grid + np.random.normal(0, 0.05, size=(z_grid.shape))
-    z = z_grid.ravel()
-    z_noise = z_grid_noise.ravel()
-
-    number_of_folds = 100
-    kf = sklearn.model_selection.KFold(n_splits=number_of_folds, shuffle=True)
-
+def perform_cross_validation(cartesian_product, z_values, number_of_folds=10, Regressor=OLS, lambd=None):
     mse_s = []
     r2_s = []
     beta_variance = []
     beta_std = []
+    beta = []
+
[…]

### 2019-09-24 assignment_one/src/auxiliary.py
Added lambda param to CV-method

@@ -1,5 +1,11 @@
 import numpy as np
 import pandas as pd
+import sklearn.model_selection
+import sklearn.preprocessing
+import tqdm
+
+from assignment_one.src.part_b import polynomial_degrees
+from assignment_one.src.regression_class import OLS
 
 
 def franke_function(x, y):
@@ -66,3 +72,54 @@ def create_variance_table(beta_variance, feature_names, polynomial_degrees):
         df[d][:len(beta_variance[i])] = beta_variance[i]
 
     return df
+
+
+def perform_cross_validation(cartesian_product, z_values, number_of_folds=10, Regressor=OLS, lambd=None):
+    mse_s = []
+    r2_s = []
+    beta_variance = []
+    beta_std = []
+    beta = []
+
+    kf = sklearn.model_selection.KFold(n_splits=number_of_folds, shuffle=True)
+
+    for d in tqdm.tqdm(polynomial_degrees):
+        mse_scores = []
+        r2_scores = []
+        bvar = []
+        std = []
+        beta_vals = []
+        for train_idx, test_idx in kf.split(cartesian_product):
+            x_train, x_test = cartesian_product[train_idx], cartesian_product[test_idx]
+            z_train, z_test = z_values[train_idx], z_values[test_idx]
+
+            poly_fit = sklearn.preprocessing.PolynomialFeatures(degree=d)
+            X_train = poly_fit.fit_transform(x_train)
+            X_test = poly_fit.fit_transform(x_test)
+
+            if lambd is None:
+                regressor = Regressor(X_train, z_train)
+            else:
+                regressor = Regressor(X_train, z_train, lambd=lambd)
+
+            z_hat_train = regressor.predict(X_train)
+            z_hat_test = regressor.predict(X_test)
+
+            mse_scores.append((mse(z_train, z_hat_train), mse(z_test, z_hat_test)))
+            r2_scores.append((r2(z_train, z_hat_train), r2(z_test, z_hat_test)))
+            bvar.append(regressor.beta_variance_estimate)
+            std.append(regressor.beta_std_dev_estimate)
+            beta_vals.append(regressor.beta)
+
+        mse_s.append(np.array(mse_scores).mean(axis=0))
+        r2_s.append(np.array(r2_scores).mean(axis=0))
+        beta_variance.append(np.array(bvar).mean(axis=0))
+        beta_std.append(np.array(std).mean(axis=0))
[…]

### 2019-09-24 assignment_one/src/part_b.py

@@ -1,9 +1,7 @@
 import matplotlib.pyplot as plt
 import numpy as np
-import sklearn.model_selection
-import tqdm
 
-from assignment_one.src.auxiliary import franke_function, mse, r2, create_variance_table
+from assignment_one.src.auxiliary import franke_function, create_variance_table, perform_cross_validation
 from assignment_one.src.plotting import latexify, format_axes
 from assignment_one.src.regression_class import OLS
 
@@ -23,54 +21,6 @@ def plot_errors():
     plt.tight_layout()
     format_axes(ax=plt.gca())
     plt.savefig('../article/images/OLS_R2_score_crossval.pdf')
-
-
-def perform_cross_validation(cartesian_product, z_values, number_of_folds=10, Regressor=OLS, lambd=None):
-    mse_s = []
-    r2_s = []
-    beta_variance = []
-    beta_std = []
-    beta = []
-
-    kf = sklearn.model_selection.KFold(n_splits=number_of_folds, shuffle=True)
-
-    for d in tqdm.tqdm(polynomial_degrees):
-        mse_scores = []
-        r2_scores = []
-        bvar = []
-        std = []
-        beta_vals = []
-        for train_idx, test_idx in kf.split(cartesian_product):
-            x_train, x_test = cartesian_product[train_idx], cartesian_product[test_idx]
-            z_train, z_test = z_values[train_idx], z_values[test_idx]
-
-            poly_fit = sklearn.preprocessing.PolynomialFeatures(degree=d)
-            X_train = poly_fit.fit_transform(x_train)
-            X_test = poly_fit.fit_transform(x_test)
-
-            regressor = Regressor(X_train, z_train)
-
-            z_hat_train = regressor.predict(X_train)
-            z_hat_test = regressor.predict(X_test)
-
-            mse_scores.append((mse(z_train, z_hat_train), mse(z_test, z_hat_test)))
-            r2_scores.append((r2(z_train, z_hat_train), r2(z_test, z_hat_test)))
-            bvar.append(regressor.beta_variance_estimate)
-            std.append(regressor.beta_std_dev_estimate)
-            beta_vals.append(regressor.beta)
-
-        mse_s.append(np.array(mse_scores).mean(axis=0))
-        r2_s.append(np.array(r2_scores).mean(axis=0))
-        beta_variance.append(np.array(bvar).mean(axis=0))
[…]

### 2019-09-24 assignment_one/src/auxiliary.py
Added tests against SKlearn.

@@ -4,7 +4,6 @@ import sklearn.model_selection
 import sklearn.preprocessing
 import tqdm
 
-from assignment_one.src.part_b import polynomial_degrees
 from assignment_one.src.regression_class import OLS
 
 
@@ -74,7 +73,8 @@ def create_variance_table(beta_variance, feature_names, polynomial_degrees):
     return df
 
 
-def perform_cross_validation(cartesian_product, z_values, number_of_folds=10, Regressor=OLS, lambd=None):
+def perform_cross_validation(cartesian_product, z_values, polynomial_degrees, number_of_folds=10, Regressor=OLS,
+                             lambd=None):
     mse_s = []
     r2_s = []
     beta_variance = []
@@ -98,9 +98,9 @@ def perform_cross_validation(cartesian_product, z_values, number_of_folds=10, Re
             X_test = poly_fit.fit_transform(x_test)
 
             if lambd is None:
-                regressor = Regressor(X_train, z_train)
+                regressor = Regressor(X_train, z_train, inversion_method='svd')
             else:
-                regressor = Regressor(X_train, z_train, lambd=lambd)
+                regressor = Regressor(X_train, z_train, lambd=lambd, inversion_method='svd')
 
             z_hat_train = regressor.predict(X_train)
             z_hat_test = regressor.predict(X_test)

### 2019-09-24 assignment_one/src/part_b.py

@@ -1,7 +1,7 @@
 import matplotlib.pyplot as plt
 import numpy as np
 
-from assignment_one.src.auxiliary import franke_function, create_variance_table, perform_cross_validation
+from assignment_one.src.auxiliary import franke_function, perform_cross_validation, create_variance_table
 from assignment_one.src.plotting import latexify, format_axes
 from assignment_one.src.regression_class import OLS
 
@@ -37,6 +37,7 @@ if __name__ == '__main__':
     number_of_folds = 10
 
     mse_s, r2_s, beta, beta_variance, beta_std, feature_names = perform_cross_validation(cartesian_product, z_noise,
+                                                                                         polynomial_degrees=polynomial_degrees,
                                                                                          number_of_folds=number_of_folds,
                                                                                          Regressor=OLS)
     # plot_errors()

### 2019-09-24 assignment_one/src/part_c.py

@@ -1,0 +1,64 @@
+import numpy as np
+import sklearn
+
+import matplotlib.pyplot as plt
+from assignment_one.src.auxiliary import franke_function, perform_cross_validation, create_variance_table, mse, r2
+from assignment_one.src.regression_class import Ridge, Lasso, OLS
+
+if __name__ == '__main__':
+    N = 10000
+    d = 5
+
+    np.random.seed(42)
+
+    lambda_vals = np.logspace(-5, 0, 50)
+
+    cartesian_product = np.random.random(size=(N, 2))
+    X, Y = np.meshgrid(cartesian_product[:, 0], cartesian_product[:, 1])
+    z_grid = franke_function(X, Y)
+    z_grid_noise = z_grid + np.random.normal(0, 0.5, size=(z_grid.shape))
+    z = z_grid.ravel()
+    z_noise = z_grid_noise.ravel()
+
+    number_of_folds = 10
+
+    results = {}
+    for i, (name, regressor) in enumerate(zip(['Ridge', 'Lasso'], [Ridge, Lasso])):
+
+        mse_results = []
+        r2_results = []
+        for l in lambda_vals:
+            kf = sklearn.model_selection.KFold(n_splits=number_of_folds, shuffle=True)
+
+            mse_point_vals = []
+            r2_point_vals = []
+
+            for train_idx, test_idx in kf.split(cartesian_product):
+                x_train, x_test = cartesian_product[train_idx], cartesian_product[test_idx]
+                z_train, z_test = z_noise[train_idx], z_noise[test_idx]
+
+                poly_fit = sklearn.preprocessing.PolynomialFeatures(degree=d)
+                X_train = poly_fit.fit_transform(x_train)
+                X_test = poly_fit.fit_transform(x_test)
+
+                R = regressor(X_train, z_train, lmbd=l, inversion_method='svd')
+
+                z_hat_train = R.predict(X_train)
+                z_hat_test = R.predict(X_test)
+
+                mse_point_vals.append((mse(z_train, z_hat_train), mse(z_test, z_hat_test)))
+                r2_point_vals.append((r2(z_train, z_hat_train), r2(z_test, z_hat_test)))
+
+            mse_results.append(np.array(mse_point_vals).mean(axis=0))
+            r2_results.append(np.array(r2_point_vals).mean(axis=0))
+
+        mse_results = np.array(mse_results)
+        r2_results = np.array(r2_results)
+
[…]

### 2019-09-24 assignment_one/src/regression_class.py

@@ -137,9 +137,6 @@ class OLS(Regression):
 
 class Ridge(Regression):
 
-    def fit(self):
-        pass
-
     def __init__(self, X, y, inversion_method='direct', lmbd=0.1):
         super().__init__(X, y, inversion_method)
         self.lmbd = lmbd

### 2019-09-24 assignment_one/src/tests/test_regressors.py

@@ -1,0 +1,42 @@
+import numpy as np
+import pytest
+import sklearn
+
+from assignment_one.src.regression_class import OLS, Lasso, Ridge
+
+
+@pytest.mark.parametrize("d", range(10))
+@pytest.mark.parametrize("regressor, sklearn_reg, lmd",
+                         [(OLS, sklearn.linear_model.LinearRegression, None), (Ridge, sklearn.linear_model.Ridge, 0.1),
+                          (Lasso, sklearn.linear_model.Lasso, 0.1)])
+def test_against_sklearn(d, regressor, sklearn_reg, lmd):
+    X, y = get_data(d)
+
+    if lmd:
+        linreg = sklearn_reg(fit_intercept=False, alpha=lmd).fit(X, y)
+        O = regressor(X, y, inversion_method='svd', lmbd=lmd)
+    else:
+        linreg = sklearn_reg(fit_intercept=False).fit(X, y)
+        O = regressor(X, y, inversion_method='svd')
+
+    beta = O.beta
+    y_hat = O.predict(X)
+
+    beta_sklearn = linreg.coef_
+    y_hat_sklearn = linreg.predict(X)
+
+    np.testing.assert_allclose(beta, beta_sklearn, rtol=1e-4)
+    np.testing.assert_allclose(y_hat, y_hat_sklearn, rtol=1e-4)
+
+
+def get_data(d):
+    np.random.seed(42)
+    X = np.array([
+        [1, 1],
+        [1, 2],
+        [2, 2],
+        [2, 3]
+    ])
+    X = sklearn.preprocessing.PolynomialFeatures(degree=d).fit_transform(X)
+    y = np.dot(X, np.random.random(X.shape[1]))
+    return X, y

### 2019-09-24 assignment_one/src/bias_variance_tradeoff.py
Added tests against sklearn. As the polynomial degree increases, my method suffers from numerical instability it seems.

@@ -27,6 +27,9 @@ if __name__ == '__main__':
     mse_scores_polynomial = np.zeros((2, len(polynomial_degrees)))
     r2_scores_polynomial = np.zeros((2, len(polynomial_degrees)))
 
+    bias = []
+    vars = []
+
     for d in polynomial_degrees:
 
         print(f'Performing {NUMBER_OF_FOLDS}-fold cross-validation for degree {d} polynomial')
@@ -39,6 +42,7 @@ if __name__ == '__main__':
         r2_values = np.zeros((2, NUMBER_OF_FOLDS))
 
         # split into test and train
+        test_predictions = []
         for i, (train_idx, test_idx) in enumerate(kf.split(design_matrix)):
             X_train, X_test = design_matrix[train_idx], design_matrix[test_idx]
             z_train, z_test = z_values[train_idx], z_values[test_idx]
@@ -51,15 +55,21 @@ if __name__ == '__main__':
             mse_values[:, i] = [mse(z_train, z_hat_train), mse(z_test, z_hat_test)]
             r2_values[:, i] = [r2(z_train, z_hat_train), r2(z_test, z_hat_test)]
 
+            test_predictions.append(z_hat_test)
+
+        test_predictions = np.array(test_predictions)
+
         # compute the mean values over all folds
         mse_scores_polynomial[:, d] = mse_values.mean(axis=1)
         r2_scores_polynomial[:, d] = r2_values.mean(axis=1)
 
-
+        bias.append(np.mean(np.mean(test_predictions)))
     # plot estimated errors
     latexify(fig_width=4)
-    plt.plot(polynomial_degrees, mse_scores_polynomial[0], ls='-', marker='o', label=r'$\mathrm{MSE}_{\mathrm{train}}(\mathbf{y}, \hat{\mathbf{y}})$')
-    plt.plot(polynomial_degrees, mse_scores_polynomial[1], ls='--', marker='*', label=r'$\mathrm{MSE}_{\mathrm{test}}(\mathbf{y}, \hat{\mathbf{y}})$')
+    plt.plot(polynomial_degrees, mse_scores_polynomial[0], ls='-', marker='o',
+             label=r'$\mathrm{MSE}_{\mathrm{train}}(\mathbf{y}, \hat{\mathbf{y}})$')
[…]

### 2019-09-24 assignment_one/src/regression_class.py

@@ -172,12 +172,3 @@ if __name__ == '__main__':
     O = OLS(X, y, inversion_method='svd')
     R = Ridge(X, y, lmbd=0.1)
     L = Lasso(X, y, lmbd=0.1)
-    print(O.beta)
-    print(R.beta)
-    print(L.beta)
-
-    print(O.predict())
-    for e in O, R, L:
-        print(e.y_variance_estimate)
-        print(e.beta_variance_estimate)
-        print(e.r2(), e.mse())

### 2019-09-24 assignment_one/src/tests/test_regressors.py

@@ -2,6 +2,7 @@ import numpy as np
 import pytest
 import sklearn
 
+from assignment_one.src.auxiliary import r2
 from assignment_one.src.regression_class import OLS, Lasso, Ridge
 
 
@@ -27,6 +28,10 @@ def test_against_sklearn(d, regressor, sklearn_reg, lmd):
 
     np.testing.assert_allclose(beta, beta_sklearn, rtol=1e-4)
     np.testing.assert_allclose(y_hat, y_hat_sklearn, rtol=1e-4)
+
+    sk_error = sklearn.metrics.mean_squared_error(y, y_hat_sklearn)
+    error = O.mse()
+    assert abs(sk_error - error) < 1.0e-6
 
 
 def get_data(d):

### 2019-09-24 assignment_one/article/main.tex
Working on article.

@@
   \subsection{The bias-variance tradeoff} 
-
+  \label{sub:bias_variance} 
   Since we are attempting to estimate a true distribution based on finite
@@
   \begin{equation}
+    \label{eq:bias_variance}
     \expect_{\data, \varepsilon}[\cost(\data, \beta)] = \frac{1}{n} \textcolor{Blue}{\sum_{i=1}^n \left(f(\x_i) - \expect_{\data}\left[\hat{y}_i\right]\right)^2}  + \frac{1}{n}\textcolor{Maroon}{\sum_{i=1}^n \expect_{\data}\big[\left(\hat{y}_i - \expect_{\data} [\hat{y}_i]\right)^2\big]} + \textcolor{Green} \sigma^2.
@@
   parameters \( \beta_{\OLS} \) of the ordinary least squares regression cannot
-  be determined. The propsed fix to this problem is to slightly perturb the
-  matrix \( \X^T \X \) by adding small positive values along the diagonal to
-  ensure its invertability. This is known as \emph{ridge regression} and yields
-  a model parameter estimator
+  be determined. The propsed fix to this problem
+  \cite{tikhonovNumericalMethodsSolution1995,hoerlRidgeRegressionBiased} is to
+  slightly perturb the matrix \( \X^T \X \) by adding small positive values
+  along the diagonal to ensure its invertability. This is known as \emph{ridge
+  regression} and yields a model parameter estimator
   \begin{equation}
@@
   will tend to \( 95\% \)''.
+
+  We are also interested in computing the model bias and variance as discussed
+  in \cref{sub:bias_variance}. Numerically, we estimate these quantities by
+  replacing the expectation values in \cref{eq:bias_variance} by averages.
+
+  \subsection{Cross-validation}
+
+  In order to obtain more reasonable estimates for these statistical
+  quantities, we perform so-called \emph{resampling} of the data. There are
+  several approaches to resampling, with two of the more common ones being
+  \emph{bootstrap} and \emph{k-fold cross-validation}. Without going into
+  details, the bootstrap method generally performs better with a significant
+  data size, with cross-validation also performing well on smaller data sets.
+  With this in mind, we will discuss cross-validation here.
 

### 2019-09-24 assignment_one/src/part_a.py
Redoing part_a.

@@ -4,52 +4,52 @@ import matplotlib.pyplot as plt
 import tqdm
 from matplotlib import cm
 from mpl_toolkits.mplot3d import Axes3D
-from assignment_one.src.aux import franke_function
+from sklearn.model_selection import KFold
+
+from assignment_one.src.auxiliary import franke_function
 from assignment_one.src.plotting import latexify, format_axes
 from assignment_one.src.regression_class import OLS
 
+
+def perform_k_fold(X, y, number_of_folds=1, regressor=OLS, seed=42):
+    """
+    Perform a single cross-validation resampling for the given regressor. All relevant
+    results are returned in the dictionary results.
+
+    :param X:
+    :param number_of_folds:
+    :param regressor:
+    :param seed:
+    :return:
+    """
+    results = {}
+    kf = KFold(n_splits=number_of_folds, shuffle=True, random_state=seed)
+
+    for i, (train_idx, test_idx) in enumerate(kf.split(X, y)):
+        print(i, train_idx, test_idx)
+        return results
+
+
+def perform_ols_estimates_with_and_without_noise(seed=42, number_of_folds=1):
+    np.random.seed(seed)
+    polynomial_degrees = range(2, 11)
+
+    N = 10
+    signal_to_noise = 0.01
+    x = np.random.random((N, 2))
+
+    z = np.array([franke_function(X, Y) for X, Y in x])
+    z_noise = z + signal_to_noise * np.random.normal(0, 1, z.shape)
+
+    regressor = OLS
+    for name, data in zip(['true', 'noise'], [z, z_noise]):
+        for d in polynomial_degrees:
+            poly_fit = sklearn.preprocessing.PolynomialFeatures(degree=d)
+            X = poly_fit.fit_transform(x)
+            results = perform_k_fold(X, data, number_of_folds, regressor, seed)
+
+    return
+
+
 if __name__ == '__main__':
-    N = 1000
-    polynomial_degrees = range(0, 11)
-
-    x = np.linspace(0, 1, N)
-    y = np.linspace(0, 1, N)
-
-    X, Y = np.meshgrid(x, y)
-    cartesian_product = np.dstack((X, Y)).reshape(-1, 2)
-    z_grid = franke_function(X, Y)
[…]

### 2019-09-26 assignment_one/src/tests/test_regressors.py
Due to numerical instabilities by inverting matrices, the results of my computations does not directly correspond to the ones given by sklearn. However, this might be solved by using SVD-decomposition, and use QR-decomposition and np.solve instead of inverting.

@@ -6,19 +6,20 @@ from assignment_one.src.auxiliary import r2
 from assignment_one.src.regression_class import OLS, Lasso, Ridge
 
 
+@pytest.mark.parametrize('inversion_method', ['svd'])
 @pytest.mark.parametrize("d", range(10))
 @pytest.mark.parametrize("regressor, sklearn_reg, lmd",
                          [(OLS, sklearn.linear_model.LinearRegression, None), (Ridge, sklearn.linear_model.Ridge, 0.1),
                           (Lasso, sklearn.linear_model.Lasso, 0.1)])
-def test_against_sklearn(d, regressor, sklearn_reg, lmd):
+def test_against_sklearn(d, regressor, sklearn_reg, lmd, inversion_method):
     X, y = get_data(d)
 
     if lmd:
         linreg = sklearn_reg(fit_intercept=False, alpha=lmd).fit(X, y)
-        O = regressor(X, y, inversion_method='svd', lmbd=lmd)
+        O = regressor(X, y, inversion_method=inversion_method, lmbd=lmd)
     else:
         linreg = sklearn_reg(fit_intercept=False).fit(X, y)
-        O = regressor(X, y, inversion_method='svd')
+        O = regressor(X, y, inversion_method=inversion_method)
 
     beta = O.beta
     y_hat = O.predict(X)
@@ -31,7 +32,7 @@ def test_against_sklearn(d, regressor, sklearn_reg, lmd):
 
     sk_error = sklearn.metrics.mean_squared_error(y, y_hat_sklearn)
     error = O.mse()
-    assert abs(sk_error - error) < 1.0e-6
+    assert abs(sk_error - error) < 1.0e-4
 
 
 def get_data(d):

### 2019-09-26 assignment_one/src/part_a.py
Removed the seed from perform_k_fold.

@@ -11,7 +11,7 @@ from assignment_one.src.plotting import latexify, format_axes
 from assignment_one.src.regression_class import OLS
 
 
-def perform_k_fold(X, y, number_of_folds=1, regressor=OLS, seed=42):
+def perform_k_fold(X, y, number_of_folds=1, regressor=OLS):
     """
     Perform a single cross-validation resampling for the given regressor. All relevant
     results are returned in the dictionary results.
@@ -23,7 +23,7 @@ def perform_k_fold(X, y, number_of_folds=1, regressor=OLS, seed=42):
     :return:
     """
     results = {}
-    kf = KFold(n_splits=number_of_folds, shuffle=True, random_state=seed)
+    kf = KFold(n_splits=number_of_folds, shuffle=True)
 
     for i, (train_idx, test_idx) in enumerate(kf.split(X, y)):
         print(i, train_idx, test_idx)
@@ -34,7 +34,7 @@ def perform_ols_estimates_with_and_without_noise(seed=42, number_of_folds=1):
     np.random.seed(seed)
     polynomial_degrees = range(2, 11)
 
-    N = 10
+    N = 100
     signal_to_noise = 0.01
     x = np.random.random((N, 2))
 
@@ -46,7 +46,7 @@ def perform_ols_estimates_with_and_without_noise(seed=42, number_of_folds=1):
         for d in polynomial_degrees:
             poly_fit = sklearn.preprocessing.PolynomialFeatures(degree=d)
             X = poly_fit.fit_transform(x)
-            results = perform_k_fold(X, data, number_of_folds, regressor, seed)
+            results = perform_k_fold(X, data, number_of_folds, regressor)
 
     return
 

### 2019-09-26 assignment_one/src/part_a.py
KFold seems to be working.

@@ -6,15 +6,15 @@ from matplotlib import cm
 from mpl_toolkits.mplot3d import Axes3D
 from sklearn.model_selection import KFold
 
-from assignment_one.src.auxiliary import franke_function
+from assignment_one.src.auxiliary import franke_function, mse, r2
 from assignment_one.src.plotting import latexify, format_axes
-from assignment_one.src.regression_class import OLS
+from assignment_one.src.regression_class import OLS, Ridge
 
 
-def perform_k_fold(X, y, number_of_folds=1, regressor=OLS):
+def perform_k_fold(X, y, number_of_folds=1, regressor=OLS, regressor_parameters={}):
     """
     Perform a single cross-validation resampling for the given regressor. All relevant
-    results are returned in the dictionary results.
+    intermediate_results are returned in the dictionary intermediate_results.
 
     :param X:
     :param number_of_folds:
@@ -22,17 +22,34 @@ def perform_k_fold(X, y, number_of_folds=1, regressor=OLS):
     :param seed:
     :return:
     """
-    results = {}
+    intermediate_results = {'mse': [], 'r2': [], 'model_bias': [], 'model_variance': [], 'beta_variance': []}
+
     kf = KFold(n_splits=number_of_folds, shuffle=True)
 
     for i, (train_idx, test_idx) in enumerate(kf.split(X, y)):
-        print(i, train_idx, test_idx)
-        return results
+        X_test, X_train = X[test_idx], X[train_idx]
+        y_test, y_train = y[test_idx], y[train_idx]
+
+        regr = regressor(X_train, y_train, **regressor_parameters)
+
+        y_hat_train = regr.predict(X_train)
+        y_hat_test = regr.predict(X_test)
+        test_pred_mean = y_hat_test.mean()
+
+        intermediate_results['mse'].append(mse(y_test, y_hat_test))
+        intermediate_results['r2'].append(r2(y_test, y_hat_test))
+        intermediate_results['model_variance'].append(np.mean((y_hat_test - test_pred_mean)) ** 2)
+        intermediate_results['model_bias'].append(np.mean((y_test - test_pred_mean) ** 2))
+        intermediate_results['beta_variance'].append(regr.beta_variance_estimate)
+
+    intermediate_results.update({k: np.array(v).mean(axis=0) for (k, v) in intermediate_results.items()})
+
+    return intermediate_results
 
 
[…]

### 2019-09-26 assignment_one/src/regression_class.py

@@ -9,6 +9,7 @@ class Regression(object):
         self.y = y
         self.xtx = np.dot(X.T, X)
         self.inv = np.linalg.inv if inversion_method == 'direct' else Regression.svd_inv
+        self.name = 'RegressionBaseClass'
 
     def fit(self):
         raise NotImplementedError()
@@ -140,6 +141,7 @@ class Ridge(Regression):
     def __init__(self, X, y, inversion_method='direct', lmbd=0.1):
         super().__init__(X, y, inversion_method)
         self.lmbd = lmbd
+        self.name = 'Ridge'
 
     @property
     def beta(self):
@@ -157,6 +159,7 @@ class Lasso(Regression):
         self.lmbd = lmbd
         self._regressor = sklearn.linear_model.Lasso(alpha=self.lmbd, fit_intercept=False)
         self._regressor.fit(self.X, self.y)
+        self.name = 'Lasso'
 
     @property
     def beta(self):

### 2019-09-26 assignment_one/src/part_a.py
KFold aggregates should be working now.

@@ -13,16 +13,16 @@ from assignment_one.src.regression_class import OLS, Ridge
 
 def perform_k_fold(X, y, number_of_folds=1, regressor=OLS, regressor_parameters={}):
     """
-    Perform a single cross-validation resampling for the given regressor. All relevant
-    intermediate_results are returned in the dictionary intermediate_results.
+    Perform a single cross-validation re-sampling for the given regressor. All relevant
+    intermediate_results are returned in the dictionary intermediate_results, and aggregated as mean-values.
 
     :param X:
     :param number_of_folds:
     :param regressor:
-    :param seed:
     :return:
     """
-    intermediate_results = {'mse': [], 'r2': [], 'model_bias': [], 'model_variance': [], 'beta_variance': []}
+    intermediate_results = {'regressor': regressor.name, 'mse': [], 'r2': [], 'model_bias': [], 'model_variance': [],
+                            'beta_variance': [], 'regressor_params': regressor_parameters}
 
     kf = KFold(n_splits=number_of_folds, shuffle=True)
 
@@ -32,7 +32,7 @@ def perform_k_fold(X, y, number_of_folds=1, regressor=OLS, regressor_parameters=
 
         regr = regressor(X_train, y_train, **regressor_parameters)
 
-        y_hat_train = regr.predict(X_train)
+        y_hat_train = regr.predict(X_train)  # this predict, while not used - performs the fit behind the scenes.
         y_hat_test = regr.predict(X_test)
         test_pred_mean = y_hat_test.mean()
 
@@ -42,7 +42,9 @@ def perform_k_fold(X, y, number_of_folds=1, regressor=OLS, regressor_parameters=
         intermediate_results['model_bias'].append(np.mean((y_test - test_pred_mean) ** 2))
         intermediate_results['beta_variance'].append(regr.beta_variance_estimate)
 
-    intermediate_results.update({k: np.array(v).mean(axis=0) for (k, v) in intermediate_results.items()})
+    intermediate_results.update(
+        {k: np.array(v).mean(axis=0) for (k, v) in intermediate_results.items() if
+         k not in ['regressor', 'regressor_params']})
 
     return intermediate_results
 
@@ -60,13 +62,16 @@ def perform_ols_estimates_with_and_without_noise(seed=42, number_of_folds=1):
 
     regressor = Ridge
     regr_params = {'lmbd': 0.001}
+    results = {}
     for name, data in zip(['true', 'noise'], [z, z_noise]):
+
[…]

### 2019-09-26 assignment_one/src/regression_class.py

@@ -3,13 +3,13 @@ import sklearn.linear_model
 
 
 class Regression(object):
+    name = 'RegressionBaseClass'
 
     def __init__(self, X, y, inversion_method='direct'):
         self.X = X
         self.y = y
         self.xtx = np.dot(X.T, X)
         self.inv = np.linalg.inv if inversion_method == 'direct' else Regression.svd_inv
-        self.name = 'RegressionBaseClass'
 
     def fit(self):
         raise NotImplementedError()
@@ -123,6 +123,7 @@ class Regression(object):
 
 
 class OLS(Regression):
+    name = 'OLS'
     @property
     def beta(self):
         if hasattr(self, '_beta'):
@@ -137,11 +138,11 @@ class OLS(Regression):
 
 
 class Ridge(Regression):
+    name = 'Ridge'
 
     def __init__(self, X, y, inversion_method='direct', lmbd=0.1):
         super().__init__(X, y, inversion_method)
         self.lmbd = lmbd
-        self.name = 'Ridge'
 
     @property
     def beta(self):
@@ -153,6 +154,7 @@ class Ridge(Regression):
 
 
 class Lasso(Regression):
+    name = 'Lasso'
 
     def __init__(self, X, y, inversion_method='direct', lmbd=0.1):
         super().__init__(X, y, inversion_method)

### 2019-10-06 assignment_one/article/main.tex
Done?

@@
 \usepackage[]{commath}
-\usepackage[]{todonotes}
 \usepackage[]{graphicx}
@@
 \usepackage[]{siunitx}
+\usepackage[]{booktabs}
 
@@
 \addbibresource{../../FYS-STK4155.bib}
+\usepackage[]{hyperref} 
 \usepackage[noabbrev, capitalize]{cleveref}
@@
 \declaretheorem[style=remark, name=Assumption,
-mdframed={
-}]{assumption}
+]{assumption}
 
@@
 \newcommand{\sgn}{\,\mathrm{sign}{}}
+\newcommand{\train}{\mathrm{train}}
+\newcommand{\test}{\mathrm{test}}
 
@@
 \newcommand{\basis}{\mathcal{B}}
+
+\newcommand*{\github}[1]{\def\@github{\color{gray!70}#1}}
+\newcommand{\githublogo}{\raisebox{-1pt}{\includegraphics[height=9pt]{images/github}}}
+
 %\newcommand{\set}[1]{\left\{\, #1 \right\}}
@@
 	  introduction to the theory and motivation behind the three methods.
+          Implementation details can be found on GitHub:
+
+	  \centering{\href{https://github.com/qTipTip/FYS-STK4155}{\githublogo \, \color{gray!50}\textit{github.com/qTipTip}}}
+
   \end{abstract}
@@
   The \(R^2\)-score measures much of the variation in \( \y \) which can be
-  attributed to a simple linear relation between \( \x \) and \( \y \). It is a
-  ratio, thus a value of one tells us that all the variation in the data can be
-  attributed to an approximate linear relationship between the data \(\x\) and
-  the response variable \( \y \). A value of zero indicates that a non-linear
-  model may be preferrable.
+  attributed to a simple linear relation between \( \x \) and \( \y \).  A
+  value of one tells us that all the variation in the data can be attributed to
+  an approximate linear relationship between the data \(\x\) and the response
+  variable \( \y \). A negative value tells us that the model performs worse
+  than a horizontal line through the data.
 
@@
   \end{equation}
-  See \cref{ap:derivation} for the full derivation. The bias-term measures the
+  The bias-term measures the
   deviation of the expectation of the model function \(f(\data, \beta)\) from
@@
 
+  Given a data-set \( \data = (\x_i, y_i)_{i=1}^n \), we divide it into two disjoint subset \( \data_\train \) and \( \data_\test \)
+  with the purpose of optimizing our model by training on \( \data_\train \) and evaluating model performance on the ``unseen data'' of \( \data_\test \).
+
+  Cross-validation entails repeated such splitting of the data-set into test
+  and training sets, evaluating model performance on each such split, and then
+  finally agregating the results to get a good estimate of model performance
+  over the whole data-set. In \emph{k-fold cross-validation}, the data is
+  specifically subdivided into \( k \) equal parts, where each sub-set takes on
+  the role of \( \data_\test \) exactly once.
+
+  In our implementation we have made use of \textsc{sklearn}'s \textsc{KFold} method, with \( k = 10 \).
+
   \chapter{Numerical Results}
@@
     sharp increase in \( \mathrm{MSE}_{\mathrm{test}} \).}
-    \label{fig:ols_scores}
+    \label{fig:ols_scores_crossval}
   \end{figure}
@@
 
-  We sample \( N = 1000 \) points in each parameter direction and compute the
-  corresponding \( N^2 \) function values of the Franke function both with and
-  without added noise. The noise was selected to be distributed as \( \N(0,
-  1) \) with a signal-to-noise ratio \( \alpha = 0.05\).
+  We randomly sampled \( N = 1000 \) points in the unit-square and compute the
+  corresponding \( N \) function values of the Franke function both with and
+  without added noise. The noise was selected to be distributed as \( \N{0,
+  1} \) with a signal-to-noise ratio \( \alpha = 0.05 \).
 
@@
 
-  In addition to the evaluation metrics, we also compute confidence intervals
-  for each model parameter. These are displayed in \cref{tab:ols_confidence}
-  for each polynomial degree up to and including \( d = 5 \).
+  In addition to the evaluation metrics, we also compute the variance for each
+  model parameter.  These are displayed in \cref{tab:ols_confidence} for each
+  polynomial degree from \(d = 2 \) up to and including \( d = 5 \).
 
@@
 	\end{tabular}
-	\caption{The variance of the model parameters \( beta \) with respect
+	\caption{The variance of the model parameters \( \beta \) with respect
 		to the polynomial degree \( d \). The corresponding basis
@@
 		variance of the model parameters seems to increase with the
-		polynomial degree, which is what is to be expected according to
-	the \emph{bias-variance tradeoff}.}
+		polynomial degree, which is what is to be expected according to the
+		\emph{bias-variance tradeoff}.}
     \label{tab:ols_confidence}
@@
   We now incorporate Ridge and Lasso regression in our numerical experiments.
-  We are still working with the noisy sampled data.
+  Both of these regression techniques rely on a hyper-parameter \( \lambda \)
+  which has to be tuned to the data at hand. By plotting the estimated bias and
+  variance of the  model as a function of the hyper-parameter \( \lambda \) we
+  find the optimal parameter. The fits were performed on the same data as in
+  the aforementioned OLS-test, and the results are presented in
+  \cref{fig:bias_variance_lambda}. However, in this case we only use bi-quintic
+  polynomial surfaces.
+
+  \begin{figure}[htpb]
+  	\centering
+	\subbottom[Ridge regression]{%
+		\includegraphics[width=0.45\linewidth]{images/bias_variance_ridge.pdf}
+	}
+	\subbottom[Lasso regression]{%
+		\includegraphics[width=0.45\linewidth]{images/bias_variance_lasso.pdf}
+	}
+	\caption{Finding the optimal hyper-parameter \( \lambda \) for both
+		Ridge and Lasso regression. The fits were perfomed for
+		bi-quintic polynomial surfaces with noisy data sampled from the
+		Franke function. As we see, the variance is fairly constant,
+	until it suddenly increases. The bias on the other hand converges to
+	zero. The optimal lambda is indicated by a vertical line. Note how the error is
+	constant after a certain value of \( \lambda \). By looking at
+	\cref{fig:ridge_lasso_beta_variance}, this is the point at which all
+	Lasso-parameters are set to zero.}%
+  	\label{fig:bias_variance_lambda}
+  \end{figure}
+
+  We also plot the model parameters \( \beta \) and their standard deviations
+  as a function of \( \lambda \).  It is expected that the parameters \(
+  \beta_\ridge\) tends to zero as \( \lambda \) increases, while the parameters
+  \( \beta_\lasso\) may in addition be set equal to zero. This is verified by
+  the plots in \cref{fig:ridge_lasso_beta_variance}.
+
+  \begin{figure}[htpb]
+  	\centering
+	\subbottom[Ridge regression]{%
+		\includegraphics[width=0.45\linewidth]{images/variance_ridge.pdf}
+	}
+	\subbottom[Lasso regression]{%
+		\includegraphics[width=0.45\linewidth]{images/variance_lasso.pdf}
+	}
+	\caption{The model parameters \( \beta \) with associated standard
+	deviations. We do not label each line to not clutter the plot, as its
+	main purpose is to illustrate how Ridge and Lasso regression, respectively,
+	penalizes model parameters. Note that for the Lasso regression, the model is
+	quite sparse, and in particular, note how the \( \lambda \) where all
+	parameters are zero correspond to the constant behaviour in
+	\cref{fig:bias_variance_lambda}.}% 
+	\label{fig:ridge_lasso_beta_variance}
+  \end{figure}
+
+  In \cref{tbl:ridge_lasso_mse_r2} we present the mean squared error and
+  \(R^2\)-score for fits performed with optimal hyper parameter \( \lambda \)
+  as indicated in \cref{fig:bias_variance_lambda} and compare them to an
+  OLS-fit.
+
+  \begin{table}[htpb]
+  	\centering
+	\caption{The MSE and \( R^2 \) score for the Franke function using
+		bi-quintic polynomial surfaces and the optimal hyper parameters
+		for Ridge and Lasso respectively. According to the reported
+		metrics, the OLS-method performs significantly better than both
+		Ridge and Lasso. The \(R^2\)-scores reported for Ridge and
+	Lasso seem unreasonably low compared to OLS, which may indicate an
+error in the implementation, or the authors lack of experience in the field.}
+  	\label{tbl:ridge_lasso_mse_r2}
+  	\begin{tabular}{llll}
+		\toprule
+		{} & OLS & Ridge & Lasso \\
+		\midrule
+		\( \mathrm{MSE} \) & \num{0.011} & \num{0.080} & \num{0.092} \\
+		\( \mathrm{R}^2 \) & \num{0.879} & \num{0.140} & \num{0.005} \\
+		\bottomrule
+  	\end{tabular}
+  \end{table}
 
@@
 
+  When testing on real data, in order to keep the data-size manageable, we
+  select a small patch from the large dataset.  This patch is subsequently
+  normalized in order to improve numerical properties of the methods.
+  Unfortunately, due to time-constraints, and numerical bugs - we only present
+  the MSE and \(R^2\) scores for the various methods with fixed
+  hyper-parameters lambda for Ridge and Lasso regression originally selected
+  for the previous dataset.  Ideally, hyper-parameter optimzation as done in
+  \cref{fig:bias_variance_lambda} should be performed for this new data-set as
+  well, and further statistical quantities reported.
+
+  The normalized patch is displayed in \cref{fig:sub_terrain}, and the performance metrics in \cref{fig:terrain_metrics}.
+  \begin{figure}[htpb]
+  	\centering
+  	\includegraphics[width=0.8\linewidth]{images/reduced_terrain_data.pdf}
+	\caption{A small patch from the larger data-set displayed in \cref{fig:terrain}. The data has been normalized.}%
+	\label{fig:sub_terrain}
+  \end{figure}
+
+  \begin{figure}[htpb]
+  	\centering
+	\subbottom{
+		\includegraphics[width=0.45\linewidth]{images/terrain_data_mse.pdf}	
+	}
+	\subbottom{
+		\includegraphics[width=0.45\linewidth]{images/terrain_data_r2.pdf}	
+	}
+	\caption{The performance metrics reported for polynomial fits to the
+	small terrain patch. As we can see, the Ordinary Least Squares
+regression seems to be the better chocice for terrain data with high local
+variance. It seems like both Ridge and Lasso struggles with finer features
+going missing, as they penalize large but ``irrelevant'' model parameters.}%
+	\label{fig:terrain_metrics}
+  \end{figure}
+
   \chapter{Conclusion}
+
+  In this project we have studied three kinds of linear regression, namely
+  Ordinary Least Squares, Ridge, and Lasso regression.  We tested the methods
+  on artificial data sampled from the Franke function, and later tested the
+  methods on terrain data.
+
+  Due to some unfortunate implementation errors and time constraints, the study
+  on the terrain data were rather uncomplete, however the one result we did
+  manage to get indicates that for data such as the terrain data presented
+  here, the ordinary least squares method may be slightly better. All numerical
+  methods were tested against \textsc{sklearn}'s methods, and the problems
+  persisted, indicating rather that there is a more fundamental
+  misunderstanding from the authors part.
 
@@
   approaches to locally refined spline spaces have started to gain momentum.
-  We note in particular that of \emph{hierarchical B-splines} (HB),
+  We note in particujlar that of \emph{hierarchical B-splines} (HB),
   \emph{truncated hierarchical B-splines} (THB) and \emph{locally refined
@@
 
+  \if 0
   \appendix
@@
   Combining the two previous equations, we obtain
+  \fi
 

### 2019-12-12 assignment_two/article/main.tex
Working on assignment two

\documentclass[a4paper, oneside, article]{memoir}

\usepackage{minted} 
\usepackage[T1]{fontenc} 
\usemintedstyle{algol}
\usepackage[]{microtype}
\usepackage[]{eulervm}
\usepackage[]{palatino} 
\usepackage[]{amsmath, amssymb} 
\usepackage[]{graphicx} 
\usepackage[]{svg} 
\usepackage[]{bm} 
\usepackage[]{commath} 
\usepackage[]{geometry} 
\usepackage[]{hyperref} 
\usepackage[]{cleveref}

\setlength{\parskip}{1em}
\setlength{\parindent}{0pt}

\newcommand{\OLS}{\mathrm{OLS}}
\newcommand{\Ridge}{\mathrm{ridge}}
\newcommand{\Lasso}{\mathrm{lasso}}
\newcommand{\X}{\bm{X}}
\newcommand{\y}{\bm{y}}
\newcommand{\yhat}{\hat{\bm{y}}}
\newcommand{\transpose}[1]{#1^{\top}}
\newcommand{\mat}[1]{\bm{#1}}
\newcommand{\cost}{\mathcal{C}}

\title{\textsc{Regression and Classification using Feed Forward Neural Networks}}
\author{Ivar Stangeby}
\begin{document}
	\maketitle

	\begin{abstract}

	\end{abstract}
	\chapter{Introduction}

	In this project we make a comparative study on feed forward neural
	networks in a regression and classification setting. In particular, we
	compare feed forward neural networks in this context with the standard
	methods of ordinary least squares for linear regression, and logistic
	regression for classification.

	\chapter{Background}

	\section{Neural Networks}

	\subsection{Construction}
	Neural networks has gained a widespread use in the approximation of
	functional relationships. In it's simplest form, a neural network
	consists of a set of \emph{neurons}, being entities that take as input
	a set number of parameters \( x_1, \ldots, x_n \), and computes a
	linear combination of these (often with added bias):
	\begin{equation}
		z = \sum_{i = 0}^n w_ix_i + b.
	\end{equation}
	The
	neuron then decides on whether to ``fire'' or not, by evaluating a
	so-called scalar activation function, \( a = g(z) \). In the architecture
	discussed herein, namely \emph{fully connected} or \emph{dense} neural
	networks, each neuron takes into account all possible parameters.

	By combining a set number of neurons we obtain what is called a
	\emph{layer}. A neural network typically consists of three types of
	layers: an \emph{input layer}, a set number of \emph{hidden layers},
	and finally an \emph{output layer}.

	\begin{figure}[htbp]
		\centering
		\includegraphics[width=0.8\textwidth]{imgs/fcc_nn.pdf}
		\caption{A simple feed forward neural network. This specific
		network has an input layer of four nodes. Two hidden layers
	consisting of six nodes each, and finally, an output layer of size
two.}%
\label{fig:fc_nn}
	\end{figure}

	For a network consisting of \( L \) layers, with each layer having \(
	n^{[\ell]} \) nodes and with activation function \( g\), the
	activations in each layer are given as 
	\begin{equation}
		a^{[\ell]}_k = g\left(\sum_{j = 1}^{n^{[\ell]}} w_{jk}^{[\ell]} a^{[\ell - 1]}_j  + b^{[\ell]}_k \right)
	\end{equation}
	for \( k = 1, \ldots, n^{[\ell]} \) and \( \ell = 1, \ldots, L \). 
	A matrix formulation reads:
	\begin{equation}
	\mat{a}^{[\ell]} = g\left(\mat{W}^{[\ell]} \mat{a}^{[\ell - 1]} + \mat{b}^{[\ell]}\right).
	\end{equation}
	We use the convention that the network input \( \mat{x} = (x_1, \ldots,
	x_{n^{[0]}}) = \mat{a}^{[0]} \) and the network output \( \yhat =
	\mat{a}^{[L]} \).

	\subsection{Optimization}

	We initialize such a network by specifying the initial weights and
	biases for each layer. That is, the matrices \( \mat{W}^{[\ell]} \) and
	\( \mat{b}^{[\ell]} \) are (most commonly) drawn from some random
	distribution. It is our goal to iteratively optimize these weights and
	biases in order for the network to produce desired output. We therefore
	choose a \emph{cost} function  \( \cost \) which in some way
	incorporates our problem at hand. For example, in regression problems,
	our cost function is typically the mean squared error (MSE) between our
	target-data and our prediction, while in classification, we often use
	cross entropy loss.

	We optimize the network by computing the gradient of the cost function
	\( \cost \) with respect to weights and bias' and perform gradient
	descent. That is
	\begin{align}
		\begin{split}
			\mat{W}^{[\ell]} \leftarrow &\mat{W}^{[\ell]} - \lambda \dpd{\cost}{{\mat{W}^{[\ell]}}}, \\ 
			\mat{b}^{[\ell]} \leftarrow &\mat{b}^{[\ell]}\;\, - \lambda \dpd{\cost}{{\mat{b}^{[\ell]}}}
		\end{split}
	\end{align}
	Here \( \lambda > 0 \) is the \emph{learning rate}.

	\subsection{Neural Network Architecture}

	In this paper we study two use-cases of feed forward neural networks,
	namely regression, and classification. The architectures are tailored
	to the data-sets at hand.

	\paragraph{Credit Card Data}

	The credit-card data-set consists of 33 numerical features describing a
	users payment history.  The task at hand is then to predict whether the
	user is prone to defaulting or not. This is therefore a binary
	classification problem. Our network therefore needs to have an input
	layer of size 33, and an output layer of size two (one probability
	value for each class). We---rather arbitrarily---decide on three hidden
	layers of size 20. We use the \emph{rectified linear unit} as our
	activation function for the hidden-layers, and (since we are predicting
	probabilities) we use the \emph{softmax} function as our
	output-activation. These are defined as follows:
	\begin{align}
		\mathrm{RELU}(\mat{x}) &= \max\{\mat{0}, \mat{x}\}, \\
		\mathrm{Softmax}_k(\mat{x}) &= \frac{\exp{x_k}}{\sum_{i=1}^n \exp{x_i}}.
	\end{align}
	Note that the softmax takes as input a vector, and spits out a scalar,
	and that the RELU takes the element-wise maximum.
	Our cost function \( \cost \) is chosen to be \emph{binary cross
	entropy} (or \emph{log-loss}) defined as
	\begin{equation}
		\cost(\mat{y}, \yhat) = -(\mat{y} \log(\yhat) + (1 - \mat{y}) \log(1 - \yhat)).
	\end{equation}

	\paragraph{Regression Data}

	For our regression task, we create a synthetic data-set by noisy
	sampling of points from the Franke test-function, and our goal is to
	create a neural network that approximates the Franke function. Thus,
	our network should take as input a set of two values, and return a
	scalar. Again, we use a rectified linear unit as our activation
	function for the hidden layers, however, since we want an unrestrained
	output range, we do not add the softmax activation to the final layer,
	as this would squeeze the network output between zero and one.
	Our cost function is the mean squared error (MSE):
	\begin{equation}
		\cost(\mat{y}, \yhat) = \frac{1}{n} (\y - \yhat)^T (\y - \yhat).
	\end{equation}

	\chapter{Implementation}

	While implementing backpropagation is a great exercise in programming,
	due to time-constraints, we have had to resort to using machine
	learning frameworks in the implementation.

	For defining our neural networks we use the Pytorch-library, and for
	standard methods like linear regression and logistic regression, we
	resort to the scikit-learn toolkit. In addition to this, we use the
	Skorch-library for wrapping the Pytorch modules in a scikit-learn
	compatible API.

	\subsection{Data Wrangling}

	We subclass the PyTorch dataset-class for loading the credit card data
	and the sampled Franke data as shown in \cref{lst:datasets}. This
	enables us to write lean training loops as the \texttt{DataLoader}
	class works as a batched iterator.

	\subsection{Neural Network}

	We implement a generic neural network that will suffice for both the
	credit card data (the classification task) and the Franke sampled data
	(the regression task). By subclassing \texttt{torch.nn.Module}, we only
	need to register the different layers, and implement the
	\texttt{forward}-method. The auto-grad framework then takes care of the
	back-propagation of gradients for us. We allow for the user to specify
	the number of input features, the number of output features, the number
	of hidden layers and the the number of hidden units (which we take as
	constant across the hidden layers). Furthermore, we also with to be
	able to specify the final activation function, as this needs to be
	specially tailored (as much else) to the task at hand. The generic architecture is layed out in
	\cref{lst:neural_network}.

	\begin{listing}
	\begin{minted}[frame=lines,tabsize=3]{python}
class CreditCardData(torch.nn.utils.data.Dataset):
	def __init__(self, *args, **kwargs):
		super().__init__()

		self.X, self.y = self.fetch_data(*args, **kwargs)
		self.num_features, self.num_items = self.X.shape

	def __getitem__(self, index):
		return self.X[index], self.y[index]

	def __len__(self):
		return self.num_items

		super().__init__()	
	def fetch_data(self):
		# Handles the loading and preprocessing of data
		...

class FrankeData(torch.nn.utils.data.Dataset):
	...
	\end{minted}
	\caption{Interfacing the data-sets with a PyTorch API, which facilitates easy iteration as shown in \cref{lst:training_loop}}
	\label{lst:datasets}
	\end{listing}

	\begin{listing}
	\begin{minted}[frame=lines,tabsize=3]{python}
credit_card_data = CreditCardData('path_to_dataset')
credit_card_loader = DataLoader(credit_card_data, batch_size=64)

# training loop
for (X_batch, y_batch) in credit_card_loader:
	...
	\end{minted}
	\caption{A simple training loop using PyTorch data-loaders.}
	\label{lst:training_loop}
	\end{listing}

	\begin{listing}
	\begin{minted}[frame=lines,tabsize=3]{python}
class GenericNN(torch.nn.Module):

	def __init__(
		self, num_in_features, num_out_features,
		num_hidden_layers=3,
		num_hidden_features=20,
		activation=torch.nn.Relu,
		final_activation=torch.nn.Identity
	):	
		super().__init__()
		self.input_layer = torch.nn.Linear(num_in_features, 
			num_hidden_features)
		self.hidden_layers = torch.nn.ModuleList(
		[
			nn.Linear(num_hidden_features, num_hidden_features)
			for i in range(num_hidden_layers) 
		])
		self.output_layer = torch.nn.Linear(num_hidden_features,
			num_out_features)
		self.g = activation
		self.f = final_activation

	def forward(self, x):
		x = self.g(self.input_layer(x))
		for layer in self.hidden_layers:
			x = self.g(layer(x))
		x = self.f(self.output_layer(x))

		return x
	\end{minted}	
	\caption{A generic dense neural network realized in the PyTorch
	framework. By implementing the forward method, gradients are kept track
	of automatically, and hence no consideration is to be made for the backward
	propagation.}
	\label{lst:neural_network}
	\end{listing}

	\subsection{Classical Methods}

	For the classical methods of (regularized) ordinary least squares and
	logistic regression, we simply use the scikit-learn toolkit. These
	classifiers can be trained by calling \texttt{fit}-method. By utilizing
	cross validation methods, we optimize the hyper-parameters. This will
	be discussed further in the results-section. As an example, for the credit-card dataset, we
	train the classifiers shown in \cref{lst:classifier_credit_card}, performing a grid-search for the best regularization strength:
	\begin{listing}
	\begin{minted}[frame=lines, tabsize=3]{python}
l1_params = [{'penalty':['l1'],'C':np.logspace(-4, 4, 40),\
		'solver':['liblinear']}]
l2_params = [{'penalty':['l2'],'C':np.logspace(-4, 4, 40),\
		'solver':['liblinear']}]

clf_l1 = GridSearchCV(LogisticRegression(), l1_params, cv=5,\
	verbose=True, n_jobs=-1)
clf_l2 = GridSearchCV(LogisticRegression(), l2_params, cv=5,\
	verbose=True, n_jobs=-1)

clf_l1.fit(X, y)
clf_l2.fit(X, y)
	\end{minted}
	\caption{Setting up and training the logistic regressor. We use both
	\(L^1\) and \( L^2\)-regularization, corresponding to Lasso and
Ridge-regression, respectively. We find the optimal regularization-strength by
5-fold cross validation over the data-set. The fits are parallelized over all
available CPU-cores, to speed up the process.}
	\label{lst:classifier_credit_card}
	\end{listing}

	\section{Numerical Results}

	In this section, we present the findings from our numerical experiments.

	\subsection{Credit card data}

	We train our sklearn-classifiers as in
	\cref{lst:classifier_credit_card}. We do a hyper-parameter search for
	the best regularization-strength for both \( L^1 \) and \( L^2 \)
	regularization. We note that \emph{regularization strength} in the
	world of scikit-learn is the inverse of what might be the convention.
	Thus, a regularizatin strength of \( \lambda \) corresponds to the
	penalization coefficient \(  1 / \lambda \). With this in mind, we
	display the results in \cref{fig:credit_cards_classical}. Note how, for
	small values of \( \lambda \), the regularization effectively sets all
	weights to zero. The fact that the network still predicts correctly
	about 78.6\% of the time is due to the fact that the network now only
	predicts zeros, and the data-set consists of about 78.6\% zeros. In the
	other limit, as \( \lambda \) increases, the regularization term is
	effectively set to zero, and both methods converge to non-penalized
	logistic regression. Incidentally, it is also this non-penalized method
	that performs the best with a classification accuracy of \( 81.6 \% \).

	The neural network classifier is trained using the
	\texttt{skorch}-framework, which acts as an interface to scikit-learn
	for PyTorch. Using this allows for  simple hyper-parameter optimization
	as in scikit-learn, and one may test for instance network depth, width,
	and regularization strength. Unfortunately, due to time constraints, we
	were only able to perform a learning-rate optimization. However, the
	implementation in \cref{lst:classification_nn} allows for other sets of
	hyper-parameters to be tested.

	\begin{figure}[htpb]
		\centering
		\includegraphics[width=0.8\linewidth]{imgs/credit_card_classification.pdf}
		\caption{The results of the \( L^1 \) and \( L^2 \) regularized logistic regression. 
			The two regularization types seems to converge to
			exactly the same classification accuracy as
			non-penalized logistic regression, with this value being 81.6\%. As \( \lambda \)
			increases, the penalization term is set to zero, hence
			converges to non-penalized regression. Why this
			coincides with the best fit is not clear to me.
		}%
		\label{fig:credit_cards_classical}
	\end{figure}

	\begin{listing}
	\begin{minted}[frame=lines, tabsize=3]{python}
net = RegularizedNeuralNetClassifier(
    GenericNN,
    module__input_size=credit_card_data.num_features,
    module__num_hidden_layers=3,
    module__output_size=2,
    criterion=torch.nn.CrossEntropyLoss,
    train_split=CVSplit(cv=5, random_state=42),
    optimizer=torch.optim.SGD,
    max_epochs=5,
    optimizer__momentum=0.9
)

params = {
    'lr': np.logspace(-4, 0, 30),
    # these hyper-params can be trained for as well
    # 'module__num_hidden_layers' : range(0, 5),
    # 'module__hidden_layer_size' : [5, 10, 15, 20]
    # 'module__regularization = ['l2', 'none', 'l1'] 
    # 'module__alpha = np.logspace(-10, 10, 40) # regularization str
}
clf = GridSearchCV(
    net,
    params,
    n_jobs=12,
    cv=5
)
clf.fit(credit_card_data.X.numpy(), credit_card_data.y.numpy())
	\end{minted}	
	\caption{By wrapping the GenericNN in a RegularizedNeuralNetClassifier
	(see \cref{lst:rnnc}) we may train the PyTorch model using the
	GridSearchCV-functionality from scikit-learn.}
	\label{lst:classification_nn}
	\end{listing}

	The results are shown in \cref{fig:credit_card_nn}, and as we see, the
	neural network, even with its basic architecture, outperforms the
	classical methods by a fair bit. The optimal learning rate yields a
	classification accuracy of 82.54\%.

	\begin{figure}[htpb]
		\centering
		\includegraphics[width=0.8\linewidth]{imgs/classification_credit_cards_nn.pdf}
		\caption{The classification accuracy as a function of the
			learning rate for the generic neural network. At the
			opimal learning rate, the network achieves a prediction
			accuracy of approximately 82.54\%, which is not
			insignificantly better than all the classical methods.
			The filled blue regions represent the standard
			deviations as reported by \texttt{GridSearchCV}. Note
			how the network at performs it's worst by guessing only
			zeros, similarily to the overpenalized logistic
		regression. The network is trained over 20 epochs, using SGD
	with momentum and 5-fold cross validation. }%
		\label{fig:credit_card_nn}
	\end{figure}
\appendix

\chapter{Regularized neural network}
	\begin{listing}
	\begin{minted}[frame=lines, tabsize=3]{python}
class RegularizedNeuralNetClassifier(NeuralNetClassifier):

    def __init__(self, module, alpha=1, regularizer='none', 
    		criterion=torch.nn.NLLLoss,
                train_split=CVSplit(5, stratified=True),
                classes=None,
                *args,
                **kwargs):
        self.alpha = alpha
        self.regularizer = regularizer

        if 'regularizer' in kwargs.keys():
            kwargs.pop('regularizer')
        if 'alpha' in kwargs.keys():
            kwargs.pop('alpha')

        super().__init__(module, *args, criterion=criterion, \
		train_split=train_split, classes=classes, **kwargs)

    def get_loss(self, y_pred, y_true, *args, **kwargs):
        loss = super().get_loss(y_pred, y_true, *args, **kwargs)

        if self.regularizer == 'none':
            pass
        elif self.regularizer == 'l1':
            loss += self.alpha * sum([w.abs().sum() 
	    					for w in self.module_.parameters()])
        elif self.regularizer == 'l2':
            loss += self.alpha * sum([(w.abs() ** 2).sum() 
	    					for w in self.module_.parameters()])
        return loss

	\end{minted}	
	\caption{We implemented a regularized neural network classifier in
	order to be able to compare the network with both \( L^2 \) and \( L^1
	\) regularization applied to the network weights. Unfortunately, we were not
	able to perform the required numerical tests in time, as the parameter space is
	rather large.}
	\label{lst:rnnc}
	\end{listing}

	\subsection{Franke function}
\end{document}

### 2019-12-12 assignment_two/src/models/logistic_regression.py

@@ -11,36 +11,67 @@ from sklearn.model_selection import GridSearchCV
 from assignment_two.src.utils.data_loaders import CreditCardData
 
 credit_card_data = CreditCardData()
+print(credit_card_data.num_features)
 X = credit_card_data.X.numpy()
 y = credit_card_data.y.numpy()
 
+print(X.shape)
+print(1 - sum(y) / len(y))
 model_accuracy = []
+
+ol_params = [
+    {'penalty': ['none'],
+     'solver': ['lbfgs']
+     }
+]
 
 l1_params = [
     {
         'penalty': ['l1'],
-        'C': np.logspace(-4, 4, 40),
+        'C': np.logspace(-10, 10, 40),
         'solver': ['liblinear']
     },
 ]
 l2_params = [
     {
         'penalty': ['l2'],
-        'C': np.logspace(-4, 4, 40),
+        'C': np.logspace(-10, 10, 40),
         'solver': ['liblinear']
     }
 ]
 
+clf_ol = GridSearchCV(LogisticRegression(), ol_params, cv=5, verbose=True, n_jobs=-1)
 clf_l1 = GridSearchCV(LogisticRegression(), l1_params, cv=5, verbose=True, n_jobs=-1)
 clf_l2 = GridSearchCV(LogisticRegression(), l2_params, cv=5, verbose=True, n_jobs=-1)
 
+clf_ol.fit(X, y)
 clf_l1.fit(X, y)
 clf_l2.fit(X, y)
 
+l1_l = clf_l1.cv_results_['mean_test_score'] - clf_l1.cv_results_['std_test_score']
+l1_h = clf_l1.cv_results_['mean_test_score'] + clf_l1.cv_results_['std_test_score']
+l2_l = clf_l2.cv_results_['mean_test_score'] - clf_l2.cv_results_['std_test_score']
+l2_h = clf_l2.cv_results_['mean_test_score'] + clf_l2.cv_results_['std_test_score']
+
+clf_best = clf_l1.best_estimator_
+clf_best.fit(X, y)
+
+
+
 plt.semilogx(l1_params[0]['C'], clf_l1.cv_results_['mean_test_score'], label='l1', color='blue')
 plt.semilogx(l2_params[0]['C'], clf_l2.cv_results_['mean_test_score'], label='l2', color='green')
+# plt.fill_between(l1_l, l1_h, alpha=0.3, color='blue')
[…]

### 2019-12-12 assignment_two/src/models/logistic_regression_nn.py

@@ -1,0 +1,106 @@
+import numpy as np
+from skorch import NeuralNetClassifier
+from sklearn.model_selection import GridSearchCV
+import torch.nn as nn
+import torch
+import torch.nn.functional as F
+import matplotlib.pyplot as plt
+
+from skorch.dataset import CVSplit
+
+from assignment_two.src.utils.data_loaders import CreditCardData
+
+
+class GenericNN(nn.Module):
+
+    def __init__(self, num_input_features=10, num_output_features=10, num_hidden_layers=1, num_hidden_features=20, activation=F.relu,
+                 alpha=1, regularization='none'):
+        super().__init__()
+
+        self.input_layer = nn.Linear(num_input_features, num_hidden_features).double()
+        self.hidden_layers = nn.ModuleList(
+            [nn.Linear(num_hidden_features, num_hidden_features).double() for i in range(num_hidden_layers)])
+        self.output_layer = nn.Linear(num_hidden_features, num_output_features).double()
+        self.f = activation
+
+    def forward(self, x):
+        x = self.f(self.input_layer(x))
+        for l in self.hidden_layers:
+            x = self.f(l(x))
+        x = self.output_layer(x)
+        return x
+
+
+class RegularizedNeuralNetClassifier(NeuralNetClassifier):
+
+    def __init__(self, module, alpha=1, regularizer='none', criterion=torch.nn.NLLLoss,
+                 train_split=CVSplit(5, stratified=True),
+                 classes=None,
+                 *args,
+                 **kwargs):
+        self.alpha = alpha
+        self.regularizer = regularizer
+
+        if 'regularizer' in kwargs.keys():
+            kwargs.pop('regularizer')
+        if 'alpha' in kwargs.keys():
+            kwargs.pop('alpha')
+
+        super().__init__(module, *args, criterion=criterion, train_split=train_split, classes=classes, **kwargs)
+
+    def get_loss(self, y_pred, y_true, *args, **kwargs):
+        loss = super().get_loss(y_pred, y_true, *args, **kwargs)
+
+        if self.regularizer == 'none':
+            pass
+        elif self.regularizer == 'l1':
+            loss += self.alpha * sum([w.abs().sum() for w in self.module_.parameters()])
+        elif self.regularizer == 'l2':
[…]

### 2019-12-17 assignment_two/article/main.tex
Added best-parameter scatterplot.

@@
-\documentclass[a4paper, oneside, article]{memoir}
+\documentclass[a4paper, 11pt, oneside, article]{memoir}
+\newsubfloat{figure}
+\linespread{1.25}
 
@@
 \usepackage[]{microtype}
+\usepackage[]{siunitx} 
 \usepackage[]{eulervm}
+\usepackage[]{booktabs} 
 \usepackage[]{palatino} 
@@
 \usepackage[]{geometry} 
+\newcommand*{\github}[1]{\def\@github{\color{gray!70}#1}}
+\newcommand{\githublogo}{\raisebox{-1pt}{\includegraphics[height=9pt]{imgs/github}}}
+\usepackage[backend=biber]{biblatex} 
+\addbibresource{../../FYS-STK4155.bib}
+
 \usepackage[]{hyperref} 
-\usepackage[]{cleveref}
+\usepackage[noabbrev, capitalize]{cleveref}
 
@@
 	 \begin{abstract}
+		 In this project we compare the performance of dense neural
+		 network architectures to more classical methods of linear and
+		 logistic regression in the context of fault detection of
+		 credit card loans and linear regression of the franke function.
 
+		 Our results indicate that for classification tasks, neural
+		 nets has a slight edge over non-regularized logistic
+		 regression, with the regularized methods falling behind. For
+		 regression, again the non-regularized methods come out on top,
+		 with the neural net and the ordinary least squares methods
+		 winning by a long shot.
+
+	 The numerical experiments are performed using \textsc{Python} with the
+	 \texttt{scikit-learn} library resonsible for fitting the models,
+	 \texttt{PyTorch} for handling data-sets and representing the nerual
+	 network architecture, and finally \texttt{skorch} for wrapping the
+	 neural networks in a \texttt{scikit-learn} friendly API. All relevant code can
+	 be found on GitHub:
+
+		  \centering{\href{https://github.com/qTipTip/FYS-STK4155}{\githublogo \, \color{gray!50}\textit{github.com/qTipTip}}}
 	  \end{abstract}
@@
 	regression for classification.
+
+	Neural networks are a very flexible framework in a wide range of tasks.
+	In this day and age where artificial intelligence is gaining traction,
+	it is certainly of interest to see how these methods perform in
+	settings where many classical methods have been developed.
 
@@
 	For our regression task, we create a synthetic data-set by noisy
-	sampling of points from the Franke test-function, and our goal is to
-	create a neural network that approximates the Franke function. Thus,
-	our network should take as input a set of two values, and return a
-	scalar. Again, we use a rectified linear unit as our activation
-	function for the hidden layers, however, since we want an unrestrained
-	output range, we do not add the softmax activation to the final layer,
-	as this would squeeze the network output between zero and one.
-	Our cost function is the mean squared error (MSE):
+	sampling of points from the Franke test-function
+	\cite{frankeCriticalComparisonMethods1979}, and our goal is to create a
+	neural network that approximates the Franke function. Thus, our network
+	should take as input a set of two values, and return a scalar. Again,
+	we use a rectified linear unit as our activation function for the
+	hidden layers, however, since we want an unrestrained output range, we
+	do not add the softmax activation to the final layer, as this would
+	squeeze the network output between zero and one.  Our cost function is
+	the mean squared error (MSE):
 	\begin{equation}
@@
 	able to specify the final activation function, as this needs to be
-	specially tailored (as much else) to the task at hand. The generic architecture is layed out in
-	\cref{lst:neural_network}.
+	specially tailored (as much else) to the task at hand. The generic
+	architecture is layed out in \cref{lst:neural_network}.
 
@@
 	cross validation methods, we optimize the hyper-parameters. This will
-	be discussed further in the results-section. As an example, for the credit-card dataset, we
-	train the classifiers shown in \cref{lst:classifier_credit_card}, performing a grid-search for the best regularization strength:
+	be discussed further in the results-section. As an example, for the
+	credit-card dataset, we train the classifiers shown in
+	\cref{lst:classifier_credit_card}, performing a grid-search for the
+	best regularization strength.
 	\begin{listing}
@@
 
-	In this section, we present the findings from our numerical experiments.
+	In this section, we present the findings from our numerical
+	experiments.
 
 	\subsection{Credit card data}
+
+	For the credit card data, we are solving a classification problem. It
+	would seem reasonable to use \emph{classification accuracy} as the
+	scoring metric of choice. However, upon further inspection of the
+	data-set, we see that the data set consists of highly unbalanced
+	classes, with about 78.6\% zeros, and the rest ones. This means, that a
+	network only predicting zeroes, will be correct 78.6\% of the time.
+	Therefore, in order to better illuminate the model performance, we use
+	in addition the \emph{balanced accuracy score} (bACC), which takes into
+	account the class imbalances. It is defined as such:
+	\begin{equation}
+		\mathrm{bACC} = \frac{TPR + TNR}{2},
+	\end{equation}
+	where \( TPR \) and \( TNR \) is the true positive rate, and the
+	true negative rate of the model, respecitvely.
 
@@
 	The neural network classifier is trained using the
-	\texttt{skorch}-framework, which acts as an interface to scikit-learn
-	for PyTorch. Using this allows for  simple hyper-parameter optimization
-	as in scikit-learn, and one may test for instance network depth, width,
-	and regularization strength. Unfortunately, due to time constraints, we
-	were only able to perform a learning-rate optimization. However, the
-	implementation in \cref{lst:classification_nn} allows for other sets of
-	hyper-parameters to be tested.
+	\texttt{skorch}-framework \cite{SkorchdevSkorch2019}, which
+	acts as an interface to scikit-learn for PyTorch. Using this allows for
+	simple hyper-parameter optimization as in scikit-learn, and one may
+	test for instance network depth, width, and regularization strength.
+	Unfortunately, due to time constraints, we were only able to perform a
+	learning-rate optimization.  However, the implementation in
+	\cref{lst:classification_nn} allows for other sets of hyper-parameters
+	to be tested.
 
@@
 		\centering
-		\includegraphics[width=0.8\linewidth]{imgs/credit_card_classification.pdf}
-		\caption{The results of the \( L^1 \) and \( L^2 \) regularized logistic regression. 
-			The two regularization types seems to converge to
-			exactly the same classification accuracy as
-			non-penalized logistic regression, with this value being 81.6\%. As \( \lambda \)
-			increases, the penalization term is set to zero, hence
-			converges to non-penalized regression. Why this
-			coincides with the best fit is not clear to me.
-		}%
+		\subbottom[Optimizing for classification accuracy]{\includegraphics[width=0.44\linewidth]{imgs/credit_card_classification.pdf}}
+		\subbottom[Optimizing for balanced classification accuracy]{\includegraphics[width=0.44\linewidth]{imgs/credit_card_classification_balanced_accuracy.pdf}}
+		\caption{The results of the \( L^1 \) and \( L^2 \) regularized
+			logistic regression.  The two regularization types
+			seems to converge to exactly the same classification
+			accuracy as non-penalized logistic regression, with
+			this value being 81.6\%. As \( \lambda \) increases,
+		the penalization term is set to zero, hence converges to
+	non-penalized regression. Why this coincides with the best fit is not
+clear to me.  }%
 		\label{fig:credit_cards_classical}
@@
 		\centering
-		\includegraphics[width=0.8\linewidth]{imgs/classification_credit_cards_nn.pdf}
+		\subbottom[]{\includegraphics[width=0.45\linewidth]{imgs/classification_credit_cards_nn.pdf}}
+		\subbottom[]{\includegraphics[width=0.45\linewidth]{imgs/classification_credit_cards_nn_balanced_accuracy.pdf}}
+
 		\caption{The classification accuracy as a function of the
@@
 			accuracy of approximately 82.54\%, which is not
-			insignificantly better than all the classical methods.
+			marginally better than all the classical methods.
 			The filled blue regions represent the standard
@@
 	\end{figure}
+
+	\subsection{Franke function}
+
+	We take two approaches to regression of the Franke function. We
+	approximate this functional relationship using a dense neural network,
+	and we compare this to a standard polynomial fit of the same data. We
+	choose the polynomial degree to be five. Such a polynomial fit were
+	discussed in more detail in
+	\citetitle{stangebyFYSSTK4155Project2019}\cite{stangebyFYSSTK4155Project2019}.
+	We re-use the optimal regularization parameters found for Ridge and
+	Lasso-regression for bi-quintic polynomials \cite[Section 4, p.
+	16]{stangebyFYSSTK4155Project2019}:
+	\begin{align}
+		\label{eq:optimal_reg_params}
+		\lambda_{\mathrm{LASSO}} = \num{4.23e-2}, \quad \lambda_{\mathrm{RIDGE}} = \num{2.36e2}.
+	\end{align}
+
+	Our neural network architecture consists of three hidden layers with 20
+	nodes each, two nodes features, and one output node. We use
+	\textsc{Relu} as our activation function, and our output-activation is
+	the identity.
+
+	We sample \( N = 2000 \) points of the Franke function, and we add
+	normally distributed noise with zero mean and unit deviation, with a
+	signal to noise ratio of 0.1.
+
+	As our performance metric we use the mean-squared error, and we perform
+	hyper-parameter optimization using grid-search for the learning rate. A
+	5-fold cross-validation scheme is used, as in the credit-card
+	experiments.
+
+	The grid search results are shown in \cref{fig:regression_nn_cv}.
+	\begin{figure}[htpb]
+		\centering
+		\includegraphics[width=0.8\linewidth]{imgs/regression_nn.pdf}
+		\caption{The mean squared error as a function of learning rate.
+		The network was trained using 5-fold cross-validation with
+	momentum SGD over 20 epochs. The lowest MSE reported was approximately
+0.012 signifying that the network does not fit to the noise.}%
+	\label{fig:regression_nn_cv}
+	\end{figure}
+
+	In \cref{tab:optimal_models_scores} we display the mean-squared error
+	and the \(R^2\)-score of the optimized Lasso, Ridge and neural-net
+	models. The numbers displayed clearly shows that the neural-network and
+	the ordinary least squares outperforms the two regularized methods
+	significantly.
+
+	\begin{table}[htpb]
+		\centering
+		\caption{The scores reported for the optimal Ridge, Lasso and neural-net models.}
+		\label{tab:optimal_models_scores}
+		\begin{tabular}{lrr}
+			\toprule
+			{Model} & MSE   & \( R^2 \) \\
+			\midrule			
+			NNet & 0.012 & 0.98 \\
+			OLS & 0.002 & 0.97 \\
+			Ridge & 0.06 & 0.25\\
+			Lasso & 0.03 & 0.64\\
+			\bottomrule
+		\end{tabular}
+	\end{table}
+
+\clearpage
 \appendix
@@
 \chapter{Regularized neural network}
-	\begin{listing}
+	\begin{listing}[htbp]
 	\begin{minted}[frame=lines, tabsize=3]{python}
@@
 
-	\subsection{Franke function}
+	\section{Conclusion}%
+	\label{sec:conclusion}
+
+	In this project we have compared artifical neural network-based methods
+	to more classical regression methods in the context of surface-fitting
+	and classification. In both cases, it seems like regularized models
+	perform siginificantly worse than their counterparts.
+
+	In the classification of credit-card defaults, the numbers reported indicate
+	that the neural network performs slightly better than logistic
+	regression. However, upon inspection of the credit-card data-set, it is
+	evident that the classes are highly unbalanced, meaning that the
+	numerical results are hard to interpret. Using the standard
+	classification accuracy score is highly inconclusive in such unbalanced
+	cases, and it is hard to know exactly how much correction the balanced
+	classification accuracy offers.
+
+	For the regression of the Franke function data-set we reuse optimal
+	parameters from \cite{stangebyFYSSTK4155Project2019}, and the
+	unpenalized least squares and the neural network methods come out on
+	top. Interestingly, the \( R^2 \)-scores reported in this project for
+	Ridge and Lasso regression are wildly different from those reported in
+	\cite{stangebyFYSSTK4155Project2019}, which indicates faults in the
+	previous implementation, and the \( R^2 \)-scores for both OLS and the
+	neural network seem suspiciously high, which may indicate overfitting
+	on the data-set.  Some additional investigation should be carried out,
+	specifically regularizing the weights in the neural net using the
+	implementation in \cref{lst:rnnc}. Alas, this will have to be for
+	another time.
+
+\printbibliography
 \end{document}

### 2019-12-17 assignment_two/src/models/logistic_regression_balanced_accuracy.py

@@ -1,0 +1,78 @@
+"""
+This file contains code for running logistic regression on the CreditCardData-set. We use the built in LogisticRegression
+model from sklearn, and perform a hyper-parameter optimization using grid search. We explore both l1 and l2 regularization.
+"""
+import numpy as np
+import matplotlib.pyplot as plt
+
+from sklearn.linear_model import LogisticRegression
+from sklearn.metrics import make_scorer, cohen_kappa_score, balanced_accuracy_score
+from sklearn.model_selection import GridSearchCV
+
+from assignment_two.src.utils.data_loaders import CreditCardData
+
+credit_card_data = CreditCardData()
+print(credit_card_data.num_features)
+X = credit_card_data.X.numpy()
+y = credit_card_data.y.numpy()
+
+print(X.shape)
+print(1 - sum(y) / len(y))
+model_accuracy = []
+
+ol_params = [
+    {'penalty': ['none'],
+     'solver': ['lbfgs']
+     }
+]
+
+l1_params = [
+    {
+        'penalty': ['l1'],
+        'C': np.logspace(-10, 10, 40),
+        'solver': ['liblinear']
+    },
+]
+l2_params = [
+    {
+        'penalty': ['l2'],
+        'C': np.logspace(-10, 10, 40),
+        'solver': ['liblinear']
+    }
+]
+
+scorer = make_scorer(balanced_accuracy_score)
+
+clf_ol = GridSearchCV(LogisticRegression(), ol_params, cv=5, verbose=True, n_jobs=-1, scoring=scorer)
+clf_l1 = GridSearchCV(LogisticRegression(), l1_params, cv=5, verbose=True, n_jobs=-1, scoring=scorer)
+clf_l2 = GridSearchCV(LogisticRegression(), l2_params, cv=5, verbose=True, n_jobs=-1, scoring=scorer)
+
+clf_ol.fit(X, y)
+clf_l1.fit(X, y)
+clf_l2.fit(X, y)
+
+l1_l = clf_l1.cv_results_['mean_test_score'] - clf_l1.cv_results_['std_test_score']
+l1_h = clf_l1.cv_results_['mean_test_score'] + clf_l1.cv_results_['std_test_score']
+l2_l = clf_l2.cv_results_['mean_test_score'] - clf_l2.cv_results_['std_test_score']
[…]

### 2019-12-17 assignment_two/src/models/logistic_regression_f1.py

@@ -1,0 +1,75 @@
+"""
+This file contains code for running logistic regression on the CreditCardData-set. We use the built in LogisticRegression
+model from sklearn, and perform a hyper-parameter optimization using grid search. We explore both l1 and l2 regularization.
+"""
+import numpy as np
+import matplotlib.pyplot as plt
+
+from sklearn.linear_model import LogisticRegression
+from sklearn.model_selection import GridSearchCV
+
+from assignment_two.src.utils.data_loaders import CreditCardData
+
+credit_card_data = CreditCardData()
+print(credit_card_data.num_features)
+X = credit_card_data.X.numpy()
+y = credit_card_data.y.numpy()
+
+print(X.shape)
+print(1 - sum(y) / len(y))
+model_accuracy = []
+
+ol_params = [
+    {'penalty': ['none'],
+     'solver': ['lbfgs']
+     }
+]
+
+l1_params = [
+    {
+        'penalty': ['l1'],
+        'C': np.logspace(-10, 10, 40),
+        'solver': ['liblinear']
+    },
+]
+l2_params = [
+    {
+        'penalty': ['l2'],
+        'C': np.logspace(-10, 10, 40),
+        'solver': ['liblinear']
+    }
+]
+
+clf_ol = GridSearchCV(LogisticRegression(), ol_params, cv=5, verbose=True, n_jobs=-1, scoring='f1')
+clf_l1 = GridSearchCV(LogisticRegression(), l1_params, cv=5, verbose=True, n_jobs=-1, scoring='f1')
+clf_l2 = GridSearchCV(LogisticRegression(), l2_params, cv=5, verbose=True, n_jobs=-1, scoring='f1')
+
+clf_ol.fit(X, y)
+clf_l1.fit(X, y)
+clf_l2.fit(X, y)
+
+l1_l = clf_l1.cv_results_['mean_test_score'] - clf_l1.cv_results_['std_test_score']
+l1_h = clf_l1.cv_results_['mean_test_score'] + clf_l1.cv_results_['std_test_score']
+l2_l = clf_l2.cv_results_['mean_test_score'] - clf_l2.cv_results_['std_test_score']
+l2_h = clf_l2.cv_results_['mean_test_score'] + clf_l2.cv_results_['std_test_score']
+
+clf_best = clf_l1.best_estimator_
+clf_best.fit(X, y)
+
[…]

### 2019-12-17 assignment_two/src/models/logistic_regression_kappa.py

@@ -1,0 +1,78 @@
+"""
+This file contains code for running logistic regression on the CreditCardData-set. We use the built in LogisticRegression
+model from sklearn, and perform a hyper-parameter optimization using grid search. We explore both l1 and l2 regularization.
+"""
+import numpy as np
+import matplotlib.pyplot as plt
+
+from sklearn.linear_model import LogisticRegression
+from sklearn.metrics import make_scorer, cohen_kappa_score
+from sklearn.model_selection import GridSearchCV
+
+from assignment_two.src.utils.data_loaders import CreditCardData
+
+credit_card_data = CreditCardData()
+print(credit_card_data.num_features)
+X = credit_card_data.X.numpy()
+y = credit_card_data.y.numpy()
+
+print(X.shape)
+print(1 - sum(y) / len(y))
+model_accuracy = []
+
+ol_params = [
+    {'penalty': ['none'],
+     'solver': ['lbfgs']
+     }
+]
+
+l1_params = [
+    {
+        'penalty': ['l1'],
+        'C': np.logspace(-10, 10, 40),
+        'solver': ['liblinear']
+    },
+]
+l2_params = [
+    {
+        'penalty': ['l2'],
+        'C': np.logspace(-10, 10, 40),
+        'solver': ['liblinear']
+    }
+]
+
+scorer = make_scorer(cohen_kappa_score)
+
+clf_ol = GridSearchCV(LogisticRegression(), ol_params, cv=5, verbose=True, n_jobs=-1, scoring=scorer)
+clf_l1 = GridSearchCV(LogisticRegression(), l1_params, cv=5, verbose=True, n_jobs=-1, scoring=scorer)
+clf_l2 = GridSearchCV(LogisticRegression(), l2_params, cv=5, verbose=True, n_jobs=-1, scoring=scorer)
+
+clf_ol.fit(X, y)
+clf_l1.fit(X, y)
+clf_l2.fit(X, y)
+
+l1_l = clf_l1.cv_results_['mean_test_score'] - clf_l1.cv_results_['std_test_score']
+l1_h = clf_l1.cv_results_['mean_test_score'] + clf_l1.cv_results_['std_test_score']
+l2_l = clf_l2.cv_results_['mean_test_score'] - clf_l2.cv_results_['std_test_score']
[…]

### 2019-12-17 assignment_two/src/models/logistic_regression_nn_balanced_accuracy.py

@@ -1,6 +1,10 @@
 import numpy as np
+from sklearn.metrics import make_scorer, balanced_accuracy_score
+from skorch import NeuralNetClassifier
 from sklearn.model_selection import GridSearchCV
+import torch.nn as nn
 import torch
+import torch.nn.functional as F
 import matplotlib.pyplot as plt
 
 from skorch.dataset import CVSplit
@@ -31,11 +35,13 @@ params = {
     # 'module__num_hidden_layers' : range(0, 5),
     # 'module__hidden_layer_size' : [5, 10, 15, 20]
 }
+scorer = make_scorer(balanced_accuracy_score)
 clf = GridSearchCV(
     net,
     params,
     n_jobs=12,
-    cv=5
+    cv=5,
+    scoring=scorer
 )
 clf.fit(credit_card_data.X.numpy(), credit_card_data.y.numpy())
 
@@ -46,9 +52,12 @@ plt.fill_between(params['lr'], clf.cv_results_['mean_test_score'] - clf.cv_resul
 plt.scatter(clf.best_params_['lr'], clf.best_score_)
 plt.title(rf'Optimal learning rate $\lambda = {clf.best_params_["lr"] :.04e}$')
 plt.xlabel(rf'Learning rate $\lambda$')
-plt.ylabel(rf'Classification accuracy')
+plt.ylabel(rf'Balanced classification accuracy')
 plt.legend()
-plt.savefig('classification_credit_cards_nn.pdf')
+plt.savefig('classification_credit_cards_nn_f1.pdf')
 plt.show()
 print(f'Best score = {clf.best_score_}')
-np.save('saved_scores/logistic_regression_nn.npy', clf.cv_results_)
+best_clf = clf.best_estimator_
+best_clf.score()
+np.save('saved_scores/logistic_regression_nn_balanced_accuracy.npy', clf.cv_results_)
+

### 2019-12-17 assignment_two/src/models/logistic_regression_nn_f1.py

@@ -1,12 +1,18 @@
 import numpy as np
+from skorch import NeuralNetClassifier
 from sklearn.model_selection import GridSearchCV
+import torch.nn as nn
 import torch
+import torch.nn.functional as F
 import matplotlib.pyplot as plt
 
 from skorch.dataset import CVSplit
 
 from assignment_two.src.models.neural_network import GenericNN, RegularizedNeuralNetClassifier
 from assignment_two.src.utils.data_loaders import CreditCardData
+
+
+
 
 credit_card_data = CreditCardData()
 
@@ -35,7 +41,8 @@ clf = GridSearchCV(
     net,
     params,
     n_jobs=12,
-    cv=5
+    cv=5,
+    scoring='f1'
 )
 clf.fit(credit_card_data.X.numpy(), credit_card_data.y.numpy())
 
@@ -46,9 +53,9 @@ plt.fill_between(params['lr'], clf.cv_results_['mean_test_score'] - clf.cv_resul
 plt.scatter(clf.best_params_['lr'], clf.best_score_)
 plt.title(rf'Optimal learning rate $\lambda = {clf.best_params_["lr"] :.04e}$')
 plt.xlabel(rf'Learning rate $\lambda$')
-plt.ylabel(rf'Classification accuracy')
+plt.ylabel(rf'F1-score')
 plt.legend()
-plt.savefig('classification_credit_cards_nn.pdf')
+plt.savefig('classification_credit_cards_nn_f1.pdf')
 plt.show()
 print(f'Best score = {clf.best_score_}')
-np.save('saved_scores/logistic_regression_nn.npy', clf.cv_results_)
+np.save('logistic_regression_nn_f1.npy', clf.cv_results_)

### 2019-12-17 assignment_two/src/models/logistic_regression_nn_kappa.py

@@ -1,6 +1,10 @@
 import numpy as np
+from sklearn.metrics import make_scorer, cohen_kappa_score
+from skorch import NeuralNetClassifier
 from sklearn.model_selection import GridSearchCV
+import torch.nn as nn
 import torch
+import torch.nn.functional as F
 import matplotlib.pyplot as plt
 
 from skorch.dataset import CVSplit
@@ -9,8 +13,6 @@ from assignment_two.src.models.neural_network import GenericNN, RegularizedNeura
 from assignment_two.src.utils.data_loaders import CreditCardData
 
 credit_card_data = CreditCardData()
-
-NN = GenericNN(num_input_features=credit_card_data.num_features, num_hidden_layers=3, num_output_features=2)
 
 net = RegularizedNeuralNetClassifier(
     GenericNN,
@@ -31,11 +33,15 @@ params = {
     # 'module__num_hidden_layers' : range(0, 5),
     # 'module__hidden_layer_size' : [5, 10, 15, 20]
 }
+
+scorer = make_scorer(cohen_kappa_score)
+
 clf = GridSearchCV(
     net,
     params,
     n_jobs=12,
-    cv=5
+    cv=5,
+    scoring=scorer
 )
 clf.fit(credit_card_data.X.numpy(), credit_card_data.y.numpy())
 
@@ -46,9 +52,9 @@ plt.fill_between(params['lr'], clf.cv_results_['mean_test_score'] - clf.cv_resul
 plt.scatter(clf.best_params_['lr'], clf.best_score_)
 plt.title(rf'Optimal learning rate $\lambda = {clf.best_params_["lr"] :.04e}$')
 plt.xlabel(rf'Learning rate $\lambda$')
-plt.ylabel(rf'Classification accuracy')
+plt.ylabel(rf'Cohens kappa score')
 plt.legend()
-plt.savefig('classification_credit_cards_nn.pdf')
+plt.savefig('classification_credit_cards_nn_kappa.pdf')
 plt.show()
 print(f'Best score = {clf.best_score_}')
-np.save('saved_scores/logistic_regression_nn.npy', clf.cv_results_)
+np.save('saved_scores/logistic_regression_nn_kappa.npy', clf.cv_results_)

### 2019-12-17 assignment_two/src/models/neural_network.py

@@ -1,0 +1,54 @@
+import torch
+from skorch import NeuralNetClassifier
+from skorch.dataset import CVSplit
+from torch import nn as nn
+from torch.nn import functional as F
+
+
+class GenericNN(nn.Module):
+
+    def __init__(self, num_input_features=10, num_output_features=10, num_hidden_layers=1, num_hidden_features=20, activation=F.relu,
+                 alpha=1, regularization='none'):
+        super().__init__()
+
+        self.input_layer = nn.Linear(num_input_features, num_hidden_features).double()
+        self.hidden_layers = nn.ModuleList(
+            [nn.Linear(num_hidden_features, num_hidden_features).double() for i in range(num_hidden_layers)])
+        self.output_layer = nn.Linear(num_hidden_features, num_output_features).double()
+        self.f = activation
+
+    def forward(self, x):
+        x = self.f(self.input_layer(x))
+        for l in self.hidden_layers:
+            x = self.f(l(x))
+        x = self.output_layer(x)
+        return x
+
+
+class RegularizedNeuralNetClassifier(NeuralNetClassifier):
+
+    def __init__(self, module, alpha=1, regularizer='none', criterion=torch.nn.NLLLoss,
+                 train_split=CVSplit(5, stratified=True),
+                 classes=None,
+                 *args,
+                 **kwargs):
+        self.alpha = alpha
+        self.regularizer = regularizer
+
+        if 'regularizer' in kwargs.keys():
+            kwargs.pop('regularizer')
+        if 'alpha' in kwargs.keys():
+            kwargs.pop('alpha')
+
+        super().__init__(module, *args, criterion=criterion, train_split=train_split, classes=classes, **kwargs)
+
+    def get_loss(self, y_pred, y_true, *args, **kwargs):
+        loss = super().get_loss(y_pred, y_true, *args, **kwargs)
+
+        if self.regularizer == 'none':
+            pass
+        elif self.regularizer == 'l1':
+            loss += self.alpha * sum([w.abs().sum() for w in self.module_.parameters()])
+        elif self.regularizer == 'l2':
+            loss += self.alpha * sum([(w.abs() ** 2).sum() for w in self.module_.parameters()])
+        return loss

### 2019-12-17 assignment_two/src/models/regression.py

@@ -1,0 +1,22 @@
+import numpy as np
+from sklearn.linear_model import Lasso, Ridge, LinearRegression
+from sklearn.metrics import mean_squared_error
+from sklearn.model_selection import GridSearchCV
+
+from assignment_two.src.utils.data_loaders import FrankeDataSet
+
+franke_data = FrankeDataSet(num_points=2000, signal_to_noise=0.1, noisy=True, random_seed=5, degree=5)
+print(franke_data.X.shape, franke_data.z_noise.shape)
+
+regressors = [('Lasso', Lasso, 4.23e-2), ('Ridge', Ridge, 2.36e2), ('OLS', LinearRegression, 0)]
+
+for name, reg_type, alpha in regressors:
+    if name != 'OLS':
+        reg = reg_type(alpha=alpha)
+    else:
+        reg = reg_type()
+    reg.fit(franke_data.X, franke_data.z_noise)
+    z_hat = reg.predict(franke_data.X)
+    print(name, 'R2', reg.score(franke_data.X, franke_data.z))
+    print(name, 'MSE', mean_squared_error(franke_data.z, z_hat))
+jk

### 2019-12-17 assignment_two/src/models/regression_nn.py

@@ -1,0 +1,62 @@
+import numpy as np
+import sklearn
+import matplotlib.pyplot as plt
+
+import torch
+from sklearn.metrics import r2_score
+from sklearn.model_selection import GridSearchCV
+from skorch import NeuralNet
+from skorch.dataset import CVSplit
+from torch.utils.data import DataLoader
+
+from assignment_two.src.models.neural_network import GenericNN
+from assignment_two.src.utils.data_loaders import FrankeDataSet
+
+franke_data_train = FrankeDataSet(num_points=2000)
+train = DataLoader(franke_data_train)
+X, y = franke_data_train.xy, franke_data_train.z_noise
+print(X.shape, y.shape)
+net = NeuralNet(
+    GenericNN,
+    module__num_input_features=2,
+    module__num_output_features=1,
+    module__num_hidden_layers=3,
+    module__num_hidden_features=20,
+    module__activation=torch.nn.functional.relu,
+    criterion=torch.nn.MSELoss,
+    optimizer=torch.optim.SGD,
+    max_epochs=100,
+    optimizer__nesterov=True,
+    optimizer__momentum=0.9,
+    # batch_size=20,
+    train_split=CVSplit(cv=5),
+)
+
+params = {
+    'lr': np.logspace(-5, 0, 20, endpoint=False),
+}
+
+reg = GridSearchCV(estimator=net, param_grid=params, scoring='neg_mean_squared_error', n_jobs=12)
+reg.fit(X, y)
+
+print(reg.cv_results_.keys())
+mean_test = -reg.cv_results_['mean_test_score']
+std_test = -reg.cv_results_['std_test_score']
+plt.semilogx(params['lr'], mean_test, label='nn_validation', color='green')
+plt.fill_between(params['lr'], mean_test - std_test,
+                 mean_test + std_test, alpha=0.1)
+# plt.semilogx(params['lr'], clf.cv_results_['mean_train_score'], label='nn_train', color='green')
+plt.scatter(reg.best_params_['lr'], -reg.best_score_)
+plt.title(rf'Optimal learning rate $\lambda = {reg.best_params_["lr"] :.04e}$')
+plt.xlabel(rf'Learning rate $\lambda$')
+plt.ylabel(rf'MSE')
+plt.legend()
+plt.savefig('regression_nn.pdf')
+plt.show()
+print(f'Best score = {-reg.best_score_}')
[…]

### 2019-12-17 assignment_two/src/utils/data_loaders.py

@@ -1,11 +1,12 @@
 from pathlib import Path
 from typing import Tuple
 
+import numpy as np
 import pandas
 import torch
 import sklearn
 from sklearn.compose import ColumnTransformer
-from sklearn.preprocessing import OneHotEncoder, normalize, scale
+from sklearn.preprocessing import OneHotEncoder, normalize, scale, PolynomialFeatures
 from torch.utils.data import Dataset, DataLoader
 
 
@@ -98,5 +99,55 @@ class CreditCardData(Dataset):
         return data
 
 
+class FrankeDataSet(Dataset):
+
+    def __init__(self, num_points, signal_to_noise=0.1, noisy=True, random_seed=None, degree=5):
+        self.num_points = num_points
+        self.random_seed = random_seed
+        self.stn = signal_to_noise
+        self.noisy = noisy
+        self.degree = degree
+        self.polyfit = PolynomialFeatures(degree)
+        if random_seed:
+            np.random.seed(random_seed)
+
+        self._sample()
+
+    def _sample(self):
+        self.xy = np.random.uniform(0, 1, (self.num_points, 2))
+        self.X = self.polyfit.fit_transform(self.xy)
+        self.z = np.array([self._eval(x, y) for x, y in self.xy]).reshape((self.num_points, 1))
+
+        if self.noisy:
+            self.z_noise = self.z + np.random.normal(0, 1, size=self.z.shape) * self.stn
+
+    def _eval(self, x, y):
+        a = 3 / 4 * np.exp(-(9 * x - 2) ** 2 / 4 - (9 * y - 2) ** 2 / 4)
+        b = 3 / 4 * np.exp(-(9 * x + 1) ** 2 / 49 - (9 * y + 1) / 10)
+        c = 1 / 2 * np.exp(-(9 * x - 7) ** 2 / 4 - (9 * y - 3) ** 2 / 4)
+        d = 1 / 5 * np.exp(-(9 * x - 4) ** 2 - (9 * y - 7) ** 2)
+
+        return a + b + c - d
+
+    def __len__(self):
+        return self.num_points
+
+    def __getitem__(self, item):
+        if self.noisy:
+            return self.xy[item], self.z_noise[item]
+        else:
+            return self.xy[item], self.z[item]
+
+
 if __name__ == '__main__':
     D = CreditCardData()
+    F = FrankeDataSet(num_points=10000)
+
[…]

### 2019-12-17 assignment_two/src/models/regression.py
Removed some stray letters.

@@ -19,4 +19,3 @@ for name, reg_type, alpha in regressors:
     z_hat = reg.predict(franke_data.X)
     print(name, 'R2', reg.score(franke_data.X, franke_data.z))
     print(name, 'MSE', mean_squared_error(franke_data.z, z_hat))
-jk

### 2019-12-17 assignment_two/src/models/logistic_regression_nn_f1.py
Added save for logistic regression.

@@ -58,4 +58,4 @@ plt.legend()
 plt.savefig('classification_credit_cards_nn_f1.pdf')
 plt.show()
 print(f'Best score = {clf.best_score_}')
-np.save('logistic_regression_nn_f1.npy', clf.cv_results_)
+np.save('saved_scores/logistic_regression_nn_f1.npy', clf.cv_results_)

### 2019-12-18 assignment_three/src/neural_network.py
Started working on project 3.

@@ -1,0 +1,23 @@
+import torch.nn as nn
+import torch.nn.functional as F
+
+
+class GenericNN(nn.Module):
+
+    def __init__(self, num_input_features=10, num_output_features=10, num_hidden_layers=1, num_hidden_features=20,
+                 activation=F.relu,
+                 alpha=1, regularization='none'):
+        super().__init__()
+
+        self.input_layer = nn.Linear(num_input_features, num_hidden_features).double()
+        self.hidden_layers = nn.ModuleList(
+            [nn.Linear(num_hidden_features, num_hidden_features).double() for i in range(num_hidden_layers)])
+        self.output_layer = nn.Linear(num_hidden_features, num_output_features).double()
+        self.f = activation
+
+    def forward(self, x):
+        x = self.f(self.input_layer(x))
+        for l in self.hidden_layers:
+            x = self.f(l(x))
+        x = self.output_layer(x)
+        return x

### 2019-12-18 assignment_three/src/numerical_solver.py

@@ -1,0 +1,63 @@
+import numpy as np
+import tqdm
+
+
+def ftcs(space_resolution, time_resolution, space_min_max=[0, 1], time_max=1, boundary_conditions=[0, 0],
+         initial_condition=None):
+    """
+    Solves the 1D-heat equation using a forward-in-time centered-in-space discretization scheme.
+
+    :param space_resolution: number of points in spatial direction
+    :param time_resolution: number of points in temporal direction
+    :param space_min_max: the spatial boundary values
+    :param time_max: the time to run the simulation for
+    :param boundary_conditions: the boundary conditions at the space_min_max-values
+    :param initial_conditions: the initial conditions at time = 0, callable.
+    """
+
+    if initial_condition is None:
+        initial_condition: lambda x: 0
+
+    dt = time_max / time_resolution
+    u0, dx = np.linspace(*space_min_max, num=space_resolution, retstep=True)
+    u = np.zeros((time_resolution, space_resolution))
+    u[:, 0] = boundary_conditions[0]
+    u[:, -1] = boundary_conditions[1]
+    u[0] = initial_condition(u0)
+
+    F = dt / dx ** 2
+    print(F)
+    for step in tqdm.trange(time_resolution - 1):
+        for i in range(1, space_resolution - 1):
+            u[step + 1, i] = u[step, i] + F * (u[step, i - 1] - 2 * u[step, i] + u[step, i + 1])
+
+    return u
+
+
+@np.vectorize
+def initial_condition(x):
+    return np.sin(np.pi * x)
+
+
+@np.vectorize
+def exact_solution(x, t):
+    return np.sin(np.pi(x)) * np.exp(-np.pi ** 2 * t)
+
+
+if __name__ == '__main__':
+
+    desired_dx = [1 / 10, 1 / 20, 1 / 30, 1 / 40]
+
+    for dx in desired_dx:
+        dt = 1 / 2 * dx ** 2
+
+        N = int(1 / dx)
+        T = int(1 / dt)
+
+        u = ftcs(N, T, initial_condition=initial_condition)
+        import matplotlib.pyplot as plt
+
+        for step in u:
+            plt.title(f'N = {N} T = {T}')
+            plt.plot(step, color='gray', alpha=0.7)
+        plt.show()

### 2019-12-20 assignment_three/article/main.tex
Working on project 3.

\documentclass[article, a4paper, oneside]{memoir}
\linespread{1.25}

\setlength{\parskip}{1em}
\setlength{\parindent}{0em}
\usepackage[]{eulervm} 
\usepackage[]{palatino} 
\usepackage[]{commath} 
\usepackage[]{microtype} 
\usepackage[]{geometry} 
\usepackage[]{minted} 
\usemintedstyle{algol}

\usepackage[]{amsmath, amssymb, amsthm, mathtools} 
\usepackage[backend=biber, bibencoding=utf8, style=ieee]{biblatex} 
\addbibresource{../../FYS-STK4155.bib}

\usepackage[]{hyperref} 
\usepackage[noabbrev, capitalize]{cleveref}

\newcommand{\cost}{\mathcal{C}}

\title{\textsc{Solving the heat-equation using neural networks}}
\author{Ivar Stangeby}

\begin{document}
	\maketitle

	\chapter{Introduction}

	Partial differential equations have widespread use in the sciences, in
	fields ranging from structural engineering to economics. The solution
	of such equations in an accurate and precise way is of paramount
	importance.

	Several approaches to the solution of such equations exists, with the
	more notable being \emph{finite difference} and \emph{finite
	element}-methods. These two methods approach the problem at hand from
	two different angles. In the finite difference schemes, the
	differential operator is discretized, typically using schemes arising
	from Taylor approximations to function. Finite element methods on the
	other hand, instead of discretizing the operators, discretize the
	function space for which the solution is sought.

	In recent years, the use of artificial intelligence in the solution of
	partial differential equations has started to gain some traction, see
	for instance the DeepXDE-network \cite{luDeepXDEDeepLearning2019} and
	the Deep Galerkin Method (DGM)
	\cite{al-aradiSolvingNonlinearHighDimensional}.

	The neural network approach to solving PDEs is meshless, that is, it
	does not rely on any underlying discretization of the domain as opposed
	to finite difference and finite element methods. Furthermore, the
	boundary conditions of the problem are encoded directly in the
	cost-function used to optimize the neural network. In finite elements
	the solution space is constructed in such a way as to automatically
	satisfy the boundary conditions, and in finite differences the boundary
	conditions must be imposed explicitly.

	\chapter{Theory}

	\section{The heat equation}

	We wish to solve the one-dimensional heat-equation. The problem reads
	as follows. Find \( u \) such that
	\begin{equation}
		\label{eq:heat_eq}
		\pd[2]{u(x, t)}{x} = \pd{u(x, t)}{t}
	\end{equation}
	subject to the boundary conditions \( u(0, t) = u(1, t) = 0 \) and the
	initial condition
	\begin{equation}
		u(x, 0) = \sin(\pi x).
	\end{equation}
	We interpret the problem as that of finding the temperature gradient in
	a rod of fixed length \( L = 1\).

	\subsection{An analytical solution}

	In order to assess a-posteriori model performance, we need to know the
	exact solution to our specific instance of the heat equation.

	Following \cite{hancockLinearPartialDifferential} we derive an
	analytical solution.  We assume separation of variables and see whether
	this leads us to anything conclusive. We write
	\begin{equation}
		u(x, t) = X(x)T(t)
	\end{equation}
	where \( X \) carries the spatial dependence and \( T \) carries the
	temporal dependence of the problem.  By taking the required partial
	derivatives we obtain \(X''(x)T(t) = X(x)T'(t)\) from which we can
	deduce that the ratios
	\begin{equation}
		\label{eq:ratio}
		\frac{X''(x)}{X(x)} = \frac{T'(t)}{T(t)} = -\lambda
	\end{equation}
	are constant. From the boundary conditions, we have that
	\begin{align}
		u(0, t) = X(0)T(t) = 0, \\
		u(1, t) = X(1)T(t) = 0.
	\end{align}
	By examination, we note that if \(T(t) = 0\) we obtain the trivial
	solution \( u = 0\), which does not satisfy the boundary conditions.
	Hence \( T \neq 0 \). Thus to satisfy the boundary conditions we must
	have \( X(0) = X(1) = 0 \).

	The ratios in \cref{eq:ratio} gives rise to two ODEs, one in time and
	one in space:
	\begin{align}
		X''(x) + \lambda X(x) &= 0, \label{eq:spatialODE} \\
		T'(t) + \lambda T(t) &= 0 \label{eq:temp_ODE}.
	\end{align}
	We have obtained boundary conditions for the spatial ODE, hence we
	start by solving that.

	As the constant \( \lambda \) is unknown, we have to consider three
	cases. The cases \( \lambda < 0 \) and \( \lambda = 0 \) can be shown
	to lead to \( u = 0\) which we discard. Thus we assume \( \lambda >
	0\).  The solution to \cref{eq:spatialODE} is then
	\begin{equation}
		X(x) = A\cos(\sqrt{\lambda}x) + B\sin(\sqrt{\lambda}x).
	\end{equation}
	By imposing the boundary conditions we obtain \( A = 0 \) and \(
	B\sin(\sqrt{\lambda}) = 0 \). Again, if we allow \( B = 0\), we obtain
	\( u = 0\), so we discard this possibility. Hence \( \sin(\sqrt{\lambda}) = 0
	\) from which we can deduce that
	\begin{equation}
		\lambda = \pi^2 n^2 
	\end{equation}
	for \( n \in \mathbb{N}^+\). This tells us that we have an infinite
	number of solutions, one for each \( n \):
	\begin{equation}
		\label{eq:solution_space}
		X_n(x) = B_n\sin(\pi n x)
	\end{equation}
	for some unknown constant \( B_n \).

	We now solve for the spatial component \( T \). For each possible
	choice of \( \lambda \) we obtain
	\begin{equation}
		T_n'(t) +\pi^2 n^2T_n(t) = 0, 
	\end{equation}
	which has solutions
	\begin{equation}
		\label{eq:solution_temp}
		T_n(t) = C_n e^{-n^2\pi^2t}.
	\end{equation}
	Combining \cref{eq:solution_space,eq:solution_temp} we get the family of solutions
	\begin{equation}
		u_n(x, t) = B_n C_n \sin(n\pi x)e^{-n^2\pi^2t}.
	\end{equation}
	In principle, we would have to consider the fourier coefficients of the
	infinite sum of these solutions, however, by examining our initial
	conditions, we see that \( u_1 \) is the desired solution, with \( B_1
	C_1 = 1 \).

	Thus, our closed form solution to the one-dimensional heat equation
	with our prescribed boundary and initial conditions is:
	\begin{equation}
		u(x, t) = \sin(\pi x) e^{-\pi^2 t}.
	\end{equation}

	\section{Finite differences}

	We wish to solve the heat equation over the domain \( \Omega = [0, L]
	\times [0, T]  \). Discretizing using \( N \) grid points in the
	spatial direction and \( M \) grid points in the temporal direction, we
	obtain step sizes \( \Delta x = L / (N - 1) \) and \( \Delta t = T / (M
	- 1) \). Of the myriad of different finite difference schemes, we
	choose the \emph{forward in time---centered in space} (FTCS).

	\subsection{Forward in time}
	We start by discretizing the time-differential in \cref{eq:heat_eq} by
	taking a first-order Taylor approximation of \( u \). We have
	\begin{equation}
		u(x, t + \Delta t) = \sum_{n = 0}^\infty \frac{\Delta t^n}{n!} \pd[n]{u(x, t)}{t} = u(x, t) + \Delta t \pd{u(x, t)}{t} + \mathcal{O}(\Delta t).
	\end{equation}
	We discard higher order terms and rearrange, yielding
	\begin{equation}
		\label{eq:forward_time}
		\pd{u(x, t)}{t} \approx \frac{u(x, t+\Delta t) - u(x, t)}{\Delta t}, 
	\end{equation}
	with a truncation error that goes as \( \mathcal{O}(\Delta t) \).

	\subsection{Centered in space}

	The centered difference scheme involves two second-order
	Taylor-aproximations in space:
	\begin{align}
		u(x + \Delta x, t) &= u(x, t) + \Delta x \dpd{u(x, t)}{x} +
		\frac{\Delta x^2}{2} \dpd[2]{u(x, t)}{x} + \mathcal{O}(\Delta x^2) \\
		u(x - \Delta x, t) &= u(x, t) - \Delta x \dpd{u(x, t)}{x} + \frac{\Delta x^2}{2} \dpd[2]{u(x, t)}{x} + \mathcal{O}(\Delta x^2).
	\end{align}
	Truncating, adding the equations to elimiate the first order
	derivatives, and solving for the second derivative yields:
	\begin{equation}
		\label{eq:centered_space}
		\dpd[2]{u(x, t)}{x} \approx \frac{u(x + \Delta x, t) - 2 u(x,
		t) + u(x - \Delta x, t)}{\Delta x^2}
	\end{equation}
	with a truncation order of \( \mathcal{O}(\Delta x^2) \).

	\subsection{The full scheme}

	Plugging \cref{eq:forward_time,eq:centered_space} into the heat
	equation given in \cref{eq:heat_eq} yields
	\begin{equation}
		\frac{u(x, t + \Delta t) - u(x, t)}{\Delta t} = \frac{u(x +
		\Delta x, t) - 2u(x, t) + u(x - \Delta x, t)}{\Delta x^2}.
	\end{equation}
	We are interesting in solving the equation forwards in time, so we
	solve for the term \( u(x, t + \Delta t) \), yielding:
	\begin{equation}
		u(x, t + \Delta t) = \left( u(x + \Delta x, t) + u(x - \Delta x, t) \right) \frac{\Delta t}{\Delta x^2} + \left(1 - 2 \frac{\Delta t}{\Delta x^2}\right)u(x, t).
	\end{equation}
	Note that the solution at the next time step requires three spatial
	solutions at the previous time-step. Hence, we are reliant on the
	boundary conditions to keep the algorithm running.

	While this scheme is easy to implement, it suffers from stability
	issues, which can be combated by tweaking the ratio \( \Delta t /
	\Delta x^2\). By ensuring that 
	\begin{equation}
		\Delta t \leq \frac{\Delta x^2}{2}, 
	\end{equation}
	the method behaves nicely.

	\section{Neural network in the context of PDEs}

	Consider a neural network \( f:\Omega \to \mathbb{R}\) parameterized by \(\theta\).  The
	question is now, how do we embed the partial differential equation in
	the neural network? The answer lies in how we engineer our loss
	function.

	\subsection{Criterion}

	From \cref{eq:heat_eq} we can see that we wish to minimize
	the following expression:
	\begin{equation}
		\label{eq:min_residual}
		\min_{\theta} \left[\dpd[2]{f(x, t; \theta)}{x} - \dpd{f(x, t; \theta)}{t}\right]
	\end{equation}
	for \( x \in [0, L] \) and \( t > 0\). Thus, we can simply use the mean
	squared error of the residuals in the right hand side of
	\cref{eq:min_residual}. We therefore define our cost function \( \cost
	\) as the mean squared residual error over all (assume \( n \))
	data-points  in \(\Omega = [0, L] \times [0, T]\).
	\begin{equation}
		\cost(f) = \frac{1}{n} \sum_{i=1}^n \left(  \dpd[2]{f(x_i, t_i)}{x}
		- \dpd{f(x_i, t_i)}{t}\right)^2
	\end{equation}

	\subsection{Embedding boundary conditions}

	Currently, we are not imposing any boundary conditions. Thus, by
	naively running the network as is, we may find any function whose
	second partial derivative in space is equal to its partial derivative
	in time. There are several approaches to solving this problem. In
	\cite{luDeepXDEDeepLearning2019} they train the network according to a
	multi-task loss following the same trick as in \cref{eq:min_residual}
	but for data-points lying on the boundary.

	We instead opt for the method of constructing a \emph{trial function}
	that acts as a mediator between the neural network and the
	loss-function which ensures that the network learns to satisfy the
	boundary conditions.

	Recall from our boundary and initial conditions that we require \( u(0,
	t) = u(1, t) = 0 \) and \( u(x, 0) = \sin(\pi x) \). The function \( g
	\) defined by
	\begin{equation}
		\label{eq:trial}
		g(x, t; \theta) = (1 - t) \sin(\pi x) + x(1-x)t f(x, t; \theta),
	\end{equation}
	satisfies these conditions. It is therefore this function we pass into
	our cost function \( \cost \).

	\subsection{Neural network architecture}

	As our architecture we wanted to reuse the \textsc{GenericNN}
	implemented in \cite{stangebyFYSSTK4155Project2019a}, which is a
	feed-forward neural network with \textsc{Relu}-activations at each
	hidden layer. However, after initial testing, using \textsc{Relu}
	yielding wildly oscilating loss, hence we chose the more stable
	\textsc{Sigmoid}-activation. This led to slower learning rates, due to
	the saturation of gradients, but allowed the network to get a tighter
	approximation.

	\section{Performance metrics}

	In order to evaluate our numerical approximations, whether they are
	computed using finite-difference or they are trained neural networks,
	we have several metrics we can consider. Calling our numerical
	approximation to \( u \) for \( \tilde{u} \), we define our errors as
	\begin{equation}
		e_p = \| u - \tilde{u} \|_p,
	\end{equation}
	where \( \| \cdot \|_p \) denotes the \(p\)-norm for vectors.

	\chapter{Implementation}

	The finite difference solver is implemented using a naive for-loop
	approach in \texttt{numpy} and can be seen in \cref{lst:ftcs}. In order
	to satisfy the stability criterion, we choose our desired \( \Delta x
	\) and compute a corresponding \( \Delta t\) by
	\begin{equation}
		\Delta t = \Delta x^2 / 2.
	\end{equation}

	The neural network is implemented in \texttt{tensorflow}, which takes
	care of the backpropagation of gradients through the network using the
	\texttt{AutoGrad}-package. Ideally, we wanted to implement this in
	\texttt{PyTorch}, however, taking element-wise gradients of the network
	output with the respect to the network input, as is required to
	evaluate \cref{eq:trial}, turned out to be hard to do in
	\texttt{PyTorch}. Due to my lacking experience in \texttt{Tensorflow},
	the implementation is a modified version of the one used to solve the
	wave equation in \cite{heinDataAnalysisMachine}. This was implemented
	in \texttt{Tensorflow V1}, thus we had to add the compatibility
	modifications shown in \cref{lst:compatibility} in order for this to
	run using the newer \texttt{Tensorflow V2} API.

	\begin{listing}
		\centering	
		\begin{minted}{python}
import tensorflow.compat.v1 as tf
tf.disable_eager_execution()		
		\end{minted}
		\caption{Compatibility modification needed to use the
		deprecated functionality of \texttt{Tensorflow} V1 in the V2
	API.}
		\label{lst:compatibility}
	\end{listing}

	\begin{listing}
		\centering
		\begin{minted}{python}
def ftcs(space_resolution, time_resolution, space_min_max=[0, 1],
		time_max=1, boundary_conditions=[0, 0], initial_condition=None):
    """
    Solves the 1D-heat equation using a forward-in-time 
    centered-in-space discretization scheme.

    :param space_resolution: number of points in spatial direction
    :param time_resolution: number of points in temporal direction
    :param space_min_max: the spatial boundary values
    :param time_max: the time to run the simulation for
    :param boundary_conditions: the boundary conditions at the space_min_max-values
    :param initial_conditions: the initial conditions at time = 0, callable.
    """

    if initial_condition is None:
        initial_condition: lambda x: 0

    dt = time_max / time_resolution
    u0, dx = np.linspace(*space_min_max, num=space_resolution, retstep=True)
    u = np.zeros((time_resolution, space_resolution))
    u[:, [0, -1]] = boundary_conditions
    u[0] = initial_condition(u0)

    F = dt / dx ** 2
    for step in tqdm.trange(time_resolution - 1):
        for i in range(1, space_resolution - 1):
            u[step + 1, i] = u[step, i] +\ 
	    	F * (u[step, i - 1] - 2 * u[step, i] + u[step, i + 1])

    return u

@np.vectorize
def initial_condition(x):
    return np.sin(np.pi * x)
		\end{minted}
		\caption{A \texttt{numpy}-based implementation of the
			\emph{forward in time---centered in space} scheme. The
			initial conditions are passed in as a callable, and the
			function returns the solution at all time-steps in form
			of a \texttt{time\_resolution} \(\times\)
		\texttt{space\_resolution} matrix.}
		\label{lst:ftcs}
	\end{listing}

	\chapter{Numerical experiments}

\clearpage
\printbibliography
\end{document}

### 2019-12-20 assignment_three/src/neural_network.py

@@ -5,8 +5,7 @@ import torch.nn.functional as F
 class GenericNN(nn.Module):
 
     def __init__(self, num_input_features=10, num_output_features=10, num_hidden_layers=1, num_hidden_features=20,
-                 activation=F.relu,
-                 alpha=1, regularization='none'):
+                 activation=F.relu):
         super().__init__()
 
         self.input_layer = nn.Linear(num_input_features, num_hidden_features).double()

### 2019-12-20 assignment_three/src/neural_network_keras.py

@@ -1,0 +1,40 @@
+import numpy as np
+import tensorflow as tf
+
+max_space = 1
+num_space = 10
+num_time = 10
+max_time = 1
+num_epochs = 10000
+t_np = np.linspace(0, max_time, num_time)
+x_np = np.linspace(0, max_space, num_space)
+
+X, T = np.meshgrid(x_np, t_np)
+xt = np.dstack((X, T)).reshape(-1, 2)
+x, t = xt[:, 0], xt[:, 1]
+
+placeholder_zeros = tf.reshape(tf.convert_to_tensor(np.zeros(t.shape), dtype=tf.float64), shape=(-1, 1))
+
+def initial_condition(x):
+    return tf.sin(np.pi * x)
+
+
+def trial_loss(dnn_output, gt=None):
+    g_trial = (1 - t) * initial_condition(x) + x * (1 - x) * t * dnn_output
+    g_trial_dt = tf.gradients(g_trial, t)
+    g_trial_dx_dx = tf.gradients(tf.gradients(g_trial, x), x)
+    return tf.losses.mean_squared_error(placeholder_zeros, g_trial_dt[0] - g_trial_dx_dx[0])
+
+
+
+# Construction
+t_tf = tf.convert_to_tensor(t.reshape((-1, 1)), dtype=tf.float64)  # x-values
+x_tf = tf.convert_to_tensor(x.reshape((-1, 1)), dtype=tf.float64)  # t-values
+p_tf = tf.convert_to_tensor(xt, dtype=tf.float64)  # (x, t)-values
+
+model = tf.keras.models.Sequential([
+    tf.keras.layers.Dense(20, activation='relu')
+])
+
+model.compile(optimizer='adam', loss=trial_loss)
+model.fit(p_tf)

### 2019-12-20 assignment_three/src/neural_network_tf.py

@@ -1,0 +1,83 @@
+import numpy as np
+import tensorflow.compat.v1 as tf
+
+# from tensorflow_core.python import set_random_seed, global_variables_initializer, Session
+# from tensorflow_core.python.framework.ops import disable_eager_execution
+# from tensorflow_core.python.layers import layers
+# from tensorflow_core.python.training.gradient_descent import GradientDescentOptimizer
+
+tf.disable_eager_execution()
+
+
+def initial_condition(x):
+    return tf.sin(np.pi * x)
+
+
+tf.set_random_seed(4155)
+
+max_space = 1
+num_space = 10
+num_time = 10
+max_time = 1
+num_epochs = 10000
+t_np = np.linspace(0, max_time, num_time)
+x_np = np.linspace(0, max_space, num_space)
+
+X, T = np.meshgrid(x_np, t_np)
+xt = np.dstack((X, T)).reshape(-1, 2)
+x, t = xt[:, 0], xt[:, 1]
+
+# Construction
+t_tf = tf.convert_to_tensor(t.reshape((-1, 1)), dtype=tf.float64)  # x-values
+x_tf = tf.convert_to_tensor(x.reshape((-1, 1)), dtype=tf.float64)  # t-values
+p_tf = tf.convert_to_tensor(xt, dtype=tf.float64)  # (x, t)-values
+
+placeholder_zeros = tf.reshape(tf.convert_to_tensor(np.zeros(t.shape)), shape=(-1, 1))
+
+num_hidden_neurons = [20, 20, 20]
+num_hidden_layers = np.size(num_hidden_neurons)
+
+# We define the dense neural network, with output 'dnn_output'.
+with tf.name_scope('dnn'):
+    previous_layer = p_tf
+
+    for l in range(num_hidden_layers):
+        current_layer = tf.layers.dense(previous_layer, num_hidden_neurons[l], name=f'hidden{l}')
+        previous_layer = current_layer
+
+    dnn_output = tf.layers.dense(previous_layer, 1, name='output')
+
+# The cost function is defined in terms of the trial function.
+with tf.name_scope('cost'):
+    g_trial = (1 - t) * initial_condition(x) + x * (1 - x) * t * dnn_output
+    g_trial_dt = tf.gradients(g_trial, t)
+    g_trial_dx_dx = tf.gradients(tf.gradients(g_trial, x), x)
[…]

### 2019-12-20 assignment_three/src/numerical_solver.py

@@ -21,12 +21,10 @@ def ftcs(space_resolution, time_resolution, space_min_max=[0, 1], time_max=1, bo
     dt = time_max / time_resolution
     u0, dx = np.linspace(*space_min_max, num=space_resolution, retstep=True)
     u = np.zeros((time_resolution, space_resolution))
-    u[:, 0] = boundary_conditions[0]
-    u[:, -1] = boundary_conditions[1]
+    u[:, [0, -1]] = boundary_conditions
     u[0] = initial_condition(u0)
 
     F = dt / dx ** 2
-    print(F)
     for step in tqdm.trange(time_resolution - 1):
         for i in range(1, space_resolution - 1):
             u[step + 1, i] = u[step, i] + F * (u[step, i - 1] - 2 * u[step, i] + u[step, i + 1])

### 2019-12-20 assignment_three/src/simulation.py

@@ -1,0 +1,83 @@
+import numpy as np
+import torch
+
+from torch.autograd import grad
+from assignment_three.src.neural_network import GenericNN
+from assignment_three.src.numerical_solver import initial_condition
+
+desired_dx = [1 / 10, 1 / 100]
+desired_dt = [dx ** 2 / 2 for dx in desired_dx]
+boundary_conditions = [0, 0]
+space_min_max = [0, 1]
+time_min_max = [0, 1]
+
+
+def trial(points, network_output):
+    x = points[:, 0]
+    t = points[:, 1]
+
+    return ((1 - t) * torch.sin(np.pi * x) + x * (1 - x) * t * network_output.view(-1)).view(-1, 1)
+
+
+def cost(d_utrial, dd_utrial):
+    x = xt_vals[:, 0]
+    t = xt_vals[:, 1]
+
+    # u_trial_dt = torch.ones_like(u_trial)
+    # u_trial_dx_dx = torch.ones_like(u_trial, requires_grad=True)
+
+    # grad(u_trial, x, grad_outputs=u_trial_dx_dx, allow_unused=True, retain_graph=True)
+    # grad(u_trial_dx_dx, x, grad_outputs=u_trial_dx_dx, allow_unused=True, retain_graph=True)
+    # grad(u_hat, t, grad_outputs=u_trial_dt, allow_unused=True, retain_graph=True)
+
+    n = u_trial.shape[0]
+
+    return torch.sum((d_utrial[:, 1] - dd_utrial[:, 0]) ** 2) / n
+
+
+def compute_gradient_penalty(network_output, network_input):
+    # ref: https://github.com/eriklindernoren/PyTorch-GAN/blob/master/implementations/wgan_gp/wgan_gp.py
+    '''
+
+    :param input: state[index]
+    :param network: actor or critic
+    :return: gradient penalty
+    '''
+    input_ = torch.tensor(network_input).requires_grad_(True)
+    mask = torch.ones_like(network_output)
+    gradients = grad(network_output, input_, grad_outputs=mask,
+                     retain_graph=True, create_graph=True,
+                     allow_unused=True)[0]  # get tensor from tuple
+    gradients = mask.view(-1, 1)
+
+    return gradients
+
+
+NUM_EPOCHS = 10000
+for dx, dt in zip(desired_dx, desired_dt):
[…]

### 2019-12-20 assignment_three/src/test..py

@@ -1,0 +1,2 @@
+import torch
+from torch.autograd import grad_mode

### 2019-12-20 assignment_three/src/tf_test.py

@@ -1,0 +1,157 @@
+import tensorflow.compat.v1 as tf
+
+tf.disable_eager_execution()
+import numpy as np
+from matplotlib import cm
+from matplotlib import pyplot as plt
+from mpl_toolkits.mplot3d import axes3d
+
+Nx = 100
+x_np = np.linspace(0, 1, Nx)
+
+Nt = 100
+t_np = np.linspace(0, 1, Nt)
+
+X, T = np.meshgrid(x_np, t_np)
+
+x = X.ravel()
+t = T.ravel()
+
+## The construction phase
+
+zeros = tf.reshape(tf.convert_to_tensor(np.zeros(x.shape)), shape=(-1, 1))
+x = tf.reshape(tf.convert_to_tensor(x), shape=(-1, 1))
+t = tf.reshape(tf.convert_to_tensor(t), shape=(-1, 1))
+
+points = tf.concat([x, t], 1)
+
+num_iter = 10000
+num_hidden_neurons = [20, 20, 20]
+
+X = tf.convert_to_tensor(X)
+T = tf.convert_to_tensor(T)
+
+with tf.variable_scope('dnn'):
+    num_hidden_layers = np.size(num_hidden_neurons)
+
+    previous_layer = points
+
+    for l in range(num_hidden_layers):
+        current_layer = tf.layers.dense(previous_layer, num_hidden_neurons[l], activation=tf.nn.sigmoid)
+        previous_layer = current_layer
+
+    dnn_output = tf.layers.dense(previous_layer, 1)
+
+
+def initial_conditions(x):
+    return tf.sin(np.pi * x)
+
+
+with tf.name_scope('loss'):
+    g_trial = (1 - t) * initial_conditions(x) + x * (1 - x) * t * dnn_output
+
+    g_trial_dt = tf.gradients(g_trial, t)
+    g_trial_d2x = tf.gradients(tf.gradients(g_trial, x), x)
+
+    loss = tf.losses.mean_squared_error(zeros, g_trial_dt[0] - g_trial_d2x[0])
+
+learning_rate = 0.01
+with tf.name_scope('train'):
+    optimizer = tf.train.GradientDescentOptimizer(learning_rate)
+    traning_op = optimizer.minimize(loss)
+
+init = tf.global_variables_initializer()
+
+exp = tf.exp(-np.pi ** 2 * t)
+sin = tf.sin(np.pi * x)
+print(exp.dtype, sin.dtype)
+g_analytic = exp * sin
[…]

### 2019-12-20 assignment_three/article/main.tex
Cleaned up. Done with project 3.

@@
 \usepackage[]{palatino} 
+\usepackage[]{graphicx} 
 \usepackage[]{commath} 
+\usepackage[]{booktabs} 
 \usepackage[]{microtype} 
@@
 
+\newcommand*{\github}[1]{\def\@github{\color{gray!70}#1}}
+\newcommand{\githublogo}{\raisebox{-1pt}{\includegraphics[height=9pt]{imgs/github}}}
 \title{\textsc{Solving the heat-equation using neural networks}}
@@
 	\maketitle
+
+	\begin{abstract}
+		In this project we compare a standard finite difference method
+		to neural networks in the solution of the heat equation. We
+		show that the neural network severely underperforms when
+		compared to the finite differences, and that the run-time is
+		significantly slower. This gives us no immediate reason to
+		chose a network based approach over classical methods. All
+		relevant code can be found on GitHub:
+
+		\centering{\href{https://github.com/qTipTip/FYS-STK4155}{\githublogo \, \color{gray!50}
+github.com/qTipTip }}
+
+	\end{abstract}
 
@@
 	\end{equation}
-	where \( \| \cdot \|_p \) denotes the \(p\)-norm for vectors.
+	where \( \| \cdot \|_p \) denotes the matrix \(p\)-norm. We consider in
+	particular the \( \infty\)-norm, coinciding with the maximum absolute
+	error, and the \( 2 \)-norm, which is the standard euclidean
+	(Frobenius) norm.
 
@@
 
+	We start by running the forward in time--central in space scheme. The
+	errors are reported in \cref{tab:ftsc_errors}. The 2-norm steadily
+	decreases, which means that on average the approximation becomes better
+	with decreased grid-size. Also, the \( \infty\)-norm decreases steadily
+	towards zero. With this in mind, the approximations are fairly good at
+	even low spatial resolutions. Due to the stability criterion, running
+	this method for even smaller grid-sizes in space quickly becomes
+	infeasible.
+
+	\begin{table}[htpb]
+		\centering
+		\caption{Solving the heat equation for varying spatial
+			resolution, with the temporal resolution chosen to
+			satisfy the stability criterion. We report both the \(
+			e_1 \) and \( e_2 \) errors.} 
+		\label{tab:ftsc_errors}
+		\begin{tabular}{lrr}
+			\toprule
+			\textsc{FTCS} & \( e_\infty \) & \( e_2 \) \\
+			\midrule
+			\( \Delta x = 1 / 10 \) & \num{0.00357} & \num{0.04674} \\
+			\( \Delta x = 1 / 20 \) & \num{0.00098} & \num{0.03688} \\
+			\( \Delta x = 1 / 30 \) & \num{0.00044} & \num{0.03113} \\
+			\( \Delta x = 1 / 40 \) & \num{0.00025}& \num{0.02743} \\
+			\( \Delta x = 1 / 100 \) & \num{0.00004} &\num{0.01787} \\
+			\bottomrule
+		\end{tabular}
+	\end{table}
+
+	We now run the neural network. During testing we realized that the
+	neural network also needs a balance between the number of data-points
+	in the temporal domain and the number of data-points in the spatial
+	domain. We therefore run the network with both spatial and temporal
+	resolution increasing at the same rate, following that of
+	\cref{tab:ftsc_errors}. We evaluate the same errors. The network is run
+	for \( 10000 \) epochs (which in hindsight might be on the fewer end of
+	things), and the results are shown in \cref{tab:nnpde_errors}. Note in
+	particular how the error not neccesarily decreases with higher
+	resolution. This might indicate that a different architecture is to be
+	prefered. Maybe a shallower network with larger width. This will also
+	reduce the run-time of the network, as the current architecture is
+	quite sluggish.
+
+	Taking the reported errors at face-value, the classical finite
+	difference method is to prefer, despite its stability issues.
+
+	\begin{table}[htpb]
+		\centering
+		\caption{Solving the heat equation for identical spatial and
+			teporal resolution.  In addition to the errors not
+			neccesarily decreasing on increased resolution, the
+			network complexity quickly increases, which incurs
+			significant run-time increase. This might indicate that
+			a different architecture is to be prefered.
+			} 
+		\label{tab:nnpde_errors}
+		\begin{tabular}{lrr}
+			\toprule
+			\textsc{FTCS} & \( e_\infty \) & \( e_2 \) \\
+			\midrule
+			\(\Delta t = \Delta x = 1 / 10 \) & \num{0.03684} & \num{0.14905} \\
+			\(\Delta t =  \Delta x = 1 / 20 \) & \num{0.14728} & \num{1.24897} \\
+			\(\Delta t =  \Delta x = 1 / 30 \) & \num{0.03979} & \num{0.39177} \\
+			\(\Delta t =  \Delta x = 1 / 40 \) & \num{0.04963}& \num{0.55036} \\
+			\(\Delta t =  \Delta x = 1 / 100 \) & \num{0.05103} &\num{1.35165} \\
+			\bottomrule
+		\end{tabular}
+	\end{table}
+
+	\chapter{Conclusion}
+
+	In this project we have discussed the solution of the heat-equation,
+	both using finite-difference methods, and a neural network based
+	approach. Based on the results, the \emph{forward in time---centered in
+	space} difference scheme outperformed the neural net both in accuracy
+	and in runtime, despite its stability issues.
+
+	The benefits of using a neural network for this problem is hard to see
+	at face value. One benefit is however that the neural network does not
+	need an underlying domain discretization. This does not constitute any
+	advantage in such a simple case as our 1D diffusion equation, but in
+	higher dimensions, this can prove to be a great advantage.
+
+	For instance in the finite element method, the numerical properties of
+	the approximation greatly hinges on the quality of the underlying mesh.
+
+	The poor results of the neural network sparked an interest in trying
+	out different architectures, however I spent too much time trying to
+	make this work using \texttt{PyTorch}, and realized too late that I had
+	to migrate to \texttt{Tensorflow}, where my experience is lacking.
+
+	In conclusion, for simple tasks like this, it seems like the neural
+	network works well, comparatively, for very coarse domain
+	discretizations, however falls behind finer resolutions.
+
+	Outside of the scope of this project would be the solution of the 2D
+	wave equation using a neural network over a non-trivial domain, which
+	might be able to highlight some of the benefits of not requiring domain
+	discretization.
 \clearpage
+
+\appendix
+
+\chapter{Finding eigenvalues}
+
+In this appendix we briefly discuss the idea behind how you can estimate
+eigenvalues and eigevectors of a real symmetric matrix using the neural network
+framework discussed in this project. Unfortunately, I was unable to make the
+network converge, and due to time-constraints this was relegated to an
+appendix.
+
+Following \cite{yiNeuralNetworksBased2004} we can also use our neural network
+to find the eigenvalues of a real symmetric matrix \( A \). In particular, let
+\( Q \)  be a \( 6 \times 6 \) random matrix and define
+\begin{equation}
+	A = \frac{Q^T + Q}{2}, 
+\end{equation}
+which then is a symmetric real matrix.
+
+In \cite{yiNeuralNetworksBased2004} they show that any non-zero equilibrium
+point of the system
+\begin{equation}
+	\pd{u(t)}{t} = -u(t) + f(u(t)), 
+\end{equation}
+where \( f(u) \) is defined as
+\begin{equation}
+	f(u) = \left[u^Tu A + \left(1 - u^TAu\right)I\right]u
+\end{equation}
+is an eigenvalue of the matrix \( A \).
+Here \( u \) is the network output.
+
+To recast this into our framework, we wish to minimize the function 
+\begin{equation}
+	\pd{u(t)}{t} + u(t) - f(u(t))
+\end{equation}
+with respect to our network parameters. The network is shown to converge to
+\emph{an} eigenvalue of the matrix \( A \). However, you can make the network
+converge to the largest eigenvalue by choosing an initial condition for the
+network non-orthogonal to the eigenspace of the largest eigenvalue. This
+eigenspace is not known a-priori, so a random initialization is made with the
+motivation that the probability is rather large that the random vector is not
+orthogonal to the eigenspace.
 \printbibliography

### 2019-12-20 assignment_three/src/neural_network_keras.py

@@ -1,40 +1,0 @@
-import numpy as np
-import tensorflow as tf
-
-max_space = 1
-num_space = 10
-num_time = 10
-max_time = 1
-num_epochs = 10000
-t_np = np.linspace(0, max_time, num_time)
-x_np = np.linspace(0, max_space, num_space)
-
-X, T = np.meshgrid(x_np, t_np)
-xt = np.dstack((X, T)).reshape(-1, 2)
-x, t = xt[:, 0], xt[:, 1]
-
-placeholder_zeros = tf.reshape(tf.convert_to_tensor(np.zeros(t.shape), dtype=tf.float64), shape=(-1, 1))
-
-def initial_condition(x):
-    return tf.sin(np.pi * x)
-
-
-def trial_loss(dnn_output, gt=None):
-    g_trial = (1 - t) * initial_condition(x) + x * (1 - x) * t * dnn_output
-    g_trial_dt = tf.gradients(g_trial, t)
-    g_trial_dx_dx = tf.gradients(tf.gradients(g_trial, x), x)
-    return tf.losses.mean_squared_error(placeholder_zeros, g_trial_dt[0] - g_trial_dx_dx[0])
-
-
-
-# Construction
-t_tf = tf.convert_to_tensor(t.reshape((-1, 1)), dtype=tf.float64)  # x-values
-x_tf = tf.convert_to_tensor(x.reshape((-1, 1)), dtype=tf.float64)  # t-values
-p_tf = tf.convert_to_tensor(xt, dtype=tf.float64)  # (x, t)-values
-
-model = tf.keras.models.Sequential([
-    tf.keras.layers.Dense(20, activation='relu')
-])
-
-model.compile(optimizer='adam', loss=trial_loss)
-model.fit(p_tf)

### 2019-12-20 assignment_three/src/neural_network_tf.py

@@ -1,83 +1,0 @@
-import numpy as np
-import tensorflow.compat.v1 as tf
-
-# from tensorflow_core.python import set_random_seed, global_variables_initializer, Session
-# from tensorflow_core.python.framework.ops import disable_eager_execution
-# from tensorflow_core.python.layers import layers
-# from tensorflow_core.python.training.gradient_descent import GradientDescentOptimizer
-
-tf.disable_eager_execution()
-
-
-def initial_condition(x):
-    return tf.sin(np.pi * x)
-
-
-tf.set_random_seed(4155)
-
-max_space = 1
-num_space = 10
-num_time = 10
-max_time = 1
-num_epochs = 10000
-t_np = np.linspace(0, max_time, num_time)
-x_np = np.linspace(0, max_space, num_space)
-
-X, T = np.meshgrid(x_np, t_np)
-xt = np.dstack((X, T)).reshape(-1, 2)
-x, t = xt[:, 0], xt[:, 1]
-
-# Construction
-t_tf = tf.convert_to_tensor(t.reshape((-1, 1)), dtype=tf.float64)  # x-values
-x_tf = tf.convert_to_tensor(x.reshape((-1, 1)), dtype=tf.float64)  # t-values
-p_tf = tf.convert_to_tensor(xt, dtype=tf.float64)  # (x, t)-values
-
-placeholder_zeros = tf.reshape(tf.convert_to_tensor(np.zeros(t.shape)), shape=(-1, 1))
-
-num_hidden_neurons = [20, 20, 20]
-num_hidden_layers = np.size(num_hidden_neurons)
-
-# We define the dense neural network, with output 'dnn_output'.
-with tf.name_scope('dnn'):
-    previous_layer = p_tf
-
-    for l in range(num_hidden_layers):
-        current_layer = tf.layers.dense(previous_layer, num_hidden_neurons[l], name=f'hidden{l}')
-        previous_layer = current_layer
-
-    dnn_output = tf.layers.dense(previous_layer, 1, name='output')
-
-# The cost function is defined in terms of the trial function.
-with tf.name_scope('cost'):
-    g_trial = (1 - t) * initial_condition(x) + x * (1 - x) * t * dnn_output
-    g_trial_dt = tf.gradients(g_trial, t)
-    g_trial_dx_dx = tf.gradients(tf.gradients(g_trial, x), x)
[…]

### 2019-12-20 assignment_three/src/numerical_solver.py

@@ -37,25 +37,31 @@ def initial_condition(x):
     return np.sin(np.pi * x)
 
 
-@np.vectorize
 def exact_solution(x, t):
-    return np.sin(np.pi(x)) * np.exp(-np.pi ** 2 * t)
+    return np.sin(np.pi * x) * np.exp(-np.pi ** 2 * t)
 
 
 if __name__ == '__main__':
 
-    desired_dx = [1 / 10, 1 / 20, 1 / 30, 1 / 40]
-
+    desired_dx = [1 / 10, 1 / 20, 1 / 30, 1 / 40, 1 / 100]
+    solutions = []
     for dx in desired_dx:
         dt = 1 / 2 * dx ** 2
 
         N = int(1 / dx)
+        n_vals = np.linspace(0, 1, N)
         T = int(1 / dt)
+        t_vals = np.linspace(0, 1, T)
 
-        u = ftcs(N, T, initial_condition=initial_condition)
-        import matplotlib.pyplot as plt
+        X_grid, T_grid = np.meshgrid(n_vals, t_vals)
 
-        for step in u:
-            plt.title(f'N = {N} T = {T}')
-            plt.plot(step, color='gray', alpha=0.7)
-        plt.show()
+        u_tilde = ftcs(N, T, initial_condition=initial_condition)
+        u_exact = exact_solution(X_grid, T_grid)
+        error = u_tilde - u_exact
+        print(f"""
+Errors with dx = {dx:.03f}
+    1-norm = {np.max(np.abs(error)):.5f}
+    2-norm = {np.linalg.norm(error, ord=2):.5f}
+""")
+
+        solutions.append(u_tilde)

### 2019-12-20 assignment_three/src/pytorch_pde_not_working.py

@@ -75,8 +75,6 @@ for dx, dt in zip(desired_dx, desired_dt):
         u_trial.backward(torch.ones_like(u_trial), retain_graph=True)  # does not work
         dd_utrial = xt_vals.grad
 
-        print(compute_gradient_penalty(u_trial, xt_vals))
-
         loss = cost(d_utrial, dd_utrial)
         loss.backward()
 

### 2019-12-20 assignment_three/src/test..py

@@ -1,2 +1,0 @@
-import torch
-from torch.autograd import grad_mode

### 2019-12-20 assignment_three/src/tf_eigenvalues.py

@@ -1,0 +1,161 @@
+import tensorflow.compat.v1 as tf
+
+tf.disable_eager_execution()
+import numpy as np
+
+Q = np.random.normal(0, 1, (6, 6))
+A = (Q.T + Q) / 2
+I = np.eye(6)
+num_iter = 100000
+
+for N in [6]:
+
+    tf.reset_default_graph()
+    x_np = np.random.random(N)
+    # Nt = 10
+    t_np = np.arange(num_iter, dtype=np.float64)
+
+    ## The construction phase
+
+    zeros = tf.reshape(tf.convert_to_tensor(np.zeros(x_np.shape)), shape=(-1, 1))
+    x = tf.reshape(tf.convert_to_tensor(x_np), shape=(-1, 1))
+    t = tf.reshape(tf.convert_to_tensor(t_np), shape=(-1, 1))
+
+    points = tf.concat([x, t], 1)
+
+    num_hidden_neurons = [20, 20, 20]
+
+    with tf.variable_scope('dnn'):
+        num_hidden_layers = np.size(num_hidden_neurons)
+
+        previous_layer = points
+
+        for l in range(num_hidden_layers):
+            current_layer = tf.layers.dense(previous_layer, num_hidden_neurons[l], activation=tf.nn.sigmoid)
+            previous_layer = current_layer
+
+        dnn_output = tf.layers.dense(previous_layer, 1)
+
+
+    def initial_conditions(x):
+        return tf.sin(np.pi * x)
+
+
+    with tf.name_scope('loss'):
+        u = dnn_output
+        u_dt = tf.gradients(u, x)[0]
+
+        f = tf.matmul(tf.transpose(u) * u * A + (1 - tf.transpose(u) * A * u) * I, u)
+
+        loss = tf.losses.mean_squared_error(zeros, u_dt + u - f)
+
+    learning_rate = 0.001
+    with tf.name_scope('train'):
+        optimizer = tf.train.GradientDescentOptimizer(learning_rate)
+        traning_op = optimizer.minimize(loss)
+
+    init = tf.global_variables_initializer()
+
+    g_dnn = None
+
+    ## The execution phase
+    with tf.Session() as sess:
+        init.run()
+        for i in range(num_iter):
+            sess.run(traning_op)
+
[…]

### 2019-12-20 assignment_three/src/tf_pde.py

@@ -1,0 +1,161 @@
+import tensorflow.compat.v1 as tf
+
+tf.disable_eager_execution()
+import numpy as np
+from matplotlib import cm
+from matplotlib import pyplot as plt
+from mpl_toolkits.mplot3d import axes3d
+
+for N in [10, 20, 30, 40, 100]:
+    tf.reset_default_graph()
+    x_np = np.linspace(0, 1, N)
+
+    # Nt = 10
+    t_np = np.linspace(0, 1, N)
+
+    X, T = np.meshgrid(x_np, t_np)
+
+    x = X.ravel()
+    t = T.ravel()
+
+    ## The construction phase
+
+    zeros = tf.reshape(tf.convert_to_tensor(np.zeros(x.shape)), shape=(-1, 1))
+    x = tf.reshape(tf.convert_to_tensor(x), shape=(-1, 1))
+    t = tf.reshape(tf.convert_to_tensor(t), shape=(-1, 1))
+
+    points = tf.concat([x, t], 1)
+
+    num_iter = 10000
+    num_hidden_neurons = [20, 20, 20]
+
+    X = tf.convert_to_tensor(X)
+    T = tf.convert_to_tensor(T)
+
+    with tf.variable_scope('dnn'):
+        num_hidden_layers = np.size(num_hidden_neurons)
+
+        previous_layer = points
+
+        for l in range(num_hidden_layers):
+            current_layer = tf.layers.dense(previous_layer, num_hidden_neurons[l], activation=tf.nn.sigmoid)
+            previous_layer = current_layer
+
+        dnn_output = tf.layers.dense(previous_layer, 1)
+
+
+    def initial_conditions(x):
+        return tf.sin(np.pi * x)
+
+
+    with tf.name_scope('loss'):
+        g_trial = (1 - t) * initial_conditions(x) + x * (1 - x) * t * dnn_output
+
+        g_trial_dt = tf.gradients(g_trial, t)
+        g_trial_d2x = tf.gradients(tf.gradients(g_trial, x), x)
+
+        loss = tf.losses.mean_squared_error(zeros, g_trial_dt[0] - g_trial_d2x[0])
+
+    learning_rate = 0.01
+    with tf.name_scope('train'):
+        optimizer = tf.train.GradientDescentOptimizer(learning_rate)
+        traning_op = optimizer.minimize(loss)
+
[…]

### 2019-12-20 assignment_three/src/tf_test.py

@@ -1,157 +1,0 @@
-import tensorflow.compat.v1 as tf
-
-tf.disable_eager_execution()
-import numpy as np
-from matplotlib import cm
-from matplotlib import pyplot as plt
-from mpl_toolkits.mplot3d import axes3d
-
-Nx = 100
-x_np = np.linspace(0, 1, Nx)
-
-Nt = 100
-t_np = np.linspace(0, 1, Nt)
-
-X, T = np.meshgrid(x_np, t_np)
-
-x = X.ravel()
-t = T.ravel()
-
-## The construction phase
-
-zeros = tf.reshape(tf.convert_to_tensor(np.zeros(x.shape)), shape=(-1, 1))
-x = tf.reshape(tf.convert_to_tensor(x), shape=(-1, 1))
-t = tf.reshape(tf.convert_to_tensor(t), shape=(-1, 1))
-
-points = tf.concat([x, t], 1)
-
-num_iter = 10000
-num_hidden_neurons = [20, 20, 20]
-
-X = tf.convert_to_tensor(X)
-T = tf.convert_to_tensor(T)
-
-with tf.variable_scope('dnn'):
-    num_hidden_layers = np.size(num_hidden_neurons)
-
-    previous_layer = points
-
-    for l in range(num_hidden_layers):
-        current_layer = tf.layers.dense(previous_layer, num_hidden_neurons[l], activation=tf.nn.sigmoid)
-        previous_layer = current_layer
-
-    dnn_output = tf.layers.dense(previous_layer, 1)
-
-
-def initial_conditions(x):
-    return tf.sin(np.pi * x)
-
-
-with tf.name_scope('loss'):
-    g_trial = (1 - t) * initial_conditions(x) + x * (1 - x) * t * dnn_output
-
-    g_trial_dt = tf.gradients(g_trial, t)
-    g_trial_d2x = tf.gradients(tf.gradients(g_trial, x), x)
-
-    loss = tf.losses.mean_squared_error(zeros, g_trial_dt[0] - g_trial_d2x[0])
-
-learning_rate = 0.01
-with tf.name_scope('train'):
-    optimizer = tf.train.GradientDescentOptimizer(learning_rate)
-    traning_op = optimizer.minimize(loss)
-
-init = tf.global_variables_initializer()
-
-exp = tf.exp(-np.pi ** 2 * t)
-sin = tf.sin(np.pi * x)
-print(exp.dtype, sin.dtype)
-g_analytic = exp * sin
[…]
